### 2011-08-23 docs/algorithm/algorithm.tex
Adding LaTeX algorithm description.

\documentclass{article}

%\usepackage[draft]{pdfdraftcopy} %http://sarovar.org/projects/pdfdraftcopy/
%\usepackage{todonotes} %http://tug.ctan.org/pkg/todonotes/
%\usepackage{fullpage}
\usepackage{graphicx}
\begin{document}
\title{Energy Efficient Routing Algorithm for Body Sensor Networks}
\author{Alvaro Prieto}
\maketitle

\pagebreak

\section{Introduction}
This document contains a brief description of the proposed routing algorithm.

\section{Overview}
The primary goal of this algorithm is to maximize network lifetime by normalizing power use across the network. Message routing will change depending on the total energy used by each sensor node. For this specific implementation, there are no relay nodes. Each sensor can relay information to and from other sensors.

This routing algorithm is based on a modified version of Dijkstra's algorithm.

\section{Operation}

\subsection{Theory}
Instead of choosing the path that maximizes the lifetime of each individual node, a path is chosen such that the overall lifetime of the network is maximized.

\subsection{Initialization}
The first step in the process consists of characterizing the links between each sensor node and all relays and access point. The minimum power required to maintain each link will be measured and stored(referred to as\emph{link power}). This initialization procedure will be re-run at some (to be determined, possibly integrated into normal operations) interval to deal with topology changes in the network.

\subsection{Routing}
\begin{itemize}
\item Compute \emph{accumulated energy} average from power required for a direct link (i.e. without using relays). This is only done once.

\item Determine \emph{link cost} by:
\begin{enumerate}
\item Add \emph{accumulated energy} to power required to use selected link.
\item Subtract \emph{accumulated energy} average
\end{enumerate}

\item Run Dijkstra's algorithm to select best path

\item Compute new accumulated energy average

\item Determine  \emph{link cost} and continue...

\end{itemize}

%\bibliographystyle{IEEEtran}
%\bibliographystyle{plain}
%\bibliography{alvaro}{}

%\pagebreak
%\listoftodos

\end{document}

### 2011-08-24 gui/itpptest/main.cpp
Added it++ ICA code. Just tested that it compiles, not that it works. Currently able to read in CSV file values and print them on screen.

@@ -1,15 +1,136 @@
 #include <itpp/itbase.h>
+#include <itpp/signal/fastica.h>
+#include <iostream>
 
 using namespace itpp;
 
 using std::cout;
 using std::endl;
 
-int main()
+#define TOTAL_LEADS (12)
+#define INBUFSIZE (1024)
+#define MAX_SAMPLES (1000)
+
+void parse_line( char*, double* );
+uint8_t parse_table( FILE* , double p_value_table[][MAX_SAMPLES] );
+
+int main( int32_t argc, char *argv[] )
 {
+  FILE *fp_in; 
+  double samples[TOTAL_LEADS][MAX_SAMPLES];
   
-  cout << "Making sure we can compile with itpp" << endl;
+  // Make sure the filename is included
+  if ( argc < 2 )
+  {
+    printf( "Usage: %s infile\r\n", argv[0] );
+    return 1;
+  }
+  
+  // Open input csv file
+  fp_in = fopen( argv[1], "r" );
+  
+  if( NULL == fp_in )
+  {
+    printf( "Error opening input file.\r\n" );
+    return 1;
+  }
+  
+  parse_table(fp_in, samples);
+  
+  fclose(fp_in);
+
+#if 0  
+  mat leads_to_reconstruct; // Whichever leads are not used to generate ICs
+  mat leads_initial; // just the 3 leads (I(1),II(2), and V2(8))
+  mat leads_new; // New leads used to generate the ICs for each new beat
+  
+  // Training  
+  Fast_ICA fastica(leads_initial); 
+  fastica.set_nrof_independent_components(3);
+  fastica.set_approach(FICA_APPROACH_DEFL);
+  fastica.separate();
+    
+  mat ICs_initial = fastica.get_independent_components();
+  mat A_initial = fastica.get_mixing_matrix();
+
+  // Generate transform from IC to remaining leads
+  mat reconst_transform = leads_to_reconstruct * ICs_initial.T() * 
+                                  inv(ICs_initial * ICs_initial.T() );
+
+  // Reconstruction
+  Fast_ICA fastica2(leads_new);
+  fastica2.set_nrof_independent_components(3);
+  fastica2.set_approach(FICA_APPROACH_DEFL);
+  fastica2.set_init_guess(A_initial);
+  fastica2.separate();
+  
+  mat ICs_current = fastica2.get_independent_components();
+  mat A_current = fastica2.get_mixing_matrix();
+  mat leads_reconst = reconst_transform * ICs_current;
+#endif
 
[…]

### 2011-08-24 gui/itpptest/main.cpp
Reading csv value into array works. Initial ICA code seems to work. Getting stuck on .separate() during the reconstruction process.

@@ -6,18 +6,18 @@ using namespace itpp;
 
 using std::cout;
 using std::endl;
-
-#define TOTAL_LEADS (12)
+#define TOTAL_LEADS (15)
 #define INBUFSIZE (1024)
-#define MAX_SAMPLES (1000)
 
-void parse_line( char*, double* );
-uint8_t parse_table( FILE* , double p_value_table[][MAX_SAMPLES] );
+void parse_line( char*, mat* );
+uint8_t parse_table( FILE* , mat* );
 
 int main( int32_t argc, char *argv[] )
 {
   FILE *fp_in; 
-  double samples[TOTAL_LEADS][MAX_SAMPLES];
+  mat *samples;
+  
+  samples = new mat;
   
   // Make sure the filename is included
   if ( argc < 2 )
@@ -35,39 +35,89 @@ int main( int32_t argc, char *argv[] )
     return 1;
   }
   
+  // Read in CSV file and populate samples matrix
   parse_table(fp_in, samples);
   
+  // Close input file
   fclose(fp_in);
+  
+  //cout << *samples << endl;
+  
+  mat *leads_initial = new mat;
+  mat *leads_to_reconstruct = new mat;
+  
+  // New leads used to generate the ICs for each new beat
+  mat *leads_new = new mat;
+  
+  // Leads I(0),II(1), and V2(7)
+  leads_initial->append_row( samples->get_col(0) );
+  leads_initial->append_row( samples->get_col(1) );
+  leads_initial->append_row( samples->get_col(7) );
+  
+ 
+ 
+  // Whichever leads are not used to generate ICs
+  leads_to_reconstruct->append_row( samples->get_col(2) );
+  leads_to_reconstruct->append_row( samples->get_col(3) );
+  leads_to_reconstruct->append_row( samples->get_col(4) );
+  leads_to_reconstruct->append_row( samples->get_col(5) );
+  leads_to_reconstruct->append_row( samples->get_col(6) );
+  leads_to_reconstruct->append_row( samples->get_col(8) );
+  leads_to_reconstruct->append_row( samples->get_col(9) );
+  leads_to_reconstruct->append_row( samples->get_col(10) );
+  leads_to_reconstruct->append_row( samples->get_col(11) );
+  leads_to_reconstruct->append_row( samples->get_col(12) );
+  leads_to_reconstruct->append_row( samples->get_col(13) );
+  leads_to_reconstruct->append_row( samples->get_col(14) );
 
[…]

### 2011-08-24 gui/itpptest/main.cpp
Reconstruction code runs. Output is saved to csv file. Need to confirm that it's actually reconstructing the leads correctly...

@@ -8,21 +8,22 @@ using std::cout;
 using std::endl;
 #define TOTAL_LEADS (15)
 #define INBUFSIZE (1024)
+#define MAX_SAMPLES (1000)
 
 void parse_line( char*, mat* );
 uint8_t parse_table( FILE* , mat* );
 
 int main( int32_t argc, char *argv[] )
 {
-  FILE *fp_in; 
+  FILE *fp_in, *fp_out; 
   mat *samples;
   
-  samples = new mat;
+  
   
   // Make sure the filename is included
-  if ( argc < 2 )
+  if ( argc < 4 )
   {
-    printf( "Usage: %s infile\r\n", argv[0] );
+    printf( "Usage: %s infile1 infile2 outfile\r\n", argv[0] );
     return 1;
   }
   
@@ -35,6 +36,7 @@ int main( int32_t argc, char *argv[] )
     return 1;
   }
   
+  samples = new mat;
   // Read in CSV file and populate samples matrix
   parse_table(fp_in, samples);
   
@@ -54,7 +56,6 @@ int main( int32_t argc, char *argv[] )
   leads_initial->append_row( samples->get_col(1) );
   leads_initial->append_row( samples->get_col(7) );
   
- 
  
   // Whichever leads are not used to generate ICs
   leads_to_reconstruct->append_row( samples->get_col(2) );
@@ -87,20 +88,44 @@ int main( int32_t argc, char *argv[] )
   
   mat A_initial = fastica.get_mixing_matrix();
   
-  cout << "done getting mixing matrix" << endl;
-         
+  cout << "done getting mixing matrix" << endl;  
 
   // Generate transform from IC to remaining leads
   mat reconst_transform = (*leads_to_reconstruct) * ICs_initial.T() *
                                   inv( ICs_initial * ICs_initial.T() );
                                   
 
+  delete samples;
+    
+  // Open input csv file
+  fp_in = fopen( argv[2], "r" );
+  
+  if( NULL == fp_in )
+  {
+    printf( "Error opening input file.\r\n" );
+    return 1;
+  }
+  
+  samples = new mat;
+  // Read in CSV file and populate samples matrix
+  parse_table(fp_in, samples);
+  
+  // Close input file
+  fclose(fp_in);
+  
+    // Leads I(0),II(1), and V2(7)
+  leads_new->append_row( samples->get_col(0) );
[…]

### 2011-08-24 gui/itpptest/main.cpp
Changed to static objects instead of many dynamic ones.

@@ -16,10 +16,19 @@ uint8_t parse_table( FILE* , mat* );
 int main( int32_t argc, char *argv[] )
 {
   FILE *fp_in, *fp_out; 
+  
   mat *samples;
+  mat leads_initial;
+  mat leads_to_reconstruct;
+  // New leads used to generate the ICs for each new beat
+  mat leads_new;
+  mat ICs_initial;
+  mat A_initial;
+  mat ICs_current;
+  mat A_current;
+  mat leads_reconst;
   
-  
-  
+
   // Make sure the filename is included
   if ( argc < 4 )
   {
@@ -45,53 +54,48 @@ int main( int32_t argc, char *argv[] )
   
   //cout << *samples << endl;
   
-  mat *leads_initial = new mat;
-  mat *leads_to_reconstruct = new mat;
-  
-  // New leads used to generate the ICs for each new beat
-  mat *leads_new = new mat;
+ 
   
   // Leads I(0),II(1), and V2(7)
-  leads_initial->append_row( samples->get_col(0) );
-  leads_initial->append_row( samples->get_col(1) );
-  leads_initial->append_row( samples->get_col(7) );
+  leads_initial.append_row( samples->get_col(0) );
+  leads_initial.append_row( samples->get_col(1) );
+  leads_initial.append_row( samples->get_col(7) );
   
- 
   // Whichever leads are not used to generate ICs
-  leads_to_reconstruct->append_row( samples->get_col(2) );
-  leads_to_reconstruct->append_row( samples->get_col(3) );
-  leads_to_reconstruct->append_row( samples->get_col(4) );
-  leads_to_reconstruct->append_row( samples->get_col(5) );
-  leads_to_reconstruct->append_row( samples->get_col(6) );
-  leads_to_reconstruct->append_row( samples->get_col(8) );
-  leads_to_reconstruct->append_row( samples->get_col(9) );
-  leads_to_reconstruct->append_row( samples->get_col(10) );
-  leads_to_reconstruct->append_row( samples->get_col(11) );
-  leads_to_reconstruct->append_row( samples->get_col(12) );
-  leads_to_reconstruct->append_row( samples->get_col(13) );
-  leads_to_reconstruct->append_row( samples->get_col(14) );
+  leads_to_reconstruct.append_row( samples->get_col(2) );
+  leads_to_reconstruct.append_row( samples->get_col(3) );
+  leads_to_reconstruct.append_row( samples->get_col(4) );
[…]

### 2011-08-24 gui/itpptest/main.cpp
Now processes entire input csv file MAX_SAMPLES samples at a time. All data is output to outfile.

@@ -11,22 +11,26 @@ using std::endl;
 #define MAX_SAMPLES (1000)
 
 void parse_line( char*, mat* );
-uint8_t parse_table( FILE* , mat* );
+int parse_table( FILE* , mat* );
 
 int main( int32_t argc, char *argv[] )
 {
   FILE *fp_in, *fp_out; 
   
-  mat *samples;
-  mat leads_initial;
-  mat leads_to_reconstruct;
-  // New leads used to generate the ICs for each new beat
-  mat leads_new;
-  mat ICs_initial;
+  mat *samples;             // Matrix with sampled data
+  mat leads_initial;        // Initial sampled leads (I,II, and V2)
+  mat leads_to_reconstruct; // Leads not used to generate ICs
+  mat *leads_new;            // New leads to generate the ICs for each new beat
+  mat ICs_initial;          //
+  mat reconst_transform;    // Transform used to reconstruct 12 leads from 3
   mat A_initial;
   mat ICs_current;
   mat A_current;
   mat leads_reconst;
+  
+  Fast_ICA *fastica_train_reconstruct;
+  
+  int rows, cols;
   
 
   // Make sure the filename is included
@@ -45,16 +49,19 @@ int main( int32_t argc, char *argv[] )
     return 1;
   }
   
+  // Create new matrix to store sampled data
   samples = new mat;
+  
   // Read in CSV file and populate samples matrix
   parse_table(fp_in, samples);
   
   // Close input file
   fclose(fp_in);
-  
-  //cout << *samples << endl;
-  
- 
+    
+  //
+  // Construct leads_initial and leads_to_reconstruct matrices from csv file
+  // data stored in the samples matrix
+  //
   
   // Leads I(0),II(1), and V2(7)
   leads_initial.append_row( samples->get_col(0) );
@@ -75,32 +82,24 @@ int main( int32_t argc, char *argv[] )
   leads_to_reconstruct.append_row( samples->get_col(13) );
   leads_to_reconstruct.append_row( samples->get_col(14) );
 
-
-  cout << "Starting fastica" << endl;
   
-  // Training  
+  //
+  // Begin ICA training
+  //  
   Fast_ICA fastica_train(leads_initial); 
   fastica_train.set_nrof_independent_components(3);
   //fastica_train.set_approach(FICA_APPROACH_DEFL);
-  fastica_train.separate();
-  
-  cout << "done separating" << endl;
-  
[…]

### 2011-08-24 gui/itpptest/main.cpp
Modified code so that output file is only opened for appending new data, instead of leaving it open the whole time.

@@ -116,7 +116,9 @@ int main( int32_t argc, char *argv[] )
   {
     printf( "Error opening output file.\r\n" );
     return 1;
-  }   
+  }
+  
+  fclose(fp_out);
   
   // Free up memory used by the samples matrix
   delete samples;
@@ -127,6 +129,7 @@ int main( int32_t argc, char *argv[] )
   // Read in CSV file and populate samples matrix
   while( parse_table(fp_in, samples) )
   {  
+    
     
     // Leads I(0),II(1), and V2(7)
     leads_new->append_row( samples->get_col(0) );
@@ -149,6 +152,15 @@ int main( int32_t argc, char *argv[] )
 
     cout << "Writing recostructed leads to file" << endl;    
 
+    // Open input csv file
+    fp_out = fopen( argv[3], "a" );
+    
+    if( NULL == fp_out )
+    {
+      printf( "Error opening output file.\r\n" );
+      return 1;
+    }
+
     // Output reconstructed leads to file
     for ( rows = 0; rows < leads_reconst.rows(); rows++ )
     {
@@ -158,6 +170,8 @@ int main( int32_t argc, char *argv[] )
       }
       fprintf(fp_out, "\n" );
     }
+    
+    fclose(fp_out);
     
     delete fastica_train_reconstruct;
     
@@ -172,9 +186,6 @@ int main( int32_t argc, char *argv[] )
   
   // Close input file
   fclose(fp_in);
-  
-  // Close output file
-  fclose(fp_out);
   
   cout << "done" << endl;
 

### 2011-08-25 gui/itpptest/main.cpp
Latest version

@@ -1,5 +1,8 @@
 #include <itpp/itbase.h>
 #include <itpp/signal/fastica.h>
+#include <itpp/signal/sigfun.h>
+#include <itpp/base/math/misc.h>
+#include <itpp/itstat.h>
 #include <iostream>
 
 using namespace itpp;
@@ -27,6 +30,8 @@ int main( int32_t argc, char *argv[] )
   mat ICs_current;
   mat A_current;
   mat leads_reconst;
+    
+  vec *temp;
   
   Fast_ICA *fastica_train_reconstruct;
   
@@ -83,12 +88,14 @@ int main( int32_t argc, char *argv[] )
   leads_to_reconstruct.append_row( samples->get_col(14) );
 
   
+  cout << "Training." << endl;
+  
   //
   // Begin ICA training
   //  
   Fast_ICA fastica_train(leads_initial); 
   fastica_train.set_nrof_independent_components(3);
-  //fastica_train.set_approach(FICA_APPROACH_DEFL);
+  fastica_train.set_approach(FICA_APPROACH_DEFL);
   fastica_train.separate();  
   ICs_initial = fastica_train.get_independent_components();  
   A_initial = fastica_train.get_mixing_matrix();
@@ -99,7 +106,8 @@ int main( int32_t argc, char *argv[] )
   reconst_transform = leads_to_reconstruct * ICs_initial.T() *
                                   inv( ICs_initial * ICs_initial.T() );
                                   
-      
+  cout << "Done training." << endl;
+        
   // Open input csv file
   fp_in = fopen( argv[2], "r" );
   
@@ -128,6 +136,7 @@ int main( int32_t argc, char *argv[] )
   
   // Read in CSV file and populate samples matrix
   while( parse_table(fp_in, samples) )
+  //parse_table(fp_in, samples);
   {  
     
     
@@ -141,14 +150,102 @@ int main( int32_t argc, char *argv[] )
     // Reconstruction
     fastica_train_reconstruct = new Fast_ICA(*leads_new);
     fastica_train_reconstruct->set_nrof_independent_components(3); 
-    //fastica_train_reconstruct.set_approach(FICA_APPROACH_DEFL);
+    //fastica_train_reconstruct->set_approach(FICA_APPROACH_DEFL);
     fastica_train_reconstruct->set_init_guess(A_initial);
[…]

### 2011-09-05 host/lib/routing.c
Added debug file output.

@@ -26,18 +26,57 @@ double watt_to_dbm( double power );
 energy_t target_rssi;
 energy_t rssi_table[MAX_DEVICES+1][MAX_DEVICES+1];
 energy_t link_power_table[MAX_DEVICES+1][MAX_DEVICES+1];
+energy_t link_powers[MAX_DEVICES];
+static energy_t previous_powers[MAX_DEVICES];
+static energy_t previous_powers_debug[MAX_DEVICES];
+uint8_t route_table_debug[MAX_DEVICES];
 
-
-static energy_t previous_powers[MAX_DEVICES];  
-
+FILE *fp_energies, *fp_routes, *fp_powers, *fp_rssi, *fp_debug;
+  
 #define AP_NODE_ID (MAX_DEVICES+1)
 
-void routing_initialize()
+uint8_t routing_initialize( )
 {    
   char node_id_string[3];
   uint8_t node_index;
   
-  target_rssi = dbm_to_watt(-75l);
+  // Open output csv files
+  fp_energies = fopen( "./logs/energies.csv", "w" );  
+  if( NULL == fp_energies )
+  {
+    printf( "Error opening energies file.\r\n" );
+    return 1;
+  }  
+
+  fp_routes = fopen( "./logs/routes.csv", "w" );  
+  if( NULL == fp_routes )
+  {
+    printf( "Error opening routes file.\r\n" );
+    return 1;
+  }
+
+  fp_powers = fopen( "./logs/powers.csv", "w" );  
+  if( NULL == fp_powers )
+  {
+    printf( "Error opening powers file.\r\n" );
+    return 1;
+  }
+
+  fp_rssi = fopen( "./logs/rssi.csv", "w" );  
+  if( NULL == fp_rssi )
+  {
+    printf( "Error opening rssi file.\r\n" );
+    return 1;
+  } 
+  
+  fp_debug = fopen( "./logs/debug.csv", "w" );  
+  if( NULL == fp_debug )
+  {
+    printf( "Error opening debug file.\r\n" );
+    return 1;
+  } 
+  
+  target_rssi = dbm_to_watt(-60l);
   
   // Add nodes 
   sprintf( node_id_string, "AP" );
@@ -65,7 +104,16 @@ void routing_initialize()
   
   printf("Round 0\n");
   
-  return;
+  return 0;
+}
+
+void routing_finalize()
+{
+  fclose( fp_energies );
+  fclose( fp_routes );
[…]

### 2011-09-05 host/lib/routing.h

@@ -80,7 +80,8 @@ static const energy_t power_values[] = {
 -4.30, -3.80, -3.70, -3.50, -3.10, -2.70, -2.30, -1.90, -1.80, -1.30, 
 -0.80, -0.60, -0.40, -0.20, -0.10, +0.00, +0.30, +0.70, +1.10, +1.50 };
 
-void routing_initialize();
+uint8_t routing_initialize();
+void routing_finalize();
 void *compute_routes_thread( void* );
 uint8_t parse_table ( uint8_t p_rssi_table[][MAX_DEVICES+1] );
 

### 2011-09-05 host/threadtest/main.c

@@ -61,7 +61,11 @@ int main( int argc, char *argv[] )
   memset( (uint8_t*)power_table, 0xff, MAX_DEVICES );
   memset( (uint8_t*)routing_table, ( MAX_DEVICES + 1 ), MAX_DEVICES );
   
-  routing_initialize();
+  if ( routing_initialize() )
+  {
+    printf("Error initializing routes.\n");
+    exit(-1);
+  }
   
   rc = pthread_create( &serial_thread, NULL, serial_read_thread, NULL );
   
@@ -86,7 +90,7 @@ int main( int argc, char *argv[] )
     
     // Wait until routing is done
     pthread_mutex_lock ( &mutex_route_done );
-  
+    
     // Send new routes to AP
     send_serial_message( (uint8_t *)rp_tables, sizeof(rp_tables) );
     
@@ -143,6 +147,8 @@ void sigint_handler( int32_t sig )
     
     pthread_cancel( serial_thread );
     pthread_cancel( routing_thread );
+    
+    routing_finalize();
     
     // Close the serial port
     serial_close();

### 2011-09-05 host/threadtest/main.h

@@ -9,7 +9,6 @@
 
 #include <stdint.h>
 
-
 #define BUFFER_SIZE ( 512 )
 
 // Function Prototypes

### 2011-09-05 sim/lib/dijkstra.c

@@ -442,6 +442,10 @@ void compute_shortest_path( uint8_t node_id )
   }
 }
 
+//
+// Store routes in an array and link powers in another
+// If the link does not exist, fill in dummy values
+//
 void compute_rp_tables( uint8_t* route_table, energy_t* link_powers )
 {
   uint8_t node_id;

### 2011-09-06 host/lib/routing.c
Fixed clean_table function. It was previously selecting the worst, instead of the best, link.
The parse_table function now changes the rssi to -999 if it is -136 dBm. This prevents the algorithm from thinking that there was in fact a link present.

@@ -18,7 +18,7 @@ void clean_table( energy_t p_rssi_table[][MAX_DEVICES+1] );
 void add_links_from_table( energy_t p_rssi_table[][MAX_DEVICES+1] );
 void compute_required_powers( energy_t*, uint8_t* );
 void print_rssi_table();
-energy_t get_power_from_setting( uint8_t setting );
+
 uint8_t find_closest_power( energy_t power );
 double dbm_to_watt( double power );
 double watt_to_dbm( double power );
@@ -32,78 +32,78 @@ static energy_t previous_powers_debug[MAX_DEVICES];
 uint8_t route_table_debug[MAX_DEVICES];
 
 FILE *fp_energies, *fp_routes, *fp_powers, *fp_rssi, *fp_debug;
-  
+
 #define AP_NODE_ID (MAX_DEVICES+1)
 
 uint8_t routing_initialize( )
-{    
+{
   char node_id_string[3];
   uint8_t node_index;
-  
+
   // Open output csv files
-  fp_energies = fopen( "./logs/energies.csv", "w" );  
+  fp_energies = fopen( "./logs/energies.csv", "w" );
   if( NULL == fp_energies )
   {
     printf( "Error opening energies file.\r\n" );
     return 1;
-  }  
+  }
 
-  fp_routes = fopen( "./logs/routes.csv", "w" );  
+  fp_routes = fopen( "./logs/routes.csv", "w" );
   if( NULL == fp_routes )
   {
     printf( "Error opening routes file.\r\n" );
     return 1;
   }
 
-  fp_powers = fopen( "./logs/powers.csv", "w" );  
+  fp_powers = fopen( "./logs/powers.csv", "w" );
   if( NULL == fp_powers )
   {
     printf( "Error opening powers file.\r\n" );
     return 1;
   }
 
-  fp_rssi = fopen( "./logs/rssi.csv", "w" );  
+  fp_rssi = fopen( "./logs/rssi.csv", "w" );
   if( NULL == fp_rssi )
   {
     printf( "Error opening rssi file.\r\n" );
     return 1;
-  } 
-  
-  fp_debug = fopen( "./logs/debug.csv", "w" );  
+  }
+
+  fp_debug = fopen( "./logs/debug.csv", "w" );
   if( NULL == fp_debug )
   {
     printf( "Error opening debug file.\r\n" );
     return 1;
-  } 
-  
+  }
+
[…]

### 2011-09-06 host/lib/routing.h

@@ -47,42 +47,43 @@ static const energy_t rssi_values[256] = {
 -74.5,-74.0,-73.5,-73.0,-72.5, };
 
 // Register settings for powers in power_values array
-static const uint8_t power_settings[] = { 
-0x00, 0x40, 0x50, 0x60, 0x70, 0x80, 0x44, 0x90, 0x41, 0x42, 
-0xC0, 0x48, 0x43, 0x54, 0xA0, 0xB0, 0x84, 0x74, 0x51, 0x52, 
-0x53, 0x58, 0x5C, 0xE0, 0x61, 0x81, 0x82, 0x72, 0x88, 0x83, 
-0x8C, 0x94, 0x6C, 0x45, 0xC4, 0x46, 0x47, 0xA4, 0xB4, 0x49, 
-0x91, 0x4E, 0x92, 0xC1, 0xC2, 0x93, 0x4B, 0x4F, 0xC3, 0xC8, 
-0xD4, 0xCC, 0xA1, 0xA2, 0xB2, 0xA3, 0xB3, 0xA8, 0xB8, 0x55, 
-0x56, 0x85, 0x86, 0xD2, 0x87, 0xD3, 0xD8, 0xE1, 0x89, 0x8D, 
-0x8A, 0x8E, 0xE3, 0xF3, 0x65, 0x8B, 0x66, 0x59, 0x76, 0x5A, 
-0x5D, 0xE8, 0x5E, 0xEC, 0xFC, 0x67, 0x77, 0x5B, 0xC5, 0x5F, 
-0xC6, 0xC7, 0x95, 0x96, 0xC9, 0xCA, 0xCE, 0x97, 0xCB, 0xCF, 
-0xB5, 0x6A, 0x6D, 0x99, 0x9D, 0x9A, 0x9E, 0xD5, 0x7E, 0xD6, 
-0x9B, 0x9F, 0xD7, 0x7B, 0x7F, 0xE5, 0xE6, 0xE7, 0xD9, 0xDA, 
-0xDE, 0xDB, 0xDF, 0xAA, 0xAE, 0xBE, 0xAB, 0xAF, 0xBB, 0xBF, 
+static const uint8_t power_settings[] = {
+0x00, 0x40, 0x50, 0x60, 0x70, 0x80, 0x44, 0x90, 0x41, 0x42,
[…]

### 2011-09-06 host/lib/serial.c
Removed trailing whitespace.

@@ -28,16 +28,16 @@ uint8_t serial_open( int32_t port_number, int32_t baud_rate,
                                        void (*callback)( uint8_t*, uint32_t) )
 {
   serial_port_number = port_number;
-  // Use RS232 library to open serial port  
+  // Use RS232 library to open serial port
   if ( OpenComport( serial_port_number, baud_rate ) )
-  {    
+  {
     // Error opening serial port
     return 1;
   }
-  
+
   // Set a callback function for receiving packets
   serial_read_callback = callback;
-  
+
   return 0;
 }
 
@@ -52,30 +52,30 @@ void serial_close()
 
 /*******************************************************************************
  * @fn     void *serial_read_thread()
- * @brief  Run as thread. Polls serial port and builds packets. Calls 
+ * @brief  Run as thread. Polls serial port and builds packets. Calls
  *         serial_callback function with each new packet
  * ****************************************************************************/
 void *serial_read_thread()
 {
-  uint8_t serial_buffer[BUFFER_SIZE]; 
+  uint8_t serial_buffer[BUFFER_SIZE];
   uint8_t packet_buffer[BUFFER_SIZE];
   uint8_t final_buffer[BUFFER_SIZE];
-  
+
   uint16_t buffer_offset = 0;
   uint8_t new_packet = 0;
- 
+
   uint16_t bytes_read = 0;
-  
+
   printf("Starting serial read.\r\n");
-  
+
   // Flush the port
   while( PollComport( serial_port_number, serial_buffer, BUFFER_SIZE ) );
-  
+
   memset( serial_buffer, 0x00, sizeof(serial_buffer) );
-  
+
   // Run forever
   for(;;)
-  {  
+  {
     // Read a buffer full of data (if available)
     bytes_read = PollComport( serial_port_number, serial_buffer, BUFFER_SIZE );
     if( bytes_read > 0 )
@@ -87,23 +87,23 @@ void *serial_read_thread()
       }
       else
       {
-        
-        memcpy( packet_buffer + buffer_offset, serial_buffer, bytes_read );                        
-       
+
+        memcpy( packet_buffer + buffer_offset, serial_buffer, bytes_read );
+
         if ( packet_in_buffer( packet_buffer ) )
-        {          
-                             
+        {
+
           buffer_offset = 0;
-          
+
           new_packet = 1;
-          bytes_read = find_and_escape_packet( packet_buffer, final_buffer );          
[…]

### 2011-09-06 sim/readcsv/main.c

@@ -28,87 +28,87 @@ double watt_to_dbm( double power );
 energy_t target_rssi;
 
 int32_t main ( int32_t argc, char *argv[] )
-{  
-  FILE *fp_in, *fp_out;      
+{
+  FILE *fp_in, *fp_out;
   energy_t rssi_table[MAX_DEVICES+1][MAX_DEVICES+1];
   uint32_t index = 0;
   char node_id_string[3];
   uint8_t node_index;
-  
+
   target_rssi = dbm_to_watt(-75l);
-  
+
   // Make sure the filename is included
   if ( argc < 3 )
   {
     printf( "Usage: %s infile outfile\r\n", argv[0] );
     return 1;
   }
-  
+
   // Open input csv file
   fp_in = fopen( argv[1], "r" );
-  
+
   if( NULL == fp_in )
   {
     printf( "Error opening input file.\r\n" );
     return 1;
   }
-  
+
   // Open output csv file
   fp_out = fopen( argv[2], "w" );
-  
+
   if( NULL == fp_out )
   {
     printf( "Error opening input file.\r\n" );
     return 1;
   }
-  
-  // Add nodes 
+
+  // Add nodes
   for( node_index = 0; node_index < (MAX_DEVICES + 1); node_index++ )
   {
     sprintf( node_id_string, "%d", node_index );
     add_labeled_node( node_index, 0, node_id_string );
   }
-  
+
   initialize_node_energy( 0 );
-  
+
   print_node_energy( 0, fp_out );
-  
-  //parse_table( fp_in, rssi_table ); 
-  while( parse_table( fp_in, rssi_table ) )  
-  {    
-    
-    clean_table( rssi_table );       
+
+  //parse_table( fp_in, rssi_table );
+  while( parse_table( fp_in, rssi_table ) )
+  {
+
+    clean_table( rssi_table );
 
     add_links_from_table( rssi_table );
-    
+
     // Run dijkstra's algorithm with 0 being the access point
     dijkstra( 0 );
-    
+
     // Display shortest paths and update energies
     for( node_index = 1; node_index < (MAX_DEVICES + 1); node_index++ )
     {
       compute_shortest_path( node_index );
       print_shortest_path( node_index );
     }
-    
+
     //print_all_links();
-    
+
     // Compute engergy mean and subtract from each individual mean
-    compute_mean_energy( 0 );    
-    
+    compute_mean_energy( 0 );
+
     printf("Round %d\n", index);
     print_node_energy( 0, fp_out );
-    
+
[…]

### 2011-09-06 sim/readcsv/main.h

@@ -39,37 +39,37 @@ static const double rssi_values[256] = {
 -74.5,-74.0,-73.5,-73.0,-72.5, };
 
 // Register settings for powers in power_values array
-static const uint8_t power_settings[] = { 
-0x00, 0x40, 0x50, 0x60, 0x70, 0x80, 0x44, 0x90, 0x41, 0x42, 
-0xC0, 0x48, 0x43, 0x54, 0xA0, 0xB0, 0x84, 0x74, 0x51, 0x52, 
-0x53, 0x58, 0x5C, 0xE0, 0x61, 0x81, 0x82, 0x72, 0x88, 0x83, 
-0x8C, 0x94, 0x6C, 0x45, 0xC4, 0x46, 0x47, 0xA4, 0xB4, 0x49, 
-0x91, 0x4E, 0x92, 0xC1, 0xC2, 0x93, 0x4B, 0x4F, 0xC3, 0xC8, 
-0xD4, 0xCC, 0xA1, 0xA2, 0xB2, 0xA3, 0xB3, 0xA8, 0xB8, 0x55, 
-0x56, 0x85, 0x86, 0xD2, 0x87, 0xD3, 0xD8, 0xE1, 0x89, 0x8D, 
-0x8A, 0x8E, 0xE3, 0xF3, 0x65, 0x8B, 0x66, 0x59, 0x76, 0x5A, 
-0x5D, 0xE8, 0x5E, 0xEC, 0xFC, 0x67, 0x77, 0x5B, 0xC5, 0x5F, 
-0xC6, 0xC7, 0x95, 0x96, 0xC9, 0xCA, 0xCE, 0x97, 0xCB, 0xCF, 
-0xB5, 0x6A, 0x6D, 0x99, 0x9D, 0x9A, 0x9E, 0xD5, 0x7E, 0xD6, 
-0x9B, 0x9F, 0xD7, 0x7B, 0x7F, 0xE5, 0xE6, 0xE7, 0xD9, 0xDA, 
-0xDE, 0xDB, 0xDF, 0xAA, 0xAE, 0xBE, 0xAB, 0xAF, 0xBB, 0xBF, 
+static const uint8_t power_settings[] = {
+0x00, 0x40, 0x50, 0x60, 0x70, 0x80, 0x44, 0x90, 0x41, 0x42,
[…]

### 2011-09-06 sim/lib/dijkstra.c
Added MAX_LINK_POWER to use when values are out of bounds. Also used to disable a link if the required power is much greater than the maximum tx power. (If the required power is greater than max TX power, the link can still be made, just not with the target RSSI)

@@ -30,18 +30,18 @@ uint8_t add_node( uint8_t node_id, uint8_t is_relay )
   if( s_nodes.current_nodes < MAX_NODES )
   {
     new_node = &s_nodes.nodes[s_nodes.current_nodes];
-    new_node->id = node_id;   
+    new_node->id = node_id;
     new_node->energy = 0;
     new_node->visited = 0;
     new_node->is_relay = is_relay;
-    
+
     s_nodes.current_nodes++;
-    
+
     if( is_relay )
     {
       s_nodes.current_relays++;
     }
-    
+
     return 0;
   }
   else
@@ -57,28 +57,28 @@ uint8_t add_node( uint8_t node_id, uint8_t is_relay )
 uint8_t add_link( uint8_t source, uint8_t destination, energy_t link_power )
 {
   link_t* new_link;
-  
+
   //TODO check if source and destination actually exist!
-    
+
   if( s_links.current_links < MAX_LINKS )
   {
-  
+
     // If the same link already exists, just update it
     new_link = find_link(source, destination);
 
     if( new_link == NULL  )
     {
       // Add new link
-      new_link = &s_links.links[s_links.current_links]; 
-      s_links.current_links++; 
+      new_link = &s_links.links[s_links.current_links];
+      s_links.current_links++;
     }
-    
-    new_link->links_power = link_power; 
+
+    new_link->links_power = link_power;
     new_link->source = source;
     new_link->destination = destination;
 
-    // Disable link if the power required is maximum
-    if( link_power > MAX_LINK_POWER )
+    // Disable link if the power required is too high
+    if( link_power > MAX_LINK_POWER * 100 )
     {
       new_link->active = 0;
     }
@@ -86,7 +86,7 @@ uint8_t add_link( uint8_t source, uint8_t destination, energy_t link_power )
     {
       new_link->active = 1;
     }
-                   
+
     return 0;
   }
   else
@@ -100,12 +100,12 @@ uint8_t add_link( uint8_t source, uint8_t destination, energy_t link_power )
 // Initialize mean energy from direct links to AP
 //
 energy_t initialize_node_energy( uint8_t source_id )
-{  
+{
   node_t* p_node;
   uint8_t link_index;
-  
+
   current_round = 1;
-  
+
   for( link_index = 0; link_index < s_links.current_links; link_index++ )
   {
     if( s_links.links[link_index].source == source_id )
[…]

### 2011-09-06 sim/lib/dijkstra.h

@@ -10,7 +10,7 @@
 #define MAX_NODES (10)
 #define MAX_LINKS (30)
 #define MAX_DISTANCE (1e99)
-#define MAX_LINK_POWER (0.002)
+#define MAX_LINK_POWER (0.001413)
 
 // Use type definition since actual datatype might change
 // (don't want to use floating point on the microcontroller...)
@@ -36,7 +36,7 @@ struct node_s
 struct link_s
 {
   energy_t links_power;   // Constant (how much power does the link require)
-  uint8_t source;         // 
+  uint8_t source;         //
   uint8_t destination;    //
   uint8_t active;         //
 };
@@ -75,4 +75,3 @@ void generate_graph( uint8_t, uint32_t );
 #endif
 
 #endif /* _NODES_H */\
-

### 2011-09-06 host/threadtest/main.c
Added graphing capabilities to program.

@@ -20,9 +20,13 @@
 #endif
 
 void send_serial_message( uint8_t* packet_buffer, int16_t buffer_size );
+void *graph_thread();
 
 pthread_t serial_thread;
 pthread_t routing_thread;
+pthread_t graphing_thread;
+
+pthread_mutex_t mutex_graph;
 
 // Routing and power tables (all in one array)
 static volatile uint8_t rp_tables[MAX_DEVICES * 2];
@@ -33,84 +37,103 @@ static volatile uint8_t *routing_table = &rp_tables[0];
 // Table storing all device transmit powers
 static volatile uint8_t *power_table = &rp_tables[MAX_DEVICES];
 
-
 int main( int argc, char *argv[] )
-{   
+{
   int32_t rc;
-  
+
   // Handle interrupt events to make sure files are closed before exiting
   (void) signal( SIGINT, sigint_handler );
 
   // Make sure input is correct
-  if( argc < 2 )
+  if( argc < 3 )
   {
-    printf("Usage: %s port baudrate\n", argv[0]);
+    printf("Usage: %s port baudrate [graph (0,1)]\n", argv[0]);
     return 0;
-  }  
-  
+  }
+
   printf("Threaded Serial Terminal\r\n");
-  
+
   // Convert string serial number to integer and open port
   if( serial_open( atoi( argv[1] ), atoi( argv[2] ), &process_packet ) )
   {
     printf("Error opening serial port.\r\n");
     exit(-1);
   }
-  
+
   // Initialize routing an power tables
   memset( (uint8_t*)power_table, 0xff, MAX_DEVICES );
   memset( (uint8_t*)routing_table, ( MAX_DEVICES + 1 ), MAX_DEVICES );
-  
+
   if ( routing_initialize() )
   {
     printf("Error initializing routes.\n");
     exit(-1);
   }
-  
+
   rc = pthread_create( &serial_thread, NULL, serial_read_thread, NULL );
-  
+
   if (rc)
   {
     printf("Error creating serial thread\n");
     exit(-1);
   }
-  
-  rc = pthread_create( &routing_thread, NULL, compute_routes_thread, 
+
+  rc = pthread_create( &routing_thread, NULL, compute_routes_thread,
                                                         (void*) routing_table );
-  
+
   if (rc)
   {
     printf("Error creating routing thread\n");
     exit(-1);
   }
-  
+
[…]

### 2011-09-07 docs/thesis/paper.tex
Initial commit of tex files for paper

\documentclass{article}
\begin{document}
\title{Specialized Cost Function for Maximizing Body Area Networks Lifetime Using Global Routing Algorithms}
\author{Alvaro Prieto}
\maketitle

\begin{abstract}
$\ll$ at the end $\gg$
\end{abstract}

\section{Introduction}
Wireless Body Area Networks (WBANs) consist of several wireless sensors located around a human body. These sensors may measure several biological signals, movement, and temperature. Due to major improvements in power consumption and constantly shrinking devices, WBANs are becoming more widely used. Due to the small form factor of these devices, the battery size is limited. While the sensors themselves may be extremely power efficient, all of the measured data must be transmitted over a much less efficient wireless link. One benefit of WBANs is that they rarely include more than a dozen wireless devices over a small area. This constraint allows for the use of routing techniques not suitable for larger wireless networks.

There are several routing protocols designed specifically for WBANs. Braem et al. developed the CICADA protocol, which consists of a spanning tree architecture with a time-division scheme for transmission scheduling~\cite{protocol:CICADA}. The primary downside to this method is that nodes closer to the root will deplete their energy source faster due to the need to relay messages from children nodes. Other protocols focus on the reliability of the network. Constant network changes are a problem with WBANs. If sensors are placed in the extremities, the network will change as the body changes positions. Quwaider et al. developed a protocol tolerant to network changes~\cite{routing:storeandforward}. They propose a store-and-forward method that maximizes the likelihood of a packet reaching its destination. Each packet is stored by multiple devices and retransmitted, which consumes more power. One solution for this problem was proposed by Ehyaie et al.~\cite{relay:networklife}, it consists of using dedicated, non-sensing, relay nodes with larger power sources. While this method increases the network lifetime, it requires more hardware. Finally, Nabi et al. propose a similar store-and-forward method, but integrates Transmit Power Adaptation (TPA)~\cite{relay:transmitpoweradaptation}. Nodes keep track of neighbors and use power control to use the smallest transmit power while maintaining a specific link quality.

In this contribution we propose a novel routing algorithm that focuses not on single devices, but the overall network. Current methods focus on maximizing device lifetime. This is generally a good goal, but certain systems require all devices working to function properly. Keeping that in mind, the proposed system maximizes the network lifetime by normalizing power use across the network.
$\ll$”In this contribution we …. (elaborate on title), describe the algorithm (“in contradistinction to previous routing approached, we define a cost function based on the deviation of the accumulated energy of a single node from the overall average power…”), summary of work (real RSSI data + simulation and hardware implementation using TI 2.4, comparison approach, network lifetime and energy per bit…), summary of results.”$\gg$

\section{Proposed Cost Function}
In the proposed system, we define a new link cost function that is then used in the routing algorithm. Traditionally, the power required to make a link possible is used as the link cost. In this case, the total energy used by a device is factored into the equation. If a device has used more energy than the others, its use as a relay will be discouraged.

Each device's energy use is accumulated as shown in Equation~\ref{eqn:accumulated_energy}, where $j$ denotes the device, $i$ is the current polling round, $\alpha$ is the channel attenuation, and $RSSI^T$ is the target RSSI. The current energy used is incremented by the energy used to transmit a single packet while maintaining $RSSI^T$. The channel attenuation for the selected link,$\alpha$, can be seen in Equation~\ref{eqn:alpha}, where RSSI is the received power and $P_tx$ is the transmitted power. The actual link cost is computed by calculating the energy used by a device \emph{if} that link is selected and subtracting the average energy used across all devices. Equation~\ref{eqn:link_cost} is normalized by the current polling round $i$.
\begin{equation}
\label{eqn:alpha}
\alpha = \frac{RSSI}{P_{tx}}
\end{equation}
\begin{equation}
\label{eqn:accumulated_energy}
E^j_{acc_{(i)}} = E^j_{acc_{(i-1)}} + \frac{RSSI^T}{\alpha}
\end{equation}
\begin{equation}
\label{eqn:link_cost}
C^j_{l,k} = \left[ \left(\frac{ E^l_{acc_{(i-1)}} + \frac{RSSI^T}{\alpha_{l,k}}}{i}\right) - \frac{\sum_jE^l_{acc_{(i-1)}}}{i-1} \right]
\end{equation}
$\ll$describe network (centralized network with powerful AP and multiple sensors able to act as relays, power control), figure 1 (like in paper fig.1 + symbols for channel attenuation, cost, sensor ID,…) and describe symbols in figure.$\gg$
$\ll$Define accumulated energy (normalized power), cost function and all relevant symbols$\gg$
$\ll$rational for cost function$\gg$

\section{Performance Analysis}
$\ll$”in what follows we …. (state simulation and hardware), implementation of cost function in Dijktra (ref),step wide pseudo code for algorithm, comparison approach to “normal” cost function with Dijkstra$\gg$
$\ll$TI platform (in detail with calibration – all data required to replicate) for RSSI and Disjkstra$\gg$

\section{Simulation}
$\ll$describe experimental setups and environment (2 setup – walking in room and sitting in from of desk), figure 2 (residential environment for setups)$\gg$

\section{Hardware Implementation}
The hardware platform used for this test consists of the Texas Instruments (TI) EZ430-RF2500. This device includes both an MSP430F2274 microcontroller along with a CC2500 2.4GHz transceiver. A single device, labeled the Access Point (AP), is connected via a USB to Serial link to the host computer. All other devices, labeled End Devices (EDs), are battery powered. In this implementation, the AP acts only as a bridge between the host and the end devices. All routing and power control calculations are done on the host.

A normal cycle of operations with $n$ devices works as follows:
\begin{enumerate}
\item AP sends synchronization beacon which includes routing and power control tables.
\item Each ED transmits its own RSSI table back to the AP, while simultaneously listening to other ED messages and storing the received power.
\item Once all EDs have transmitted their data, the AP sends a table with the RSSI data from all devices.
\item The host uses the RSSI table to compute the routes along with the required powers to meet the selected links.
\item The host sends both routing and power tables back to the AP so that a new cycle may begin.
\end{enumerate}

To minimize the number of control packets from being transmitted, the EDs are not individually polled. The only control packet that is sent is the synchronization beacon, which also carries the routing and power tables. Once the EDs are synchronized, they transmit on a pre-defined schedule to avoid collisions. Each ED has a network ID ranging from 1 to $n$. The time between synchronization packets is divided into time-slots than are used by each ED. The time slot used depends on the network ID of each device. This removes the need to come up with time schedules during runtime.

The routing table is a simple array which lists the destination for each ED packet. The ED does not need to know the entire route its packets will take, but only next device in the path. Similarly, the power table lists the transmit power setting each device needs to use. The size of these tables is directly proportional to the number of devices in the network.

$\ll$describe change from simulation (RSSI from power controlled transmission, address benefits$\gg$

\section{Results}
$\ll$describe figure  3 and interpret results – network lifetime (like fig.8 in paper with Jeff)$\gg$ 
$\ll$describe figure  4 and interpret results – energy per bit (like fig.9 in paper with Jeff)$\gg$ 
$\ll$* maybe use CDFs?...$\gg$

\section{Conclusion}
$\ll$at the end: different from abstract -  The abstract in past tense + concentrate on meaning of results1

\bibliography{sources/thesis.bib}{}
\bibliographystyle{plain}
\end{document}

### 2011-09-07 docs/thesis/proposal.tex
Initial commit of tex files for paper

\documentclass{article}

%\usepackage[draft]{pdfdraftcopy} %http://sarovar.org/projects/pdfdraftcopy/
%\usepackage{todonotes} %http://tug.ctan.org/pkg/todonotes/
%\usepackage{fullpage}
\usepackage{graphicx}
\begin{document}
\title{Full-body Synchronized Sampling with Power Efficient Relays}
\author{Alvaro Prieto}
\maketitle

\begin{abstract}
This is the thesis proposal for a full-body, synchronized sampling system with power efficient relays. The main goal of the system is to provide a real-time electrocardiogram(ECG) over a low power wireless Body Sensor Network (BSN). 
The primary contribution from this research is a method for synchronized sampling over relayed networks with a focus on energy efficiency. The proposed relaying method consists of a dynamic relaying model with a focus on energy efficiency in order to maximize network lifetime.
\end{abstract}

\pagebreak

\tableofcontents

\pagebreak

\section{The Case for Synchronized Sampling}
Why is synchronized sampling necessary? Until recently, most monitoring systems have been wired. The time difference between two concurrent measurements from interconnected devices is well below the required system tolerance. In most cases, all samples are scheduled, and connected directly, to one single clock source. Unfortunately, this is no longer the case with wireless monitoring systems. The time delay introduced by wireless transmission and reception is, in most cases, not negligible. Another major source of error comes from the introduction of more than one clock sources. In general, each device has its own clock source. Even though the clock sources might be physically similar, they do not oscillate at exactly the same rate. Minute differences in frequency add up over time and, eventually, the clock sources drift apart. Samples that previously occurred simultaneously now are separated by an unknown amount of time. The purpose of synchronized sampling in Wireless Sensor Networks (WSNs) consists of taking the time difference between samples into account and achieving performance similar to that of a wired system.

% Include literature survey from other file
\input{sources/literature_survey}

\section{Project Plan}\label{section:plan}
\subsection{Development Phases}
During the first phase (\ref{section:firstphase}), a working reference system is developed. The second phase (\ref{section:secondphase}) consists of improving the reference system by adding relays(\ref{subsection:relays}) and finer time synchronization(\ref{subsection:synchronization}). The third and final stage (\ref{section:thirdphase}) combines the improvements of the second stage(\ref{subsection:proposed}) and compares the results with current published solutions (\ref{subsection:performance}).

\subsection{Project Timeline}
These are the proposed milestones for the project:

\begin{itemize}
	\item \textbf{April 8} - Have working platform of the reference system.
	\begin{itemize}
		\item Reference system consists of full wireless ECG implementation on non-relaying network.
	\end{itemize}
	\item \textbf{May 6} - Implement reference relaying system.
	\item \textbf{May 27} - Formulate new protocol with energy efficient relaying for network lifetime maximization.
	\item \textbf{June 10} - Implement formulated relaying protocol.
	\item \textbf{June 24} - Compare performance between reference and formulated protocol.
	\item \textbf{August 12} - Submit research thesis.
\end{itemize}

\section{First Phase}\label{section:firstphase}

\subsection{Reference System}
The first step is the development of a working reference system. This system is based on the Imperial College London BSN Motes running TinyOS. Multiple nodes, arranged in a single-hop star topology, act as wireless electrodes for an ECG system. The access point node interfaces with the host computer and display measured data. The goal of the reference system is to have a working base system that can be built on and improved. The reference system should demonstrate a fully working wireless ECG.

\emph{Note: In the first reference system, the access point node interfaces directly with the host computer through a wired connection, reducing initial system complexity and points of failure.}

\subsection{Power Control}
Once the reference system is working, power control is introduced to reduce power consumption and increase network lifetime. Power control allows each wireless device to control Tx power and measure Rx power. This allows each connection to use the minimum required power, decreasing the energy consumed by each transmission.

\section{Second Phase}\label{section:secondphase}
In order to minimize complexity and focus on specific tasks, both sections in the second stage are to occur in parallel. Once working, both will be merged to implement the third, and final stage (\ref{section:thirdphase}).

\subsection{New Synchronization System}\label{subsection:synchronization}
In addition to the single-hop reference system, a novel synchronization method is introduced. The goal is to reduce power consumption, therefore increasing network lifetime, while maintaining an acceptable level of time synchronization between nodes. A detailed description of the proposed synchronization method can be seen in Section~\ref{section:synchronization}.

\subsection{Relaying on Reference System}\label{subsection:relays}
Without worrying about novel time synchronization, the reference system is expanded to support relaying, and increasing network coverage. The goal is to have a working wireless ECG platform over larger networks.  The use or relays in BSNs has been shown to greatly reduce power consumption in various situations~\cite{relay:creepingwave}. Tight integration of power control, relaying, and routing algorithms will increase the overall network lifetime by minimizing power usage.

The relaying system will maximize system lifetime by using centralized power control and routing. The main node, or access point, has some control of the power consumption of each node by deciding which route the data will take. Depending on the route, different Tx power will be used by each node, changing its power consumption. If a node is running low on power, the access point will try to avoid using it as a relay to increase the overall life of the entire system. That node can further decrease power consumption by relaying its data through the closest neighbor, requiring less Tx power. A method to measure the power consumption is explained in Section~\ref{subsection:performance}.

\section{Third Phase}\label{section:thirdphase}

\subsection{Proposed System}\label{subsection:proposed}
Once the novel synchronization method and relaying work independently, they are combined to form the final system. The proposed system includes time synchronization over relayed networks with tight integration of power control.

\subsection{Proposed System Benefitsdynamic relaying model with a focus on energy efficiency in order to maximize network lifetime}
Unlike many WSN applications, the wireless ECG application implemented with a BSN has certain properties that simplify the design. The area over which BSN nodes are distributed is smaller than most WSNs. This means that only few relaying nodes are needed to cover the required area. This physical constraint allows for simple routing algorithms to be used and reduces overall power consumption due to close-by nodes. For the same reason, the network topology rarely changes, minimizing the number of network restructuring events. Another benefit comes in the fact that the wireless ECG application is fully deterministic. There is a constant stream of bio-potential measurements flowing from each node at a known rate, and there are no unscheduled events. This allows more time to be allocated to measurement data, instead of having down time to account for unscheduled interrupts.

\subsection{Performance Comparison}\label{subsection:performance}
Once a working proposed system is obtained, performance comparisons with the reference system are required. This includes comparisons of the reference system using other synchronization methods, like~\cite{synchronization:FTSP}, if possible.

The proposed system does not need to achieve time synchronization as precise as~\cite{synchronization:FTSP}. The goal is to have acceptable synchronization while minimizing power use and therefore maximizing network life. If the system underperforms in time precision yet utilizes less resources, it can be considered a successful improvement for the specific application of wireless ECG.

Power use during transmit/receive and idle times can be measured by using the voltage across a current-sense(precision) resistor placed in series with the power supply~\cite{mac:lowdutycycle}. Fixed size-messages can be transmitted to calculate the energy-per-bit used during transmission. The measurement can then be used to calculate the power consumption as long as the number of transmitted bits is known, without requiring the actual power measurements. By measuring the energy-per-bit at different transmit power levels, a mathematical model for the (radio) energy consumption can be created. This model can then be used to compare various synchronization and relaying methods.

\section{Proposed Synchronization Method}\label{section:synchronization}

\subsection{Function}
The proposed method both compensates for time offsets between samples and minimizes the effect of clock drift by periodically synchronizing independent clock sources together. Figure~\ref{fig:network_layout} shows an example system consisting of three end devices (\emph{ED}s), one access point (\emph{AP}) and one host computer. 
\begin{figure}[htb]
\begin{center}
\includegraphics[width=0.4\textwidth]{figures/network-map-test2.pdf}
\end{center}
\caption{Test Network Layout}
\label{fig:network_layout}
\end{figure}

\subsection{Synchronization}
All devices have a timer used for time-stamping and synchronization. All events, including samples and data transfers, occur according to the value of the timer. Due to certain hardware limitations, the value of the timer cannot be altered, it can only be reset to zero. This does not, in any way, affect the synchronization of devices.

The timer inside the \emph{AP} can be considered the primary, or global, timer. The goal is to synchronize all \emph{ED}, or local, timers with the global timer. Whenever the global timer reaches its maximum value, and rolls over back to zero, the \emph{AP} sends a broadcast synchronization beacon instructing each end device it is time zero. When an \emph{ED} receives this message, it compares its current timer value to zero, to calculate the clock drift, and resets it to zero. This ensures that all \emph{ED} timers are aligned, which then allows measurements to occur at close to the same time.

\subsection{Scheduler}
A round robin-like scheme is used to poll devices in the network. The previous polling system was created as a demonstration for a specific hardware platform and was not as efficient or clean as possible. This new iteration focuses on simplicity and portability.

\begin{figure}[htb]
\begin{center}
\includegraphics[width=1\textwidth]{figures/sync_schedule.pdf}
\end{center}
\caption{System Schedule}
\label{fig:system_schedule}
\end{figure}

Previously, each \emph{ED} was polled for sampled data, which required all \emph{ED}s to be on at all times. Once all device clocks are synchronized, polling is no longer necessary. Instead of sending poll messages to each device, the \emph{AP} just listens for \emph{ED} transmissions. The \emph{ED}s will wake up when they are scheduled to transmit and send the required data automatically. This method greatly reduces the power consumption of both \emph{AP} and \emph{ED}s. After initialization, the \emph{AP} only needs to send a periodic time synchronization beacon and the collected data back to the host computer. Once synchronized, end devices will only need to turn their receiver on around the time synchronization beacon is expected.

Figure~\ref{fig:system_schedule} shows a sample system schedule. The figure shows the time each device is on, and the state of each radio. The ``sleep'' time for each \emph{ED} does not show when a device wakes up to sample data. In this specific case, the \emph{AP} has one extra time-slot allotted for a fourth \emph{ED} to connect and use.

Certain devices can further reduce power consumption by sampling without using the main processor. This works by using direct-memory access(\emph{DMA}) which allows the analog-to-digital converter(\emph{ADC}) to store measurements in memory without first going through the processor. Using this method, Figure~\ref{fig:system_schedule} accurately presents the time each \emph{ED} is powered on.

\section{Proposed Relaying Method}\label{section:relaying}
The proposed relaying method, while not fully defined yet, will utilize some concepts presented by previous work while introducing novel methods. A basic description of the system consists of a dynamic relaying model with a focus on energy efficiency in order to maximize network lifetime. One of the main contributions is the use of the creeping wave model~\cite{relay:creepingwave}, along with power consumption, as a main factor in the routing algorithm. The relaying method will be integrated with the previously described synchronization method to create the final system.

%\bibliographystyle{IEEEtran}
\bibliographystyle{plain}
\bibliography{sources/thesis}{}

%\pagebreak
%\listoftodos

\end{document}

### 2011-09-07 docs/thesis/sources/literature_survey.tex
Initial commit of tex files for paper

\section{Literature Survey}\label{section:litsurvey}
\subsection{Time Synchronization}
Time synchronization has been the subject of research for many years.

There are robust and time-tested synchronization methods, such as the Network Time Protocol (NTP)~\cite{synchronization:NTP}, which are currently in use around the world. Unfortunately these methods were designed for wired systems, and do not perform well once power consumption, latency, and other wireless effects taken into account.

One of the first synchronization protocols designed specifically for wireless networks is Reference Broadcast Synchronization (RBS)~\cite{synchronization:RBS}. The main idea behind RBS is to synchronize all wireless sensor nodes to each other. This creates a local time, within the network of nodes, where all clocks are synchronized. For many applications, sensor nodes do not need to know the actual time, as long as they are synchronized to one another. \cite{synchronization:RBS} lists an example application that measures the time-of-flight of sound. If needed, RBS can be extended to use an external global time source to provide a relate the local and global times.

There are several sources of nondeterminism in a wireless network. ~\cite{synchronization:RBS} and ~\cite{synchronization:FTSP} decompose the sources of nondeterminism into several components. These components account for the delays due to the sending and receiving of a message in a wireless network. These components(from ~\cite{synchronization:RBS}) are: 
\begin{itemize}
\item \emph{Send Time} -- Time the sender takes to construct a message. Includes delays incurred by the operating system and time required to transfer the message from the host to the network interface.
\item \emph{Access Time} -- Time spent waiting for access to transmit channel.
\item \emph{Propagation Time} -- Time taken by the message to physically travel from sender to receiver.
\item \emph{Receive Time} -- Time spent processing a message and notifying the host of its arrival.
\end{itemize}

As suggested by its name, RBS uses broadcast messages to synchronize wireless nodes to one another. RBS does not synchronize a set of wireless nodes to a reference clock. Because of this property, RBS effectively eliminates the \emph{Send Time} and \emph{Access Time} as sources of error. The \emph{Access Time}, while unknown, is the same across all devices for any single message. RBS does not take \emph{Propagation Time} into account and considers it to be effectively zero. This is only a valid assumption if all nodes are equidistant from the broadcast message source. As stated in~\cite{synchronization:VHT}, the main benefit of RBS was due to high transmission nondeterminism. With new radio technology, this is no longer the main problem.

RBS is expandable for multi-hop networks, but has several requirements to do so. The first is that there must be multiple broadcast transmitters in order to cover the whole network. This creates several sub-networks of synchronized nodes, but does not synchronize them all together. The second requirement is that there must be a sort of overlap between broadcast transmitters. A sensor node receiving broadcasts from two different transmitters can compute the offset between the local times of both networks. 
\begin{figure}[htb]
\begin{center}
\includegraphics[width=0.4\textwidth]{figures/RBS_multihop.pdf}
\end{center}
\caption{RBS Multi-Hop Network Layout. From~\cite{synchronization:RBS}.}
\label{fig:RBS_multihop}
\end{figure}

Another time synchronization protocol introduced in~\cite{synchronization:TPSN} is the Timing-sync Protocol for Sensor Networks (TPSN). In contrast to RBS, TPSN synchronizes a set of notes to a single global time source. A hierarchical structure is established before synchronization can take place. Once the structure is in place, the root node begins to synchronize all nodes directly connected to it. Those nodes then do the same for their children nodes, etc...

TPSN utilizes the sender-receiver approach~\cite{synchronization:NTP} to synchronize a pair of nodes. A two-way message exchange provides enough information to calculate the clock difference and propagation delay between a pair of nodes. In order to synchronize the whole network, the root node initiates the process. In order to reduce packet collisions, each child node waits for a random amount of time before synchronizing to the parent. This is then repeated with the next level of nodes until they are all synchronized. Due to the very structured nature of TPSN, special protocols are provided for dealing with dying nodes, as well as newly introduced ones.

One of the costs of using TPSN over RBS is that the \emph{Send Time} is once again a source of error in the system. This is, however, mitigated by closely integrating TPSN at the Medium Access Control (MAC) layer. TPSN requires that both incoming and outgoing messages be timestamped at the MAC level in order to minimize uncertainties in transmit and receive times.

The Flooding-Time Synchronization Protocol (FTSP)~\cite{synchronization:FTSP} is another protocol specifically designed for resource limited wireless platforms. Like TPSN, FTSP relies on MAC-layer time-stamping, along with several other new techniques, to synchronize devices. FTSP uses a single message to synchronize the clocks of to wireless nodes. By comparison, RBS uses a single message to synchronize the clocks of several devices to one-another, but not to the senders clock. TPSN uses two messages to synchronize two device clocks. Because of this, FTSP needs to account to other sources of error, such as the \emph{Byte Alignment Time}. All three protocols are susceptible to the jitter in \emph{Interrupt Handling Time}.

In FTSP, each device sends a broadcast message with its clock time. Devices in the vicinity listen to the broadcast and calculate the difference between their local and received times. In order to reduce jitter, FTSP records multiple times to compute a singe, more accurate time-stamp. This occurs on both the sending and receiving sides.

Another important contribution from FTSP is a method to deal with clock drift. Due to differences in the exact frequency between local clocks, the time-difference between them grows over time.~\cite{synchronization:FTSP} proposes a solution that takes place off-line. In the short-term, the clock drift is fairly linear. Because of the linear fashion of clock drift, linear regression can be used to distribute the error over a period of time. If clocks are being synchronized every \emph{X} seconds, and clocks drift some amount \emph{Y}, the algorithm distributes the drift \emph{Y} over all samples in a \emph{X} second interval. The reason this must be done off-line is because \emph{Y} can only be computed after each synchronization.

FTSP scales well to multi-hop wireless networks. Reference points are utilized to keep track of the time differences between the root node and the local node. FTSP is also resistant against losing nodes as it includes root  re-election protocols in case the root node fails. According to ~\cite{synchronization:VHT}, FTSP is currently the \emph{de facto} time synchronization protocol in wireless sensor networks.

The Gradient Time Synchronization Protocol (GTSP)~\cite{synchronization:GTSP} focuses on synchronization of close-by nodes in multi-hop networks. If two physically nearby nodes are part of two different synchronization trees, their clocks might be different. GTSP uses radio broadcasts overheard from all neighbors in order to provide a more accurate time in relation to close-by nodes.

There are certain disadvantages to each of these methods. RBS requires additional messages for time synchronization across multi-hop networks. TPSN does not account for the clock drift of sensor nodes. Both RBS and TPSN are vulnerable to jitter in interrupt handling and decoding times. While FTSP accounts for interrupt handling jitter, it does not take propagation time into account. This is not a problem if nodes are physically close, but can affect performance in long distance cases. GTSP and RBS are also vulnerable to propagation delay errors.

\subsection{Synchronized Sampling}
There are several applications, such as monitoring an active volcano ~\cite{applications:volcano}, structural monitoring~\cite{applications:structural} and counter sniper systems~\cite{applications:sniper}, which rely on synchronized measurements across sensor networks. Currently there is not much research focusing specifically on synchronized sampling in low-power WSNs. Apart from ~\cite{sampling:earthquake}, most research utilizes the previously described time synchronization protocols to keep device clocks aligned. With synchronized clocks, synchronized sampling is no longer a complex problem.

\subsection{Routing and Relaying}
Due to power and physical constraints, sensor nodes placed along the body may not be able to communicate with the main node directly. Even when sensor nodes are able reach the main node, the power required to do so could be prohibitive. The use of relays to re-transmit the data has been shown to dramatically reduce power consumption in body sensor networks~\cite{relay:creepingwave}.

There are several routing protocols currently used in WSNs~\cite{survey:wirelessrouting}, unfortunately, most are not well suited for BSNs. Several of the protocols focus on mobility, Quality of Service (QoS), scalability and reliability, among other things. In general, BSNs do not have the same requirements as other WSNs. For example, in BSNs, scalability is no longer a problem since networks are limited to a human body. In most cases, power consumption becomes the main limiting factor for overall system performance.

Some routing protocols are specifically designed with BSN constraints in mind. One example is the Wireless Autonomous Spanning tree Protocol (WASP) \cite{protocol:WASP}. Braem et al. describe the WASP as ``a slotted protocol that uses a spanning tree for medium access coordination and traffic routing''. The main idea is that each node sends a ``WASP-scheme'' through a broadcast message that both parent and children nodes receive. This scheme contains the time allotments for each node for the current time frame. The children of the current node need the scheme to know when they are allowed to transmit data. The scheme also informs the parent of the current node as to how many children each node has, which allows it to allocate enough time slots for that branch. This method allows for a distributed time-allocation protocol. Each time frame also includes contention periods where new nodes can join the network. Routing is simple as data from sensor nodes is just sent up the network tree and broadcast messages are used to send configuration down from the root node. The main disadvantage of the WASP is the need to send the scheme packets from the root node down the tree during every time frame.

An improvement over the WASP is the Cascading Information retrieval by Controlling Access with Distributed slot Assignment (CICADA) protocol~\cite{protocol:CICADA}. The improvement over the WASP has to do with the separation of control and data subcycles. The WASP re-configures the time slots during every frame, which produces a significant time delay. The information has to move from the root node to all the children and back. The CICADA protocol configures the network in a separate subcycle and later transfers data. The data transfer is initiated from the bottom of the tree, so there is no delay waiting for the root node to send configurations down during each data subcycle.

Other BSN routing methods focus on postural mobility~\cite{routing:storeandforward}. Quwaider et al. propose a protocol tolerant to constant network changes due to body/sensor movement. When sensor nodes are located in body extremities, the wireless links vary with body position changes. The proposed protocol uses a store-and-forward method to transfer data from sensor nodes to the main node. While the protocol is well suited for dealing with mobile nodes, it requires more energy to do so. ``Hello'' packets are constantly sent by each node to determine what the current neighboring nodes are. A neighbor table is generated using this information and is necessary to determine where packets will be sent to. Another problem with the method is due to multiple copies of packets being transmitted to increase the probability of delivery. Finally, the main node transmits at full power in order to poll each sensor node, increasing its energy consumption.

Most current routing methods for BSNs use network trees. One of the main problems present with network trees is that the nodes closer to the main node deplete their energy supply faster than children nodes. This is due to the fact that all trafic from children nodes has to be relayed through. Ehyaie et al. propose using dedicated relay nodes instead of having sensor nodes act as relays~\cite{relay:networklife}. Using dedicated relays increases the individual sensor life, with the added cost of more devices.

So far, most BSNs have used tree-like networks to join nodes together. Tree networks, while simple, are negatively affected by node failures and movement.  Nabi et al. use a gossiping data routing strategy along with TDMA based MAC to transmit information to the main node \cite{relay:transmitpoweradaptation}. Each node receives and stores the last sample data from every neighboring node. During transmission, it sends its own sampled data, along with one or more of the stored packets from other nodes. This method maximizes the likelihood of a packed reaching the main node, while consuming more power due to retransmissions of the same data. Timestamps are used to decide whether or not a received packed will be stored or discarded. The main node is assumed to have higher battery capacity and transmission range. It is required to be able to reach all nodes with a single transmission. It is in charge of sending a synchronization beacon to use as a time reference for TDMA frames. This method is not always feasible, for example, when two nodes are on opposite sides of the body.

Another valuable contribution from~\cite{relay:transmitpoweradaptation} deals with Transmit Power Adaptation (TPA). The goal for TPA is to minimize power consumption while maintaining a specific link quality. It achieves this goal by adjusting Tx power according to link quality. Unlike the link quality metric in ~\cite{routing:storeandforward}, which was bidirectional, nodes have both inlink and outlink qualities. The outlink metric is then maintained within predefined thresholds by varying Tx power. Inlink quality is stored and later transmitted in order to share the information with neighboring nodes.

\subsection{Similar Work}
In order to determine the improvements of novel methods, reference systems need to be available to compare with. Both~\cite{mac:tdma}~and~\cite{mac:lowdutycycle} provide a suitable reference system for a synchronized, non-relaying BSN. Both proposed systems present, in detail, a TDMA MAC protocol specifically designed with energy efficiency in mind. Details of packet format, TDMA frame calculation, and power consumption measurements are available.
%Add more details later

%[The current system is further simplified by the complete determinism of network traffic. Each sensor node is continuously sampling data at a known rate and sending it back to the main node at known intervals. This allows for protocol simplifications and other energy saving things EXPLAIN!!! ]

### 2011-09-07 sim/lib/dijkstra.c
Fixed algorithm! The link cost function was adding the mean energy instead of subtracting! Increased the maximum number of links.

@@ -378,7 +378,7 @@ uint8_t dijkstra( uint8_t source_id )
           // Calculate the current cost of the link
           current_cost = p_source_node->energy + p_current_link->links_power;
           current_cost /= current_round;
-          current_cost += s_mean_energy /( current_round - 1 );
+          current_cost -= s_mean_energy /( current_round - 1 );
 
 #ifdef REGULAR_DIJKSTRA
           current_cost = p_current_link->links_power;
@@ -757,3 +757,4 @@ void generate_graph( uint8_t source_id, uint32_t file_number )
 }
 
 #endif
+

### 2011-09-07 sim/lib/dijkstra.h

@@ -8,7 +8,7 @@
 #define _NODES_H
 
 #define MAX_NODES (10)
-#define MAX_LINKS (30)
+#define MAX_LINKS (100)
 #define MAX_DISTANCE (1e99)
 #define MAX_LINK_POWER (0.001413)
 
@@ -74,4 +74,5 @@ void print_all_nodes( uint8_t );
 void generate_graph( uint8_t, uint32_t );
 #endif
 
-#endif /* _NODES_H */\
+#endif /* _NODES_H */
+

### 2011-09-07 host/lib/routing.h
Added newlines at end of file that were mistakenly removed.

@@ -87,3 +87,4 @@ uint8_t parse_table ( uint8_t p_rssi_table[][MAX_DEVICES+1] );
 energy_t get_power_from_setting( uint8_t setting );
 
 #endif /*_ROUTING_H */
+

### 2011-09-07 host/lib/serial.c

@@ -254,3 +254,4 @@ void dummy_callback( uint8_t* buffer, uint32_t size )
 {
   // Do nothing
 }
+

### 2011-09-07 host/lib/serial.h

@@ -17,3 +17,4 @@ void *serial_read_thread();
 void serial_close();
 
 #endif /* _SERIAL_H */
+

### 2011-09-07 host/threadtest/main.c

@@ -263,3 +263,4 @@ void sigint_handler( int32_t sig )
     printf("\nExiting...\n");
     exit(sig);
 }
+

### 2011-09-07 host/threadtest/main.h

@@ -17,3 +17,4 @@ void process_packet( uint8_t*, uint32_t );
 void sigint_handler( int32_t sig );
 
 #endif
+

### 2011-09-07 host/lib/routing.c
Removed debug printout

@@ -119,17 +119,11 @@ void routing_finalize()
 void *compute_routes_thread( void *rp_tables )
 {
   static uint32_t index;
+  uint8_t node_index;
   uint8_t *route_table = &((uint8_t*)rp_tables)[0];
   uint8_t *power_table = &((uint8_t*)rp_tables)[MAX_DEVICES];
-
-  uint8_t node_index;
-  for( node_index = 0; node_index < MAX_DEVICES; node_index++ )
-    {
-
-      printf("%5.1f ", previous_powers[node_index] );
-    }
-    printf("<--\n");
-
+  
+  
   // loop forever
   for (;;)
   {
@@ -161,23 +155,11 @@ void *compute_routes_thread( void *rp_tables )
     // Compute power table
     compute_required_powers( link_powers, power_table );
 
-    for( node_index = 0; node_index < MAX_DEVICES; node_index++ )
-    {
-      //printf("%02X ", power_table[node_index] );
-      printf("%g(0x%02X) ", previous_powers[node_index], power_table[node_index] );
-    }
-    printf("\n");
-
-    //print_all_links();
-
     // Compute engergy mean and subtract from each individual mean
     compute_mean_energy( AP_NODE_ID );
 
-    //print_node_energy( 0, fp_out );
-
-    //generate_graph( 0, index );
-
     print_rssi_table();
+
     print_node_energy( AP_NODE_ID, fp_energies );
 
     index++;
@@ -319,16 +301,12 @@ void compute_required_powers( energy_t* p_link_powers, uint8_t* power_table )
     power_table[node_index] =
                 find_closest_power( watt_to_dbm( p_link_powers[node_index] ) );
 
-    printf("%g ", watt_to_dbm( p_link_powers[node_index] ));
-
     // Save current required power to be used as tx_power next round
     previous_powers[node_index] =
                             //get_power_from_setting( 0xff );  // no power control
                             get_power_from_setting( power_table[node_index] );
 
-
   }
-  printf("\n");
 }
 
 void print_rssi_table()
@@ -340,51 +318,51 @@ void print_rssi_table()
 
   for( col_index = 0; col_index <= MAX_DEVICES; col_index++ )
   {
-    printf(" %3d   ", col_index );
+    //printf(" %3d   ", col_index );
   }
 
-  printf("\r\n");
+  //printf("\r\n");
 
[…]

### 2011-09-08 docs/thesis/paper.tex
Added changes suggested by Dr. Tsouri. Still needs tons of work.

@@
 These sensors may measure several biological signals, movement, and temperature.
-Due to major improvements in power consumption and constantly shrinking devices, WBANs are becoming more widely used.
+Due to major improvements in power consumption and constantly shrinking devices, WBANs are becoming ubiquitous.
 Due to the small form factor of these devices, the battery size is limited.
@@
 One benefit of WBANs is that they rarely include more than a dozen wireless devices over a small area.
-This constraint allows for the use of routing techniques not suitable for larger wireless networks.
+This constraint allows for the use of routing techniques not suitable for larger wireless sensor networks.
 
@@
 The primary downside to this method is that nodes closer to the root will deplete their energy source faster due to the need to relay messages from children nodes.
-Other protocols focus on the reliability of the network.
-Constant network changes are a problem with WBANs.
+Other protocols focus on the reliability of the network$\ll$citation needed$\gg$.
+Dynamic networks are a problem with WBANs.
 If sensors are placed in the extremities, the network will change as the body changes positions.
@@
 One solution for this problem was proposed by Ehyaie et al.~\cite{relay:networklife}, it consists of using dedicated, non-sensing, relay nodes with larger power sources.
-While this method increases the network lifetime, it requires more hardware.
+While this method increases the network lifetime$\ll$define term$\gg$, it requires more hardware.
 Finally, Nabi et al.
@@
 
-Each device's energy use is accumulated as shown in Equation~\ref{eqn:accumulated_energy}, where $j$ denotes the device, $i$ is the current polling round, $\alpha$ is the channel attenuation, and $RSSI^T$ is the target RSSI.
+Each device's energy use is accumulated as shown in Equation~\ref{eqn:accumulated_energy}, where $j$ denotes the device ID, $i$ is the current polling round, $\alpha$ is the channel attenuation, and $RSSI^T$ is the target RSSI.
 The current energy used is incremented by the energy used to transmit a single packet while maintaining $RSSI^T$.
@@
 \end{equation}
+
+Once all link costs are calculated, Dijkstra's algorithm~\cite{dijkstra:algorithm} is used to compute the shortest path.
+
+This algorithm works under several assumptions.
+The first is that this is a centralized network with one central Access Point (AP) with virtually unlimited energy and  computing power.
+The second is that each wireless sensor, or End Device(ED), can act as a relay for messages from other nodes.
+The Third and last requirement is that the link quality,or channel attenuation, between all EDs and AP is known.
+All routing calculations take place in the AP.
+
 $\ll$describe network (centralized network with powerful AP and multiple sensors able to act as relays, power control), figure 1 (like in paper fig.1 + symbols for channel attenuation, cost, sensor ID,…) and describe symbols in figure.$\gg$
+
 $\ll$Define accumulated energy (normalized power), cost function and all relevant symbols$\gg$
+
 $\ll$rational for cost function$\gg$
@@
 (state simulation and hardware), implementation of cost function in Dijktra (ref),step wide pseudo code for algorithm, comparison approach to “normal” cost function with Dijkstra$\gg$
+
 $\ll$TI platform (in detail with calibration – all data required to replicate) for RSSI and Disjkstra$\gg$
@@
 \section{Results}
+
 $\ll$describe figure  3 and interpret results – network lifetime (like fig.8 in paper with Jeff)$\gg$
+
 $\ll$describe figure  4 and interpret results – energy per bit (like fig.9 in paper with Jeff)$\gg$
+
 $\ll$* maybe use CDFs?...$\gg$

### 2011-09-08 host/lib/routing.c
Moved graphing mutex initialization to BEFORE thread is created. Added parse_table_d to make it easier to load rssi table from file. Added extra newline at the end of each rssi line...

@@ -176,7 +176,7 @@ void *compute_routes_thread( void *rp_tables )
 /*******************************************************************************
  * @fn    uint8_t parse_table ( int8_t p_rssi_table[][MAX_DEVICES+1] )
  *
- * @brief Get RSSI table, convert, store and print it
+ * @brief Get RSSI table, convert, store and print it(gets uint8_t rssi array)
  * ****************************************************************************/
 uint8_t parse_table ( uint8_t p_rssi_table[][MAX_DEVICES+1] )
 {
@@ -202,6 +202,53 @@ uint8_t parse_table ( uint8_t p_rssi_table[][MAX_DEVICES+1] )
                                rssi_values[p_rssi_table[row_index][col_index]];
       }
 
+      if( 0 == col_index )
+      {
+        // AP always transmits with max power
+        tx_power = dbm_to_watt( get_power_from_setting( 0xff ) );
+      }
+      else
+      {
+        // Use previous transmit settings
+        tx_power = dbm_to_watt( previous_powers[col_index + 1] );
+      }
+
+      //tx_power = dbm_to_watt( 1.5l ); // uncomment for no power control
+
+      // Compute the minimum power required to meet this link with 'target_rssi'
+      // alpha is the channel attenuation, that is received/transmitted power
+      alpha = dbm_to_watt( rssi_table[row_index][col_index] ) / tx_power;
+
+      // Transmit power required is the target rssi / channel attenuation
+      link_power_table[row_index][col_index] = target_rssi / alpha;
+
+    }
+  }
+
+  // Did not read table successfully
+  return 0;
+}
+
+/*******************************************************************************
+ * @fn    uint8_t parse_table_d ( energy_t p_rssi_table[][MAX_DEVICES+1] )
+ *
+ * @brief Get RSSI table, convert, store and print it (gets energy_t rssi array)
+ * ****************************************************************************/
+uint8_t parse_table_d ( energy_t p_rssi_table[][MAX_DEVICES+1] )
+{
+  uint8_t row_index;
+  uint8_t col_index;
+  energy_t tx_power;
+  energy_t alpha;
+
+  // Convert table to rssi values from raw data and copy to local array
+  for( row_index = 0; row_index <= MAX_DEVICES; row_index++ )
+  {
+    for( col_index = 0; col_index <= MAX_DEVICES; col_index++ )
+    {
+      // Copy rssi table
[…]

### 2011-09-08 host/lib/routing.h

@@ -84,6 +84,7 @@ uint8_t routing_initialize();
 void routing_finalize();
 void *compute_routes_thread( void* );
 uint8_t parse_table ( uint8_t p_rssi_table[][MAX_DEVICES+1] );
+uint8_t parse_table_d ( energy_t p_rssi_table[][MAX_DEVICES+1] );
 energy_t get_power_from_setting( uint8_t setting );
 
 #endif /*_ROUTING_H */

### 2011-09-08 host/threadtest/main.c

@@ -87,18 +87,18 @@ int main( int argc, char *argv[] )
     exit(-1);
   }
 
+  pthread_mutex_init( &mutex_graph, NULL );
+
+  // Start mutex locked
+  pthread_mutex_lock ( &mutex_graph );
+
   rc = pthread_create( &graphing_thread, NULL, graph_thread, NULL );
 
   if (rc)
   {
     printf("Error creating serial thread\n");
     exit(-1);
-  }
-
-  pthread_mutex_init( &mutex_graph, NULL );
-
-  // Start mutex locked
-  pthread_mutex_lock ( &mutex_graph );
+  }  
 
   for(;;)
   {

### 2011-09-08 host/lib/routing.c
Modified sim/readcsv/main.c to read rssi file dumped from host/lib/routing.c Verified that it produces the same results every time.

@@ -361,7 +361,7 @@ void print_rssi_table()
   uint8_t row_index;
   uint8_t col_index;
 
-  printf("   ");
+  //printf("   ");
 
   for( col_index = 0; col_index <= MAX_DEVICES; col_index++ )
   {

### 2011-09-08 sim/readcsv/main.c

@@ -1,6 +1,6 @@
 /** @file main.c
 *
-* @brief Read CSV file with RSSI table and generate links with it
+* @brief Read CSV file with RSSI table and simulate routing algorithm
 *
 * @author Alvaro Prieto
 */
@@ -8,39 +8,52 @@
 #include <stdint.h>
 #include <stdlib.h>
 #include <string.h>
-#include <math.h>
-
+#include <signal.h>
+#include <unistd.h>
+#include <pthread.h>
+#include "routing.h"
 #include "dijkstra.h"
-
 #include "main.h"
 
-#define INBUFSIZE (512)
-#define MAX_DEVICES (6)
+#define INBUFSIZE (4096)
 
-void parse_line( char*, energy_t* );
-uint8_t parse_table( FILE* , energy_t p_rssi_table[][MAX_DEVICES+1] );
-void clean_table( energy_t p_rssi_table[][MAX_DEVICES+1] );
-void add_links_from_table( energy_t p_rssi_table[][MAX_DEVICES+1] );
+#ifndef MAX_DEVICES
+#define MAX_DEVICES (3)
+#warning MAX_DEVICES not defined, defaulting to 3
+#endif
 
-double dbm_to_watt( double power );
-double watt_to_dbm( double power );
+void read_line( char*, energy_t* );
+uint8_t read_table( FILE* , energy_t p_rssi_table[][MAX_DEVICES+1] );
+void *graph_thread();
+void sigint_handler( int32_t sig );
 
-energy_t target_rssi;
+pthread_t routing_thread;
+pthread_t graphing_thread;
+
+pthread_mutex_t mutex_graph;
+
+// Routing and power tables (all in one array)
+static volatile uint8_t rp_tables[MAX_DEVICES * 2];
+
+// Table storing all device routes
+static volatile uint8_t *routing_table = &rp_tables[0];
+
+// Table storing all device transmit powers
+static volatile uint8_t *power_table = &rp_tables[MAX_DEVICES];
 
 int32_t main ( int32_t argc, char *argv[] )
 {
-  FILE *fp_in, *fp_out;
+  FILE *fp_in;
+  int32_t rc;
   energy_t rssi_table[MAX_DEVICES+1][MAX_DEVICES+1];
-  uint32_t index = 0;
-  char node_id_string[3];
-  uint8_t node_index;
 
-  target_rssi = dbm_to_watt(-75l);
+  // Handle interrupt events to make sure files are closed before exiting
[…]

### 2011-09-08 sim/readcsv/main.h

@@ -8,68 +8,5 @@
 #define _MAIN_H
 
 
-
-// Lookup table for converting cc2500 rssi value to received power in dBm
-static const double rssi_values[256] = {
--72.0,-71.5,-71.0,-70.5,-70.0,-69.5,-69.0,-68.5,-68.0,-67.5,-67.0,
--66.5,-66.0,-65.5,-65.0,-64.5,-64.0,-63.5,-63.0,-62.5,-62.0,
--61.5,-61.0,-60.5,-60.0,-59.5,-59.0,-58.5,-58.0,-57.5,-57.0,
--56.5,-56.0,-55.5,-55.0,-54.5,-54.0,-53.5,-53.0,-52.5,-52.0,
--51.5,-51.0,-50.5,-50.0,-49.5,-49.0,-48.5,-48.0,-47.5,-47.0,
--46.5,-46.0,-45.5,-45.0,-44.5,-44.0,-43.5,-43.0,-42.5,-42.0,
--41.5,-41.0,-40.5,-40.0,-39.5,-39.0,-38.5,-38.0,-37.5,-37.0,
--36.5,-36.0,-35.5,-35.0,-34.5,-34.0,-33.5,-33.0,-32.5,-32.0,
--31.5,-31.0,-30.5,-30.0,-29.5,-29.0,-28.5,-28.0,-27.5,-27.0,
--26.5,-26.0,-25.5,-25.0,-24.5,-24.0,-23.5,-23.0,-22.5,-22.0,
--21.5,-21.0,-20.5,-20.0,-19.5,-19.0,-18.5,-18.0,-17.5,-17.0,
--16.5,-16.0,-15.5,-15.0,-14.5,-14.0,-13.5,-13.0,-12.5,-12.0,
--11.5,-11.0,-10.5,-10.0,-9.5,-9.0,-8.5,-136.0,-135.5,-135.0,
--134.5,-134.0,-133.5,-133.0,-132.5,-132.0,-131.5,-131.0,-130.5,-130.0,
--129.5,-129.0,-128.5,-128.0,-127.5,-127.0,-126.5,-126.0,-125.5,-125.0,
--124.5,-124.0,-123.5,-123.0,-122.5,-122.0,-121.5,-121.0,-120.5,-120.0,
--119.5,-119.0,-118.5,-118.0,-117.5,-117.0,-116.5,-116.0,-115.5,-115.0,
[…]

### 2011-09-08 results/scripts/csv2dat.py
Adding file conversion and plotting scripts.

@@ -1,0 +1,30 @@
+#!/usr/bin/env python
+# This script takes a list of files as an input (it only takes    
+# files named 'energies.csv') Change if needed...
+# It replaces commas with spaces and ads a time as the first column
+# assuming a 5 Hz sampling rate
+# sample usage: find . | ./csv2dat.py
+
+import csv
+import fileinput
+for line in fileinput.input():
+    if 'energies.csv' in line:
+      line = line.replace('\n','')
+      print "Processing ", line
+      #open output file
+      outfile = open(line.replace('csv','dat'),'w')
+
+      #open csv file to read
+      with open(line, 'rb') as f:
+        reader = csv.reader(f)
+        x = 0
+        for row in reader:
+          outfile.write(str(x*.2))
+          outfile.write(' ')
+          for item in row:
+            outfile.write(item)
+            outfile.write(' ')
+          x = x + 1
+          outfile.write('\n')
+      outfile.close()
+

### 2011-09-12 host/lib/routing.c
Added c_factor to Dijkstra's algorithm. Modified parse_table_d to take previous power input. Removed all instances of mean_energy. Added find_min_energy calculation (needs updating) Implemented new cost function. Modified readcsv/main.c to read power tables from file. Added sample limit to read only 'sample_limit' samples from file.

@@ -31,11 +31,18 @@ static energy_t previous_powers[MAX_DEVICES];
 static energy_t previous_powers_debug[MAX_DEVICES];
 uint8_t route_table_debug[MAX_DEVICES];
 
+energy_t c_factor;
+
 FILE *fp_energies, *fp_routes, *fp_powers, *fp_rssi, *fp_debug;
 
 #define AP_NODE_ID (MAX_DEVICES+1)
 
-uint8_t routing_initialize( )
+/*******************************************************************************
+ * @fn    uint8_t routing_initialize( energy_t dijkstra_c_factor )
+ *
+ * @brief Open all debugging files and initialize routing system
+ * ****************************************************************************/
+uint8_t routing_initialize( energy_t dijkstra_c_factor )
 {
   char node_id_string[3];
   uint8_t node_index;
@@ -76,6 +83,9 @@ uint8_t routing_initialize( )
     return 1;
   }
 
+  // Store c_factor for later use
+  c_factor = dijkstra_c_factor;
+
   target_rssi = dbm_to_watt(-60l);
 
   // Add nodes
@@ -107,6 +117,11 @@ uint8_t routing_initialize( )
   return 0;
 }
 
+/*******************************************************************************
+ * @fn    void routing_finalize()
+ *
+ * @brief Call when finished. Close all files.
+ * ****************************************************************************/
 void routing_finalize()
 {
   fclose( fp_energies );
@@ -116,6 +131,11 @@ void routing_finalize()
   fclose( fp_debug );
 }
 
+/*******************************************************************************
+ * @fn    void *compute_routes_thread( void *rp_tables )
+ *
+ * @brief Thread that takes care of routing
+ * ****************************************************************************/
 void *compute_routes_thread( void *rp_tables )
 {
   static uint32_t index;
@@ -136,7 +156,7 @@ void *compute_routes_thread( void *rp_tables )
     add_links_from_table( rssi_table );
 
     // Run dijkstra's algorithm with 0 being the access point
-    dijkstra( AP_NODE_ID );
+    dijkstra( AP_NODE_ID, c_factor );
 
     // Display shortest paths and update energies
     for( node_index = 1; node_index < (MAX_DEVICES + 1); node_index++ )
@@ -154,9 +174,6 @@ void *compute_routes_thread( void *rp_tables )
 
     // Compute power table
     compute_required_powers( link_powers, power_table );
-
-    // Compute engergy mean and subtract from each individual mean
-    compute_mean_energy( AP_NODE_ID );
 
[…]

### 2011-09-12 host/lib/routing.h

@@ -80,11 +80,12 @@ static const energy_t power_values[] = {
 -4.30, -3.80, -3.70, -3.50, -3.10, -2.70, -2.30, -1.90, -1.80, -1.30,
 -0.80, -0.60, -0.40, -0.20, -0.10, +0.00, +0.30, +0.70, +1.10, +1.50 };
 
-uint8_t routing_initialize();
+uint8_t routing_initialize( energy_t );
 void routing_finalize();
 void *compute_routes_thread( void* );
 uint8_t parse_table ( uint8_t p_rssi_table[][MAX_DEVICES+1] );
-uint8_t parse_table_d ( energy_t p_rssi_table[][MAX_DEVICES+1] );
+uint8_t parse_table_d ( energy_t p_rssi_table[][MAX_DEVICES+1], 
+                                      energy_t *p_previous_powers );
 energy_t get_power_from_setting( uint8_t setting );
 
 #endif /*_ROUTING_H */

### 2011-09-12 sim/lib/dijkstra.c

@@ -8,6 +8,7 @@
 #include <stdlib.h>
 #include <string.h>
 #include <stdint.h>
+#include <math.h>
 #include "dijkstra.h"
 
 static nodes_t s_nodes;
@@ -31,7 +32,7 @@ uint8_t add_node( uint8_t node_id, uint8_t is_relay )
   {
     new_node = &s_nodes.nodes[s_nodes.current_nodes];
     new_node->id = node_id;
-    new_node->energy = 0;
+    new_node->energy = MAX_LINK_POWER;
     new_node->visited = 0;
     new_node->is_relay = is_relay;
 
@@ -97,54 +98,38 @@ uint8_t add_link( uint8_t source, uint8_t destination, energy_t link_power )
 }
 
 //
-// Initialize mean energy from direct links to AP
+// Initialize node energy from direct links to AP
 //
 energy_t initialize_node_energy( uint8_t source_id )
 {
-  node_t* p_node;
-  uint8_t link_index;
+  uint8_t node_index;
 
   current_round = 1;
 
-  for( link_index = 0; link_index < s_links.current_links; link_index++ )
+  for( node_index = 0; node_index < s_nodes.current_nodes; node_index++ )
   {
-    if( s_links.links[link_index].source == source_id )
+    if( s_nodes.nodes[node_index].id != source_id )
     {
-      p_node = find_node( s_links.links[link_index].destination );
-
-#ifdef DEBUG_ON
-  //print_node_name( p_node->id );
-#endif
-      if ( !p_node->is_relay )
-      {
-        p_node->energy = s_links.links[link_index].links_power;
-        s_mean_energy += p_node->energy;
-#ifdef DEBUG_ON
-  //printf("(%g)", p_node->energy);
-#endif
-      }
-#ifdef DEBUG_ON
-  //printf("\n");
-#endif
+      s_nodes.nodes[node_index].energy = MAX_LINK_POWER; 
     }
   }
-
-  // Compute actual mean (subtracting 1 to account for source)
-  s_mean_energy /= (s_nodes.current_nodes - 1 - s_nodes.current_relays);
-
-#ifdef DEBUG_ON
-  //printf("Mean Energy: %g\n", s_mean_energy);
-#endif
 
   return s_mean_energy;
 }
 
 //
-// Re-calculate mean energy
+// Find node using the last amount of energy
 //
-energy_t compute_mean_energy( uint8_t source_id )
[…]

### 2011-09-12 sim/lib/dijkstra.h

@@ -57,8 +57,8 @@ typedef struct
 uint8_t add_node( uint8_t, uint8_t );
 uint8_t add_link( uint8_t, uint8_t, energy_t );
 energy_t initialize_node_energy( uint8_t source_id );
-energy_t compute_mean_energy( uint8_t source_id );
-uint8_t dijkstra( uint8_t );
+energy_t find_min_energy( uint8_t source_id );
+uint8_t dijkstra( uint8_t, energy_t );
 void compute_shortest_path( uint8_t node_id );
 void compute_rp_tables( uint8_t*, energy_t* );
 

### 2011-09-13 results/scripts/plotcsv.py
Adding whole plotting script in python. Will take csv files in and plot in SVG and PNG

@@ -1,0 +1,63 @@
+#!/usr/bin/env python
+# This script takes a list of files as an input (it only takes    
+# files named 'energies.csv') Change if needed...
+# finds all energies.csv files in the data directory and plots in svg and png
+# sample usage: find data/ | ./plotcsv.py
+#
+
+import csv
+import fileinput
+import string
+import os
+max_devices = 8
+sample_rate = 5 # in Hz
+for line in fileinput.input():
+    if 'energies.csv' in line:      
+      line = line.replace('\n','')
+      dirs = line.rsplit('/')
+      #print(dirs[len(dirs)-2])
+      
+      datafilename = 'tmp/'+ dirs[len(dirs)-2] + '.dat'
+      scriptfilename = 'tmp/'+ dirs[len(dirs)-2] + '.scr'
+      
+      # generate data file in gnuplot format
+      datafile = open(datafilename,'w')
+
+      #open csv file to read
+      with open(line, 'rb') as f:
+        reader = csv.reader(f)
+        x = 0
+        for row in reader:
+          datafile.write(str(x/sample_rate))
+          datafile.write(' ')
+          for item in row:
+            datafile.write(item)
+            datafile.write(' ')
+          x = x + 1
+          datafile.write('\n')
+      datafile.close()
+      
+      # generate gnuplot script to process file
+      print "Processing ", line
+      scriptfile = open( scriptfilename,'w')
+      
+      scriptfile.write("# gnuplot script for '" + datafilename + "'\n")
+      scriptfile.write("set terminal svg size 640,480 dynamic enhanced fname 'arial'\n")
+      scriptfile.write("set output 'plots/svg/" + dirs[len(dirs)-2] + ".svg'\n")
+      scriptfile.write("set title '" + dirs[len(dirs)-2] +  "'\n")
+      scriptfile.write("set xlabel 'Time (s)'\n")
+      scriptfile.write("set ylabel 'Power Used (mW)'\n")
+      scriptfile.write("plot '" + datafilename + "' using 1:2 with lines title 'Average'\n")
+      for ed in range(1,max_devices+1):
+        scriptfile.write("replot '" + datafilename + "' using 1:" + str(ed + 2) + " with lines title 'ED" + str(ed) + "'\n")
+      
+      scriptfile.write("set terminal png size 640,480\n")
+      scriptfile.write("set output 'plots/png/" + dirs[len(dirs)-2] + ".png'\n")
[…]

### 2011-09-13 host/lib/serial.c
Updatedn threadtest to sample up to MAX_ROUNDS. Added timeout before sampling starts, to allow for hardware setup. Plays music using banshee and stops it when sampling is finished. Added serial port flush on error in /lib/serial.c

@@ -12,20 +12,20 @@ static int32_t serial_port_number;
 
 uint8_t packet_in_buffer( uint8_t* );
 uint16_t find_and_escape_packet( uint8_t*, uint8_t* );
-static void dummy_callback( uint8_t* buffer, uint32_t size );
+static uint8_t dummy_callback( uint8_t* buffer, uint32_t size );
 
-static void (*serial_read_callback)( uint8_t*, uint32_t ) = dummy_callback;
+static uint8_t (*serial_read_callback)( uint8_t*, uint32_t ) = dummy_callback;
 
 #define SYNC_BYTE   ( 0x7E )
 #define ESCAPE_BYTE ( 0x7D )
 
 /*******************************************************************************
  * @fn     uint8_t serial_open( int32_t port_number, int32_t baud_rate,
- *                                     void (*callback)( uint8_t*, uint32_t) )
+ *                                     uint8_t (*callback)( uint8_t*, uint32_t) )
  * @brief  Open serial port
  * ****************************************************************************/
 uint8_t serial_open( int32_t port_number, int32_t baud_rate,
-                                       void (*callback)( uint8_t*, uint32_t) )
+                                       uint8_t (*callback)( uint8_t*, uint32_t) )
 {
   serial_port_number = port_number;
   // Use RS232 library to open serial port
@@ -115,8 +115,17 @@ void *serial_read_thread()
 
         if( new_packet == 1 )
         {
+          // Call serial callback. If it returns 0, flush the buffer
+          if ( !serial_read_callback( final_buffer, (bytes_read-1) ) )
+          {
+            // Wait ~250ms
+            usleep(250000);
+            // Flush the port
+            while( PollComport( serial_port_number, serial_buffer, 
+                                                                BUFFER_SIZE ) );
 
-          serial_read_callback( final_buffer, (bytes_read-1) );
+            memset( serial_buffer, 0x00, sizeof(serial_buffer) );
+          }
 
           new_packet = 0;
           memset( packet_buffer, 0x00, sizeof(packet_buffer) );
@@ -247,11 +256,12 @@ void send_serial_message( uint8_t* packet_buffer, int16_t buffer_size )
 }
 
 /*******************************************************************************
- * @fn    void dummy_callback( uint8_t* buffer, uint32_t size )
- * @brief  Dummy function so the callback function isn't initially NULL
+ * @fn    uint8_t dummy_callback( uint8_t* buffer, uint32_t size )
[…]

### 2011-09-13 host/lib/serial.h

@@ -12,7 +12,7 @@
 #define BUFFER_SIZE ( 512 )
 
 // Function Prototypes
-uint8_t serial_open( int32_t, int32_t, void (*)( uint8_t*, uint32_t) );
+uint8_t serial_open( int32_t, int32_t, uint8_t (*)( uint8_t*, uint32_t) );
 void *serial_read_thread();
 void serial_close();
 

### 2011-09-13 host/threadtest/main.c

@@ -19,6 +19,8 @@
 #warning MAX_DEVICES not defined, defaulting to 3
 #endif
 
+#define MAX_ROUNDS (3000)
+
 void send_serial_message( uint8_t* packet_buffer, int16_t buffer_size );
 void *graph_thread();
 
@@ -40,14 +42,16 @@ static volatile uint8_t *power_table = &rp_tables[MAX_DEVICES];
 int main( int argc, char *argv[] )
 {
   int32_t rc;
+  int32_t timeout;
 
   // Handle interrupt events to make sure files are closed before exiting
   (void) signal( SIGINT, sigint_handler );
-
+  
   // Make sure input is correct
-  if( argc < 3 )
+  if( argc < 6 )
   {
-    printf("Usage: %s port baudrate [graph (0,1)]\n", argv[0]);
+    printf("Usage: %s port baudrate C(0.0-1.0) [graph (0,1)] timeout\n", 
+                                                                      argv[0]);
     return 0;
   }
 
@@ -60,11 +64,17 @@ int main( int argc, char *argv[] )
     exit(-1);
   }
 
+  timeout = atoi( argv[5] );
+  printf("Data collection will start in ~%d seconds...\n", timeout);
+
+  sleep(timeout);
+  system("banshee --play");
+
   // Initialize routing an power tables
   memset( (uint8_t*)power_table, 0xff, MAX_DEVICES );
   memset( (uint8_t*)routing_table, ( MAX_DEVICES + 1 ), MAX_DEVICES );
 
-  if ( routing_initialize() )
+  if ( routing_initialize( (energy_t)strtod( argv[3], NULL ) ) )
   {
     printf("Error initializing routes.\n");
     exit(-1);
@@ -100,10 +110,12 @@ int main( int argc, char *argv[] )
     exit(-1);
   }  
 
+  uint32_t round = 0;
+
   for(;;)
   {
     //uint8_t index;
-
+    
     // Wait until routing is done
     pthread_mutex_lock ( &mutex_route_done );
 
@@ -111,7 +123,7 @@ int main( int argc, char *argv[] )
     send_serial_message( (uint8_t *)rp_tables, sizeof(rp_tables) );
 
     // Only graph when asked to
-    if ( argv[3][0] == '1')
+    if ( argv[4][0] == '1')
     {
       // Start generating the graph
       pthread_mutex_unlock ( &mutex_graph );
@@ -131,6 +143,16 @@ int main( int argc, char *argv[] )
       printf( "%02X ", power_table[index] );
     }
[…]

### 2011-09-13 docs/thesis/paper.tex
Updated paper.tex and added thesis.tex

@@
 \documentclass{article}
+\usepackage{graphicx}
 \begin{document}
@@
 The primary downside to this method is that nodes closer to the root will deplete their energy source faster due to the need to relay messages from children nodes.
-Other protocols focus on the reliability of the network$\ll$citation needed$\gg$.
+Other protocols focus on the reliability of the network~\cite{routing:storeandforward}.
 Dynamic networks are a problem with WBANs.
@@
 
+\begin{figure}[htb]
+\begin{center}
+\includegraphics[width=0.5\textwidth]{figures/body.pdf}
+\end{center}
+\caption{Experimental Setup}
+\label{fig:body}
+\end{figure}
+
 \section{Proposed Cost Function}
@@
 \label{eqn:accumulated_energy}
-E^j_{acc_{(i)}} = E^j_{acc_{(i-1)}} + \frac{RSSI^T}{\alpha}
+E^j_{(i)} = E^j_{(i-1)} + \frac{RSSI^T}{\alpha}
 \end{equation}
@@
 \label{eqn:link_cost}
-C^j_{l,k} = \left[ \left(\frac{ E^l_{acc_{(i-1)}} + \frac{RSSI^T}{\alpha_{l,k}}}{i}\right) - \frac{\sum_jE^l_{acc_{(i-1)}}}{i-1} \right]
+C^j_{l,k} = \left[ \left(\frac{ E^l_{(i-1)} + \frac{RSSI^T}{\alpha_{l,k}}}{i}\right) - \frac{\sum_jE^l_{(i-1)}}{i-1} \right]
+\end{equation}
+
+\begin{equation}
+\label{eqn:link_cost}
+C^j_{l,k} = \frac{RSSI^T}{\alpha_{l,k}} \times \left( \frac{1+\left(\frac{E^j_i}{E^{min}_{i}} \right)^c}{2} \right)
 \end{equation}
@@
 \section{Hardware Implementation}
+\subsection{Hardware Platform}
 The hardware platform used for this test consists of the Texas Instruments (TI) EZ430-RF2500.
@@
 
+The host is a laptop computer with an Intel(R) Core(TM) i5-2410M CPU and 4GB of RAM.
+Eight EDs were placed on a 170cm, 70kg male subject as shown in Figure~\ref{fig:body}.
+The host was carried on a backpack and connected to the AP via a USB cable.
+The data was sampled, and the routing algorithm was run, at a rate of 5Hz.
+This sampling rate is appropriate as long as the subject, or its environment, does not move faster than this.$\ll$What was the name for this?$\gg$ If a data needs to be sampled at a faster rate, some hardware and firmware changes might be necessary, but the host software will still work.
+
+\subsection{Cycle of Operations}
 A normal cycle of operations with $n$ devices works as follows:
@@
 
+\subsection{Details}
 To minimize the number of control packets from being transmitted, the EDs are not individually polled.
 The only control packet that is sent is the synchronization beacon, which also carries the routing and power tables.
-Once the EDs are synchronized, they transmit on a pre-defined schedule to avoid collisions.
+Once the EDs are synchronized, they transmit their data on a pre-defined schedule to avoid collisions.
 Each ED has a network ID ranging from 1 to $n$.
@@
 
+\begin{figure}[htb]
+\begin{center}
+\includegraphics[width=0.5\textwidth]{figures/walk2-c100.pdf}
+\end{center}
+\caption{Walking (C=100)}
+\label{fig:walk2-c100}
+\end{figure}
+
+\begin{figure}[htb]
+\begin{center}
+\includegraphics[width=0.5\textwidth]{figures/walk2-c0.pdf}
+\end{center}
+\caption{Walking (C=0)}
+\label{fig:walk2-c0}
+\end{figure}
+
 $\ll$describe figure  3 and interpret results – network lifetime (like fig.8 in paper with Jeff)$\gg$

### 2011-09-13 docs/thesis/thesis.tex
Updated paper.tex and added thesis.tex

\documentclass{article}
\usepackage{graphicx}
\begin{document}
\title{Masters Thesis}
\author{Alvaro Prieto}
\maketitle

\pagebreak

\tableofcontents

\pagebreak

\section{Introduction}

% Include literature survey from other file
\input{sources/literature_survey} %uncomment to include literature survey

\section{Software Libraries}
One of the primary contributions of this project consists of software libraries for working with WBANs. Due to the project requirements, a choice was made to use free and open source tools, as opposed to proprietary software. Because of this choice, most of the radio libraries had to be either ported or re-written from scratch. There were two main hardware devices that were targeted in these libraries. The first consisted of a CC430 device, which has a built in 915MHz radio. The second was an MSP430 with an external CC2500 2.4GHz radio.

\subsection{Project History}
The first software libraries were initially developed by Corey Provencher, a student at RIT, for a senior design project. He developed a hardware board with a variety of sensors to measure body movements and the corresponding software to use it. All of his software development was done using the Texas Instruments (TI) software tools on Windows(TM).

Jeffery Robble, another student, took over and ported most of the code to use Linux based open source tools. The code ported over by Robble used TI's radio stack \emph{SimpliciTI}. Unfortunately, SimpliciTI was too complex and resource heavy to do time-sensitive operations, like clock synchronization, over the radio. It also required the use of other TI software libraries in order to work.

After Robble left, the work continued. The main goal of the project at the time was to develop wireless electrocardiogram(ECG)(Section~\ref{section:wirelessecg}). While planning for the wireless ECG, a decision was made to develop an entire WBAN platform that could later be used for other projects. The new platform would be developed in-house and included both hardware and software. Because of this, the entire library was to be written from the ground up. This would prevent any rights issues from coming up if published.

\subsection{Design Choices}
The decision to write all software libraries from scratch was not an easy one. Initially the SimpliciTI libraries were modified to try and meet the project needs, but this proved to be a complex task. Due to the simple nature of the planned WBAN, a much less complex approach could be used. The main application only required streaming data from various wireless sensors.$\ll$ Should I elaborate on the design constraints here? The number of devices is known, the time of transmission is known etc...$\gg$ This could be accomplished with a very minimal radio library.

\subsection{CC430 Device}
The first device used in this project was Provencher's initial hardware platform. Due to the limited quantities of that device, a new version of the hardware platform was developed. The new device removed all unused sensors from the board to minimize its footprint. It included a CC430 device with a 32.768Khz crystal oscillator for precise timekeeping. The analog-to-digital(A/D) ports were used to attach any needed sensors.

Libraries were written for serial communication, visual indicators, analog measurements, time synchronization and radio communications. The serial communication was used to transmit data to the host computer, while the rest were used in the wireless ECG application.

\subsection{MSP430-RF2500 Device}
The second major hardware platform was the MSP430-RF2500. This device is manufactured by TI and sold as a development platform. The MSP430-RF2500 device consists of an MSP430 microcontroller and a CC2500 radio. There were several reasons why this device was used. The primary use for the MSP430-RF2500 was to develop the libraries to use the CC2500 radio. The next planned hardware platform for the project consisted of a CC430 with a CC2500. The previous software libraries would need to be expanded to include a second radio for communication.

Unlike the CC430, which has a radio core built in, the CC2500 is a standalone unit. In order to expedite the software development process, a working platform was needed. Software development new, untested, hardware is not a feasible option. Debugging both hardware and software simultaneously is not a good idea.

The previously developed CC430 libraries were ported to work with the MSP430 microcontroller and expanded to use the CC2500 radio. Features of the new radio libraries include power control, addressing and channel switching.

\section{Specialized Cost Function for Maximizing Body Area Networks Lifetime Using Global Routing Algorithms}
$\ll$ Insert most of the paper here. $\gg$

\section{Wireless ECG Application}\label{section:wirelessecg}
\subsection{Description}
The main goal of the Wireless ECG application is to produce the equivalent output of a traditional, wired, 12-lead ECG using three wireless sensors.

The three main parts of the project are: 
\begin{itemize}
\item Developing the analog front-end to measure the biopotential signals from the subject.
\item Developing the software/firmware libraries to measure, store, and transmit the data from the EDs to the host.
\item Developing the lead reconstruction algorithms to convert the 3 measurements back to a physician friendly 12-lead display.
\end{itemize}
The work presented here focuses on the software/firmware development for the application.

\subsection{TinyOS Platform}
The first wireless ECG platform was developed using the Imperial College London BSN Motes running TinyOS. These devices

\subsection{CC430 Platform}

\subsection{Time Synchronization}
The method used compensates for time offsets between samples and minimizes the effect of clock drift by periodically synchronizing independent clock sources together. Figure~\ref{fig:network_layout} shows an example system consisting of three end devices (\emph{ED}s), one access point (\emph{AP}) and one host computer. 
\begin{figure}[htb]
\begin{center}
\includegraphics[width=0.4\textwidth]{figures/network-map-test2.pdf}
\end{center}
\caption{Test Network Layout}
\label{fig:network_layout}
\end{figure}
All devices have a timer used for time-stamping and synchronization. All events, including samples and data transfers, occur according to the value of the timer. Due to certain hardware limitations, the value of the timer cannot be altered, it can only be reset to zero. This does not, in any way, affect the synchronization of devices.

The timer inside the \emph{AP} can be considered the primary, or global, timer. The goal is to synchronize all \emph{ED}, or local, timers with the global timer. Whenever the global timer reaches its maximum value, and rolls over back to zero, the \emph{AP} sends a broadcast synchronization beacon instructing each end device it is time zero. When an \emph{ED} receives this message, it compares its current timer value to zero, to calculate the clock drift, and resets it to zero. This ensures that all \emph{ED} timers are aligned, which then allows measurements to occur at close to the same time.

While many, more precise, time synchronization methods are available~\cite{synchronization:FTSP}, it was determined that they were not needed. Due to the low sampling rate and the use of crystal oscillators to run the timers, the precision obtained with this simple method was good enough for the system. More complex methods would not have significantly improved results.

\subsection{Scheduling}
A round robin-like scheme is used to poll devices in the network. It focuses on simplicity and portability.

In previous implementations, each \emph{ED} was polled individually for sampled data, which required all \emph{ED}s to be on at all times. Once all device clocks are synchronized, polling is no longer necessary. Instead of sending poll messages to each device, the \emph{AP} just listens for \emph{ED} transmissions. The \emph{ED}s will wake up when they are scheduled to transmit and send the required data automatically. This method greatly reduces the power consumption of both \emph{AP} and \emph{ED}s. After initialization, the \emph{AP} only needs to send a periodic time synchronization beacon and the collected data back to the host computer. Once synchronized, end devices will only need to turn their receiver on around the time the synchronization beacon is expected.

Figure~\ref{fig:system_schedule} shows a sample system schedule. The figure shows the time each device is on, and the state of each radio. The ``sleep'' time for each \emph{ED} does not show when a device wakes up to sample data. In this specific case, the \emph{AP} has one extra time-slot allotted for a fourth \emph{ED}.

\begin{figure}[htb]
\begin{center}
\includegraphics[width=1\textwidth]{figures/sync_schedule.pdf}
\end{center}
\caption{System Schedule}
\label{fig:system_schedule}
\end{figure}

Certain devices can further reduce power consumption by sampling without using the main processor. This works by using direct-memory access(\emph{DMA}) which allows the analog-to-digital converter(\emph{ADC}) to store measurements in memory without first going through the processor. Using this method, Figure~\ref{fig:system_schedule} accurately presents the time each \emph{ED} is powered on. Sampling with DMA has not been implemented in the current wireless ECG platform.

\subsection{Host Interface}

\section{Conclusion}

\bibliography{sources/thesis.bib}{}
\bibliographystyle{plain}
\end{document}

### 2011-09-13 sim/readcsv/main.c
Modified readcsv to also read the tx powers for more accurate channel data.

@@ -22,7 +22,8 @@
 #warning MAX_DEVICES not defined, defaulting to 3
 #endif
 
-void read_line( char*, energy_t* );
+void read_energy_line( char*, energy_t* );
+uint8_t read_power_line ( FILE* , energy_t* );
 uint8_t read_table( FILE* , energy_t p_rssi_table[][MAX_DEVICES+1] );
 void *graph_thread();
 void sigint_handler( int32_t sig );
@@ -43,26 +44,40 @@ static volatile uint8_t *power_table = &rp_tables[MAX_DEVICES];
 
 int32_t main ( int32_t argc, char *argv[] )
 {
-  FILE *fp_in;
+  FILE *fp_rssi;
+  FILE *fp_powers;
   int32_t rc;
+  uint8_t node_index;
   energy_t rssi_table[MAX_DEVICES+1][MAX_DEVICES+1];
-
+  energy_t previous_powers[MAX_DEVICES];
+  uint32_t sample_limit = 10000;
+  
   // Handle interrupt events to make sure files are closed before exiting
   (void) signal( SIGINT, sigint_handler );
 
   // Make sure the filename is included
-  if ( argc < 3 )
+  if ( argc < 4 )
   {
-    printf( "Usage: %s infile [graph (0,1)]\r\n", argv[0] );
+    printf( "Usage: %s rssi.csv powers.csv C(0.0-1.0) [graph (0,1)]\r\n"
+                                                                    , argv[0] );
     return 1;
   }
+  
+  // Open input rssi csv file
+  fp_rssi = fopen( argv[1], "r" );
 
-  // Open input csv file
-  fp_in = fopen( argv[1], "r" );
-
-  if( NULL == fp_in )
+  if( NULL == fp_rssi )
   {
-    printf( "Error opening input file.\r\n" );
+    printf( "Error opening rssi input file.\r\n" );
+    return 1;
+  }
+  
+  // Open input rssi csv file
+  fp_powers = fopen( argv[2], "r" );
+
+  if( NULL == fp_powers )
+  {
+    printf( "Error opening tx powers input file.\r\n" );
     return 1;
   }
   
@@ -70,7 +85,16 @@ int32_t main ( int32_t argc, char *argv[] )
   memset( (uint8_t*)power_table, 0xff, MAX_DEVICES );
   memset( (uint8_t*)routing_table, ( MAX_DEVICES + 1 ), MAX_DEVICES );
   
-  if ( routing_initialize() )
[…]

### 2011-09-13 results/scripts/plotcsv.py
Updated plotcsv.py to fix replotting in svg file.

@@ -41,16 +41,16 @@ for line in fileinput.input():
       print "Processing ", line
       scriptfile = open( scriptfilename,'w')
       
-      scriptfile.write("# gnuplot script for '" + datafilename + "'\n")
-      scriptfile.write("set terminal svg size 640,480 dynamic enhanced fname 'arial'\n")
-      scriptfile.write("set output 'plots/svg/" + dirs[len(dirs)-2] + ".svg'\n")
+      scriptfile.write("# gnuplot script for '" + datafilename + "'\n")  
       scriptfile.write("set title '" + dirs[len(dirs)-2] +  "'\n")
       scriptfile.write("set xlabel 'Time (s)'\n")
       scriptfile.write("set ylabel 'Power Used (mW)'\n")
       scriptfile.write("plot '" + datafilename + "' using 1:2 with lines title 'Average'\n")
       for ed in range(1,max_devices+1):
         scriptfile.write("replot '" + datafilename + "' using 1:" + str(ed + 2) + " with lines title 'ED" + str(ed) + "'\n")
-      
+      scriptfile.write("set terminal svg size 640,480\n")
+      scriptfile.write("set output 'plots/svg/" + dirs[len(dirs)-2] + ".svg'\n")
+      scriptfile.write("replot\n")
       scriptfile.write("set terminal png size 640,480\n")
       scriptfile.write("set output 'plots/png/" + dirs[len(dirs)-2] + ".png'\n")
       scriptfile.write("replot\n")

### 2011-09-13 host/threadtest/main.h
Updated process_packet function.

@@ -12,7 +12,7 @@
 #define BUFFER_SIZE ( 512 )
 
 // Function Prototypes
-void process_packet( uint8_t*, uint32_t );
+uint8_t process_packet( uint8_t*, uint32_t );
 
 void sigint_handler( int32_t sig );
 

### 2011-09-13 host/rssistream/main.h
Commit before merge.

@@ -18,7 +18,7 @@
 #define SYNC_BYTE   ( 0x7E )
 #define ESCAPE_BYTE ( 0x7D )
 
-#define MAX_DEVICES (6)
+#define MAX_DEVICES (3)
 
 #define BROADCAST_ADDRESS (0x00)
 

### 2011-09-13 results/scripts/csv2dat.py

@@ -8,7 +8,7 @@
 import csv
 import fileinput
 for line in fileinput.input():
-    if 'energies.csv' in line:
+    if '.csv' in line:
       line = line.replace('\n','')
       print "Processing ", line
       #open output file

### 2011-09-14 docs/thesis/sources/literature_survey.tex
Added figures and updated thesis.tex

@@
-\section{Literature Survey}\label{section:litsurvey}
+\section{Previous Work}\label{section:litsurvey}
 \subsection{Time Synchronization}
@@
 
-\subsection{Similar Work}
-In order to determine the improvements of novel methods, reference systems need to be available to compare with.
+%\subsection{Similar Work}
+%In order to determine the improvements of novel methods, reference systems need to be available to compare with.
 Both~\cite{mac:tdma}~and~\cite{mac:lowdutycycle} provide a suitable reference system for a synchronized, non-relaying BSN.

### 2011-09-14 docs/thesis/thesis.tex
Added figures and updated thesis.tex

@@
 \title{Masters Thesis}
-\author{Alvaro Prieto}
+\author{Alvaro Prieto\\
+BS/MS Electrical Engineering\\
+\texttt{axp8664@rit.edu}}
 \maketitle
@@
 
-\section{Introduction}
+\begin{abstract}
+...
+\end{abstract}
 
+\section{Summary of Contributions}
+\begin{itemize}
+\item BSN Software Libraries
+\item wireless ECG platform
+\item New routing algorithm
+\end{itemize}
 % Include literature survey from other file
@@
 
-The three main parts of the project are: 
+The three main parts of the wireless ECG project are: 
 \begin{itemize}
@@
 \subsection{TinyOS Platform}
-The first wireless ECG platform was developed using the Imperial College London BSN Motes running TinyOS. These devices
+The first wireless ECG platform was developed using the Imperial College London BSN Motes, which use the TinyOS operating system.
+These devices are programmed in a derivative of the C language called NesC.
+While working with these devices and operating system was not particularly simple, the hardware worked, and had been used in other projects.
+A full 3-sensor wireless ECG application (minus the lead reconstruction) was implemented with these devices.
+Since the devices run an operating system, several, non-ideal, modifications were required to obtain precise time synchronization.
 
 \subsection{CC430 Platform}
+After demonstrating the system using the TinyOS devices, the development of the in-house system continued.
+After several hardware implementations, a similar 3-sensor wireless ECG application was developed.
+In contrast to the TinyOS system, the CC430 platform only used hardware and software that was developed by the team.
 
@@
 More complex methods would not have significantly improved results.
+Figure~\ref{fig:sync_data} shows the result of three separate wireless devices measuring the same signal.
+This method was used to confirm that the samples were aligned.
+
+\begin{figure}[htb]
+\begin{center}
+\includegraphics[width=0.8\textwidth]{figures/syncdata.pdf}
+\end{center}
+\caption{Synchronization Test}
+\label{fig:sync_data}
+\end{figure}
 
@@
 \subsection{Host Interface}
+The primary focus of this project was to obtain synchronized samples from several wireless devices.
+A secondary, but important, goal was to display the data in real time.
+To speed up development, the live-plotting tool kst(http://kst-plot.kde.org/) was used.
+A application, written in C, communicated via a USB-to-Serial link with the AP and received all of the ED data.
+It then output the data to CSV files, which were then used as an input for kst.
+Figure~\ref{fig:heartbeats} shows an ECG lead wirelessly measured from a human subject.
+
+\begin{figure}[htb]
+\begin{center}
+\includegraphics[width=1\textwidth]{figures/heartbeats.pdf}
+\end{center}
+\caption{Sample Measurement from Single ECG Lead}
+\label{fig:heartbeats}
+\end{figure}
 
 \section{Conclusion}
+...
+
+\pagebreak
+
+\listoffigures
+
+%\listoftables
 

### 2011-09-14 sim/lib/dijkstra.c
I AM AN IDIOT.

@@ -364,7 +364,7 @@ uint8_t dijkstra( uint8_t source_id, energy_t c_factor )
       
           // Calculate the current cost of the link
           current_cost = 1 + 
-            pow( ( p_source_node->energy / current_minimum ), c_factor ) ;
+            pow( ( p_destination_node->energy / current_minimum ), c_factor ) ;
           
           // Normalize
           current_cost /= ( 2 );          

### 2011-09-14 host/lib/routing.c
Restored power control compensation to readcsv

@@ -280,7 +280,7 @@ uint8_t parse_table_d ( energy_t p_rssi_table[][MAX_DEVICES+1],
         tx_power = dbm_to_watt( p_previous_powers[col_index + 1] );
       }
 
-      tx_power = dbm_to_watt( 1.5l ); // uncomment for no power control
+      //tx_power = dbm_to_watt( 1.5l ); // uncomment for no power control
 
       // Compute the minimum power required to meet this link with 'target_rssi'
       // alpha is the channel attenuation, that is received/transmitted power

### 2011-09-14 docs/thesis/paper.tex
Updated paper with pseudocode and new figures

@@
 \usepackage{graphicx}
+\usepackage{algorithm}
+\usepackage{algorithmic}
 \begin{document}
@@
 One solution for this problem was proposed by Ehyaie et al.~\cite{relay:networklife}, it consists of using dedicated, non-sensing, relay nodes with larger power sources.
-While this method increases the network lifetime$\ll$define term$\gg$, it requires more hardware.
+While this method increases the network lifetime, it requires more hardware.
 Finally, Nabi et al.
@@
 Nodes keep track of neighbors and use power control to use the smallest transmit power while maintaining a specific link quality.
+
+For the purpose of this paper, \textbf{network lifetime} is defined as the time it takes a single node to deplete its energy source.
 
@@
 \end{center}
-\caption{Experimental Setup}
+\caption{Experimental Setup (Not all links shown)}
 \label{fig:body}
@@
 
-Each device's energy use is accumulated as shown in Equation~\ref{eqn:accumulated_energy}, where $j$ denotes the device ID, $i$ is the current polling round, $\alpha$ is the channel attenuation, and $RSSI^T$ is the target RSSI.
+\subsection{Device Energy}
+Each device's energy use is accumulated as shown in Equation~\ref{eqn:accumulated_energy}, where $j$ denotes the device ID, $i$ is the current polling round, $\alpha_{j,k}$ is the channel attenuation for the selected link, and $RSSI^T$ is the target RSSI.
 The current energy used is incremented by the energy used to transmit a single packet while maintaining $RSSI^T$.
-The channel attenuation for the selected link,$\alpha$, can be seen in Equation~\ref{eqn:alpha}, where RSSI is the received power and $P_tx$ is the transmitted power.
-The actual link cost is computed by calculating the energy used by a device \emph{if} that link is selected and subtracting the average energy used across all devices.
-Equation~\ref{eqn:link_cost} is normalized by the current polling round $i$.
+The channel attenuation for the selected link,$\alpha_{j,k}$, can be seen in Equation~\ref{eqn:alpha}, where RSSI is the received power measured by device $k$ and $P_{tx}$ is the transmitted power used by device $j$.
+
+\subsection{Link Cost}
+The actual link cost is computed by calculating the energy used by the device, \emph{if} that link is selected, and multiplying by a cost factor(Equation~\ref{eqn:link_cost}).
+The cost factor calculation takes the energy used by the destination device, $E^k_i$, and divides it by the current energy minimum, $E^{min}_{i}$, which is the energy used by the device with the minimum energy used by round $i$.
+This ratio is then raised to the power $M$, which determines how strong the effect will be $\ll$Need a better way of saying this..$\gg$.
+If a device's energy is much greater than the current minimum, it will be avoided as a relay.
+The $M$ factor in Equation~\ref{eqn:link_cost} can be any number in $0 < M < \infty$.
+The cost factor is normalized so when $M=0$, it goes back to the traditional cost of required power to meet the link.
+
+Once all link costs are calculated, Dijkstra's algorithm~\cite{dijkstra:algorithm} is used to compute the shortest path.
+
 \begin{equation}
 \label{eqn:alpha}
-\alpha = \frac{RSSI}{P_{tx}}
+\alpha_{j,k} = \frac{RSSI}{P_{tx}}
 \end{equation}
@@
 \label{eqn:link_cost}
-C^j_{l,k} = \left[ \left(\frac{ E^l_{(i-1)} + \frac{RSSI^T}{\alpha_{l,k}}}{i}\right) - \frac{\sum_jE^l_{(i-1)}}{i-1} \right]
+C^j_{j,k} = \frac{RSSI^T}{\alpha_{j,k}} \times \left( \frac{1+\left(\frac{E^k_i}{E^{min}_{i}} \right)^M}{2} \right)
 \end{equation}
 
-\begin{equation}
-\label{eqn:link_cost}
-C^j_{l,k} = \frac{RSSI^T}{\alpha_{l,k}} \times \left( \frac{1+\left(\frac{E^j_i}{E^{min}_{i}} \right)^c}{2} \right)
-\end{equation}
-
-Once all link costs are calculated, Dijkstra's algorithm~\cite{dijkstra:algorithm} is used to compute the shortest path.
-
+\subsection{Assumptions}
 This algorithm works under several assumptions.
@@
 The second is that each wireless sensor, or End Device(ED), can act as a relay for messages from other nodes.
-The Third and last requirement is that the link quality,or channel attenuation, between all EDs and AP is known.
+The Third is that the link quality,or channel attenuation, between all EDs and AP is known.
+Finally, all EDs begin with an equal energy used, that is, they all have the same energy stored.
 All routing calculations take place in the AP.
+Figure~\ref{fig:body} shows a sample network with EDs labeled 1 through 8, and one AP.
 
-$\ll$describe network (centralized network with powerful AP and multiple sensors able to act as relays, power control), figure 1 (like in paper fig.1 + symbols for channel attenuation, cost, sensor ID,…) and describe symbols in figure.$\gg$
-
-$\ll$Define accumulated energy (normalized power), cost function and all relevant symbols$\gg$
+\subsection{Performance Metric}
+The goal of the new cost function is to increase network lifetime.
 
@@
 \section{Performance Analysis}
+In what follows, we analyze the performance of the proposed cost function. 
 $\ll$”in what follows we ….
 (state simulation and hardware), implementation of cost function in Dijktra (ref),step wide pseudo code for algorithm, comparison approach to “normal” cost function with Dijkstra$\gg$
+
+\subsection{Algorithm Implementation}
+\begin{algorithm}
+\caption{Dijkstra's Algorithm with new cost function}
+\label{alg:dijkstra}
+\begin{algorithmic}
+\FOR{$i = 1$ to Total Nodes} 
+	\STATE $node_i.visited \leftarrow 0$
+	\STATE $node_i.distance \leftarrow \infty$
+\ENDFOR
+\STATE $node_{AP}.distance \leftarrow 0$
+\STATE $energy_{min} = min(node.energy)$
+\WHILE{unvisited nodes available}
+	\STATE $node_{source} \leftarrow $ node with smallest distance
+	\FOR{$i = 1$ to Total Links}
+		\IF{$link_i.source = node_{source}$}
+			\STATE $node_{dest} = link_i.destination$
+			\IF{$node_{dest}.visited = 0$}
+				\STATE $cost \leftarrow link_i.power * (1 + node_{dest}.energy/energy_{min} )^M/2$
+				\STATE $newdistance \leftarrow node_{source}.distance + cost$
+				\IF{$distance < node_{dest}.distance$} 
+				\STATE$node_{dest}.distance \leftarrow newdistance$
+				\ENDIF
+			\ENDIF
+		\ENDIF
+	\ENDFOR
+	\STATE $node_{source}.visited \leftarrow 1$
+\ENDWHILE
+\end{algorithmic}
+\end{algorithm}
 
@@
 This sampling rate is appropriate as long as the subject, or its environment, does not move faster than this.$\ll$What was the name for this?$\gg$ If a data needs to be sampled at a faster rate, some hardware and firmware changes might be necessary, but the host software will still work.
+
+Two experimental setups were used.
+For the first, the subject remained relatively static in a sitting position.
+The second consisted of a more dynamic environment, where the subject walked around a room.
 
@@
 
+\begin{figure}[htb]
+\label{tab:rssi}
+\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}
+\hline  & AP & ED1 & ED2 & ED3 & ED4 & ED5 & ED6 & ED7 & ED8 \\ 
+\hline AP & - & -60 & -60 & -60 & -60.5 & -60.5 & -60.5 & -60 & -79.5 \\ 
+\hline ED1 & -66 & - & -62 & -75 & -63.5 & -69 & -75.5 & -73 & -67.5 \\ 
+\hline ED2 & -72 & -61.5 & - & -80 & -69 & -70 & -74 & -75 & -66 \\ 
+\hline ED3 & -77.5 & -74.5 & -80.5 & - & -67.5 & -70 & -72.5 & -70.5 & -70 \\ 
+\hline ED4 & -39 & -63 & -69.5 & -67 & - & -82.5 & -74.5 & -69 & -65.5 \\ 
+\hline ED5 & -67.5 & -68 & -71 & -69.5 & -84 & - & -55.5 & -51.5 & -62 \\ 
+\hline ED6 & -67 & -75 & -75 & -72.5 & -75 & -55.5 & - & -44.5 & -53 \\ 
+\hline ED7 & -46 & -70 & -77.5 & -70.5 & -69 & -50.5 & -45 & - & -51.5 \\ 
+\hline ED8 & -78 & -67.5 & -66.5 & -71 & -66 & -62 & -54 & -51.5 & - \\ 
+\hline 
+\end{tabular} 
+\caption{Sample RSSI Table}
+\end{figure}
+
 \section{Results}
+
+Figure~\ref{fig:sit1plots} shows the resulting accumulated energies while the subject was in a sitting position. The experiment was run with both an $M=0$ as a reference, and $M=100$.
 
@@
 \begin{center}
-\includegraphics[width=0.5\textwidth]{figures/walk2-c100.pdf}
+\includegraphics[width=1.0\textwidth]{figures/sit1plots.pdf}
 \end{center}
-\caption{Walking (C=100)}
-\label{fig:walk2-c100}
+\caption{Sitting experiment with M=0 and M=100}
+\label{fig:sit1plots}
 \end{figure}
@@
 \begin{center}
-\includegraphics[width=0.5\textwidth]{figures/walk2-c0.pdf}
+\includegraphics[width=1.0\textwidth]{figures/walk2plots.pdf}
 \end{center}
-\caption{Walking (C=0)}
-\label{fig:walk2-c0}
+\caption{Walking experiment with M=0 and M=100}
+\label{fig:walk2plots}
 \end{figure}

### 2011-09-14 docs/thesis/paper.tex
Updated paper.tex and figures.

@@
 \begin{document}
-\title{Specialized Cost Function for Maximizing Body Area Networks Lifetime Using Global Routing Algorithms}
+\title{Specialized Cost Function for Maximizing Body Area Network Lifetime Using Global Routing Algorithms}
 \author{Alvaro Prieto}
@@
 
-For the purpose of this paper, \textbf{network lifetime} is defined as the time it takes a single node to deplete its energy source.
+For the purpose of this paper, \textbf{network lifetime} is defined as the time it takes any single node to deplete its energy source.
 
-In this contribution we propose a novel routing algorithm that focuses not on single devices, but the overall network.
-Current methods focus on maximizing device lifetime.
-This is generally a good goal, but certain systems require all devices working to function properly.
+In this contribution we propose a novel routing algorithm that focuses not on single devices, but the network as a whole.
+Current methods focus on maximizing individual device lifetime or minimizing the energy-per-bit.
+While those are generally good goals, certain systems require all devices functioning to be considered operational.
 Keeping that in mind, the proposed system maximizes the network lifetime by normalizing power use across the network.
+
 $\ll$”In this contribution we ….
@@
 
-\begin{figure}[htb]
+\begin{figure}[!htb]
 \begin{center}
@@
 \section{Proposed Cost Function}
-In the proposed system, we define a new link cost function that is then used in the routing algorithm.
+In the proposed system, we define a new link cost function that is then used in a routing algorithm.
 Traditionally, the power required to make a link possible is used as the link cost.
@@
 \subsection{Device Energy}
-Each device's energy use is accumulated as shown in Equation~\ref{eqn:accumulated_energy}, where $j$ denotes the device ID, $i$ is the current polling round, $\alpha_{j,k}$ is the channel attenuation for the selected link, and $RSSI^T$ is the target RSSI.
+Each device's normalized energy use is calculated as shown in Equation~\ref{eqn:accumulated_energy}, where $j$ denotes the device ID, $i$ is the current polling round, $\alpha_{j,k}$ is the channel attenuation for the selected link, and $RSSI^T$ is the target RSSI.
 The current energy used is incremented by the energy used to transmit a single packet while maintaining $RSSI^T$.
@@
 The $M$ factor in Equation~\ref{eqn:link_cost} can be any number in $0 < M < \infty$.
-The cost factor is normalized so when $M=0$, it goes back to the traditional cost of required power to meet the link.
+The cost factor is normalized so when $M=0$, it returns to the traditional cost of required power to meet the link.
 
-Once all link costs are calculated, Dijkstra's algorithm~\cite{dijkstra:algorithm} is used to compute the shortest path.
+Since the link costs are directly related to the destination device, they are calculated during the routing algorithm computation.
+For this paper, Dijkstra's algorithm~\cite{dijkstra:algorithm} is used to select the best route.
 
@@
 This algorithm works under several assumptions.
-The first is that this is a centralized network with one central Access Point (AP) with virtually unlimited energy and  computing power.
+The first being that this is a centralized network with one Access Point (AP) with virtually unlimited energy and computing power.
 The second is that each wireless sensor, or End Device(ED), can act as a relay for messages from other nodes.
-The Third is that the link quality,or channel attenuation, between all EDs and AP is known.
+The Third is that the link quality, or channel attenuation, between all EDs and AP is known.
 Finally, all EDs begin with an equal energy used, that is, they all have the same energy stored.
@@
 
-\subsection{Performance Metric}
-The goal of the new cost function is to increase network lifetime.
-
-$\ll$rational for cost function$\gg$
-
 \section{Performance Analysis}
 In what follows, we analyze the performance of the proposed cost function.
+The new cost function was tested in both simulated and experimental environments.
+In order to gauge the efficiency of the cost function, it will be compared to an identical system which uses a traditional link cost function.
+The traditional link cost function consists of the required power to meet a link with the desired $RSSI^T$.
+This can be achieved by setting a value of $M=0$ in the new link cost function.
+
+Performance is determined by the time it takes any node's accumulated normalized energy to cross an arbitrary threshold.
+The threshold could be equivalent to the amount of energy stored in a device battery, but for this experiment, is purely arbitrary.
+
 $\ll$”in what follows we ….
@@
 \subsection{Algorithm Implementation}
-\begin{algorithm}
-\caption{Dijkstra's Algorithm with new cost function}
-\label{alg:dijkstra}
-\begin{algorithmic}
-\FOR{$i = 1$ to Total Nodes} 
-	\STATE $node_i.visited \leftarrow 0$
-	\STATE $node_i.distance \leftarrow \infty$
-\ENDFOR
-\STATE $node_{AP}.distance \leftarrow 0$
-\STATE $energy_{min} = min(node.energy)$
-\WHILE{unvisited nodes available}
-	\STATE $node_{source} \leftarrow $ node with smallest distance
-	\FOR{$i = 1$ to Total Links}
-		\IF{$link_i.source = node_{source}$}
-			\STATE $node_{dest} = link_i.destination$
-			\IF{$node_{dest}.visited = 0$}
-				\STATE $cost \leftarrow link_i.power * (1 + node_{dest}.energy/energy_{min} )^M/2$
-				\STATE $newdistance \leftarrow node_{source}.distance + cost$
-				\IF{$distance < node_{dest}.distance$} 
-				\STATE$node_{dest}.distance \leftarrow newdistance$
-				\ENDIF
-			\ENDIF
-		\ENDIF
-	\ENDFOR
-	\STATE $node_{source}.visited \leftarrow 1$
-\ENDWHILE
-\end{algorithmic}
-\end{algorithm}
-
-$\ll$TI platform (in detail with calibration – all data required to replicate) for RSSI and Disjkstra$\gg$
+Detailed pseudo-code for the routing algorithm can be found in Algorithm~\ref{alg:dijkstra} in Appendix~\ref{appendix:algorithms}.
+Similarly, code for energy used calculation can be found in Algorithm~\ref{alg:energyused}.
 
@@
 This device includes both an MSP430F2274 microcontroller along with a CC2500 2.4GHz transceiver.
-A single device, labeled the Access Point (AP), is connected via a USB to Serial link to the host computer.
+The CC2500 is configured to run at 250kbps.
+A single device, labeled the Access Point (AP), is connected via a USB to Serial link, running at 115200 BAUD, to the host computer.
 All other devices, labeled End Devices (EDs), are battery powered.
@@
 The data was sampled, and the routing algorithm was run, at a rate of 5Hz.
-This sampling rate is appropriate as long as the subject, or its environment, does not move faster than this.$\ll$What was the name for this?$\gg$ If a data needs to be sampled at a faster rate, some hardware and firmware changes might be necessary, but the host software will still work.
+This sampling rate is appropriate as long as the subject, or its environment, does not move faster than this.$\ll$What was the name for this?$\gg$ If data needs to be sampled at a faster rate, some hardware and firmware changes might be necessary, but the host software will still work.
 
@@
 For the first, the subject remained relatively static in a sitting position.
-The second consisted of a more dynamic environment, where the subject walked around a room.
+The second consisted of a more dynamic environment, where the subject walked around a room as seen in Figure~\ref{fig:floorplan}.
+The target RSSI ($RSSI^T$) was arbitrarily chosen to be $-60 dBm$.
+
+\begin{figure}[!htb]
+\begin{center}
+\includegraphics[width=0.6\textwidth]{figures/floorplan.pdf}
+\end{center}
+\caption{Walking environment.}
+\label{fig:floorplan}
+\end{figure}
 
@@
 \item AP sends synchronization beacon which includes routing and power control tables.
-\item Each ED transmits its own RSSI table back to the AP, while simultaneously listening to other ED messages and storing the received power.
-\item Once all EDs have transmitted their data, the AP sends a table with the RSSI data from all devices.
+\item Each ED transmits its own RSSI table back to the AP, while simultaneously listening to other ED messages and storing the received power from each.
+\item Once all EDs have transmitted their data, the AP sends a table with the RSSI data from all devices. (See Table~\ref{tab:rssi} in Appendix~\ref{appendix:tables} for an example)
 \item The host uses the RSSI table to compute the routes along with the required powers to meet the selected links.
@@
 
-\begin{figure}[htb]
+\section{Results}
+
+\begin{figure}[!ht]
+\includegraphics[width=0.8\textwidth]{figures/sit1-c0.pdf}
+\caption{Sitting M=0}
+\label{fig:sit1-c0}
+\end{figure}
+
+\begin{figure}[!ht]
+\includegraphics[width=0.8\textwidth]{figures/sit1-c100.pdf}
+\caption{Sitting M=100}
+\label{fig:sit1-c100}
+\end{figure}
+
+Figures~\ref{fig:sit1-c0}~and~\ref{fig:sit1-c100} show the resulting accumulated energies while the subject was in a sitting position.
+The experiment was run with both an $M=0$ as a reference, and $M=100$.
+Unfortunately, the results with a static subject were rather neutral.
+All EDs consume energy at different rates for both $M=0$ and $M=100$.
+
+Figures~\ref{fig:walk2-c0}~and~\ref{fig:walk2-c100}, on the other hand, show an increase in network lifetime.
+At the end of the 5 minute sampling time with $M=0$, the device which consumed the most power was ED7 with approximately 1.7 mW.
+On the other hand, with $M=100$, all EDs consume power in a uniform fashion and end up with less than 1.3 mW each.
+One important observation is that the network with $M=0$ consumed less overall power (~9.6mW) than the one with $M=100$ (~10.1mW).
+This is expected, due to the fact that the new cost function specifically avoids the most efficient path for a specific node in favor of the most efficient path for the whole network.
+While more energy is consumed overall, no single ED consumes significantly more power.
+
+To demonstrate the network lifetime improvement, an ED will be considered have depleted it's power source after consuming 1mW of normalized energy.
+In the previous example, the trial with $M=0$ reaches 1mW after approximately 175 seconds, while the case with $M=100$ reaches it around 240 seconds.
+
+\begin{figure}[!ht]
+\includegraphics[width=0.8\textwidth]{figures/walk2-c0.pdf}
+\caption{Walking M=0}
+\label{fig:walk2-c0}
+\end{figure}
+
+\begin{figure}[!ht]
+\includegraphics[width=0.8\textwidth]{figures/walk2-c100.pdf}
+\caption{Walking M=100}
+\label{fig:walk2-c100}
+\end{figure}
+
+A good example of the algorithm working was a result of hardware problems.
+During of the experiments, a single ED went off-line for several seconds and was manually restarted.
+Figure~\ref{fig:walk4-c100} shows the effect the ED failure had on the network.
+Because a device that is powered off does not consume power, it became the one with the least energy used.
+When it returned to the network, it was immediately used as a relay by other nodes.
+
+\begin{figure}[!ht]
+\includegraphics[width=0.8\textwidth]{figures/walk4-c100.pdf}
+\caption{Walking M=100}
+\label{fig:walk4-c100}
+\end{figure}
+
+$\ll$describe figure  3 and interpret results – network lifetime (like fig.8 in paper with Jeff)$\gg$
+
+$\ll$describe figure  4 and interpret results – energy per bit (like fig.9 in paper with Jeff)$\gg$
+
+\section{Conclusion}
+$\ll$at the end: different from abstract -  The abstract in past tense + concentrate on meaning of results$\gg$
+
+One improvement that can be done deals with the example from Figure~\ref{fig:walk4-c100}.
+While the failed ED was not part of the network, it was still taken into account when it came to link cost calculation.
+A better implementation would not include off-line devices in the calculation.
+
+\bibliography{sources/thesis.bib}{}
+\bibliographystyle{plain}
+
+\appendix
+\section{Algorithms}\label{appendix:algorithms}
+\begin{algorithm}[!ht]
+\caption{Dijkstra's Algorithm with new cost function}
+\label{alg:dijkstra}
+\begin{algorithmic}
+\FOR{$i = 1$ to \# of Nodes} 
+	\STATE $node_i.visited \leftarrow 0$
+	\STATE $node_i.distance \leftarrow \infty$
+	\STATE $node_i.previous \leftarrow node_i$
+\ENDFOR
+\STATE $node_{AP}.distance \leftarrow 0$
+\STATE $energy_{min} = min(node.energy)$
+\WHILE{unvisited nodes available}
+	\STATE $node_{source} \leftarrow $ node with smallest distance
+	\FOR{$i = 1$ to \# of Links}
+		\IF{$link_i.source = node_{source}$}
+			\STATE $node_{dest} \leftarrow link_i.destination$
+			\IF{$node_{dest}.visited = 0$}
+				\STATE $cost \leftarrow link_i.power * (1 + node_{dest}.energy/energy_{min} )^M/2$
+				\STATE $newdistance \leftarrow node_{source}.distance + cost$
+				\IF{$distance < node_{dest}.distance$} 
+				\STATE$node_{dest}.distance \leftarrow newdistance$
+				\STATE$node_{dest}.previous \leftarrow node_{source}$
+				\ENDIF
+			\ENDIF
+		\ENDIF
+	\ENDFOR
+	\STATE $node_{source}.visited \leftarrow 1$
+\ENDWHILE
+\end{algorithmic}
+\end{algorithm}
+
+\begin{algorithm}[!ht]
+\caption{Energy Used Calculation}
+\label{alg:energyused}
+\begin{algorithmic}
+\FOR{$i = 1$ to \# of Nodes}
+	\STATE $node_{ptr} = node_i$
+	\IF{$node_{ptr}.previous = node_{ptr}$}
+		\STATE No path to $node_i$
+	\ELSE
+		\WHILE{$node_{ptr}.previous \neq node_{ptr}$}
+			\STATE $link_{ptr} \leftarrow $ link between $node_{ptr}$ and $node_{ptr}.previous$
+			\STATE $node_{ptr}.energy += link_{ptr}.power$
+			\STATE $node_{ptr} \leftarrow = node_{ptr}.previous$
+		\ENDWHILE
+	\ENDIF
+\ENDFOR
+\end{algorithmic}
+\end{algorithm}
+
+\section{Tables}\label{appendix:tables}
+\begin{table}[htb]
 \label{tab:rssi}
@@
 \caption{Sample RSSI Table}
-\end{figure}
+\end{table}
 
-\section{Results}
-
-Figure~\ref{fig:sit1plots} shows the resulting accumulated energies while the subject was in a sitting position. The experiment was run with both an $M=0$ as a reference, and $M=100$.
-
-\begin{figure}[htb]
-\begin{center}
-\includegraphics[width=1.0\textwidth]{figures/sit1plots.pdf}
-\end{center}
-\caption{Sitting experiment with M=0 and M=100}
-\label{fig:sit1plots}
-\end{figure}
-
-\begin{figure}[htb]
-\begin{center}
-\includegraphics[width=1.0\textwidth]{figures/walk2plots.pdf}
-\end{center}
-\caption{Walking experiment with M=0 and M=100}
-\label{fig:walk2plots}
-\end{figure}
-
-$\ll$describe figure  3 and interpret results – network lifetime (like fig.8 in paper with Jeff)$\gg$
-
-$\ll$describe figure  4 and interpret results – energy per bit (like fig.9 in paper with Jeff)$\gg$
-
-$\ll$* maybe use CDFs?...$\gg$
-
-\section{Conclusion}
-$\ll$at the end: different from abstract -  The abstract in past tense + concentrate on meaning of results1
-
-\bibliography{sources/thesis.bib}{}
-\bibliographystyle{plain}
 \end{document}

### 2011-09-14 results/scripts/plotcsv.py
Updated plotcsv.py to have the correct labels.

@@ -37,16 +37,22 @@ for line in fileinput.input():
           datafile.write('\n')
       datafile.close()
       
+      title = dirs[len(dirs)-2].rsplit('c')
+      M = title[len(title)-1]
+      
+      
       # generate gnuplot script to process file
       print "Processing ", line
       scriptfile = open( scriptfilename,'w')
       
       scriptfile.write("# gnuplot script for '" + datafilename + "'\n")  
-      scriptfile.write("set title '" + dirs[len(dirs)-2] +  "'\n")
+      scriptfile.write("set grid'\n show grid\n")
+      scriptfile.write("set key left top\n")
+      scriptfile.write("set title 'Accumulated Normalized Energy vs. Time (M=" + M + ")'\n")
       scriptfile.write("set xlabel 'Time (s)'\n")
-      scriptfile.write("set ylabel 'Power Used (mW)'\n")
-      scriptfile.write("plot '" + datafilename + "' using 1:2 with lines title 'Average'\n")
-      for ed in range(1,max_devices+1):
+      scriptfile.write("set ylabel 'Accumulated Normalized Energy (mW)'\n")
+      scriptfile.write("plot '" + datafilename + "' using 1:3 with lines title 'ED1'\n")
+      for ed in range(2,max_devices+1):
         scriptfile.write("replot '" + datafilename + "' using 1:" + str(ed + 2) + " with lines title 'ED" + str(ed) + "'\n")
       scriptfile.write("set terminal svg size 640,480\n")
       scriptfile.write("set output 'plots/svg/" + dirs[len(dirs)-2] + ".svg'\n")

### 2011-09-14 docs/thesis/paper.tex
More changes to thesis.tex

@@
 \begin{abstract}
-$\ll$ at the end $\gg$
+TODO
 \end{abstract}
@@
 
-$\ll$”In this contribution we ….
-(elaborate on title), describe the algorithm (“in contradistinction to previous routing approached, we define a cost function based on the deviation of the accumulated energy of a single node from the overall average power…”), summary of work (real RSSI data + simulation and hardware implementation using TI 2.4, comparison approach, network lifetime and energy per bit…), summary of results.”$\gg$
+In order to provide accurate channel conditions, RSSI measurements of an eight ED network were used as input to our simulations(\ref{sec:simulation}).
+Real-time hardware implementations were also run to demonstrate the algorithm(\ref{sec:hardware}).
+The new cost function succeeded in improving network lifetimes when run in a dynamic environment.
+Static environments did not produce successful results.
 
@@
 The cost factor calculation takes the energy used by the destination device, $E^k_i$, and divides it by the current energy minimum, $E^{min}_{i}$, which is the energy used by the device with the minimum energy used by round $i$.
-This ratio is then raised to the power $M$, which determines how strong the effect will be $\ll$Need a better way of saying this..$\gg$.
+This ratio is then raised to the power $M$, which determines how strong the effect will be.
 If a device's energy is much greater than the current minimum, it will be avoided as a relay.
@@
 
-$\ll$”in what follows we ….
-(state simulation and hardware), implementation of cost function in Dijktra (ref),step wide pseudo code for algorithm, comparison approach to “normal” cost function with Dijkstra$\gg$
+\subsection{Algorithm Implementation}
+Detailed pseudo-code for the routing algorithm can be found in Algorithm~\ref{alg:dijkstra}. Similarly, code for energy used calculation can be found in Algorithm~\ref{alg:energyused}.
 
-\subsection{Algorithm Implementation}
-Detailed pseudo-code for the routing algorithm can be found in Algorithm~\ref{alg:dijkstra} in Appendix~\ref{appendix:algorithms}.
-Similarly, code for energy used calculation can be found in Algorithm~\ref{alg:energyused}.
+\begin{algorithm}[!ht]
+\caption{Dijkstra's Algorithm with new cost function}
+\label{alg:dijkstra}
+\begin{algorithmic}[1]
+\FOR{$i = 1$ to \# of Nodes} 
+	\STATE $node_i.visited \leftarrow 0$
+	\STATE $node_i.distance \leftarrow \infty$
+	\STATE $node_i.previous \leftarrow node_i$
+\ENDFOR
+\STATE $node_{AP}.distance \leftarrow 0$
+\STATE $energy_{min} = min(node.energy)$
+\WHILE{unvisited nodes available}
+	\STATE $node_{source} \leftarrow $ node with smallest distance
+	\FOR{$i = 1$ to \# of Links}
+		\IF{$link_i.source = node_{source}$}
+			\STATE $node_{dest} \leftarrow link_i.destination$
+			\IF{$node_{dest}.visited = 0$}
+				\STATE $cost \leftarrow link_i.power * (1 + node_{dest}.energy/energy_{min} )^M/2$
+				\STATE $newdistance \leftarrow node_{source}.distance + cost$
+				\IF{$distance < node_{dest}.distance$} 
+				\STATE$node_{dest}.distance \leftarrow newdistance$
+				\STATE$node_{dest}.previous \leftarrow node_{source}$
+				\ENDIF
+			\ENDIF
+		\ENDIF
+	\ENDFOR
+	\STATE $node_{source}.visited \leftarrow 1$
+\ENDWHILE
+\end{algorithmic}
+\end{algorithm}
 
-\section{Simulation}
-$\ll$describe experimental setups and environment (2 setup – walking in room and sitting in from of desk), figure 2 (residential environment for setups)$\gg$
+\begin{algorithm}[!ht]
+\caption{Energy Used Calculation}
+\label{alg:energyused}
+\begin{algorithmic}[1]
+\FOR{$i = 1$ to \# of Nodes}
+	\STATE $node_{ptr} = node_i$
+	\IF{$node_{ptr}.previous = node_{ptr}$}
+		\STATE No path to $node_i$
+	\ELSE
+		\WHILE{$node_{ptr}.previous \neq node_{ptr}$}
+			\STATE $link_{ptr} \leftarrow $ link between $node_{ptr}$ and $node_{ptr}.previous$
+			\STATE $node_{ptr}.energy += link_{ptr}.power$
+			\STATE $node_{ptr} \leftarrow = node_{ptr}.previous$
+		\ENDWHILE
+	\ENDIF
+\ENDFOR
+\end{algorithmic}
+\end{algorithm}
 
-\section{Hardware Implementation}
+\section{Simulation}\label{sec:simulation}
+In order to test the new cost function in an efficient manner, a simulation was used.
+Measurements of real channel data were acquired by running the procedure described in~\ref{subsec:cycleofoperations} without the routing or power control enabled.
+Each ED continuously sampled RSSI data from all other devices and transmitted it to the host.
+
+Two experimental setups were used to capture RSSI data.
+For the first, the subject remained relatively static in a sitting position.
+The second consisted of a more dynamic environment, where the subject walked around a room as seen in Figure~\ref{fig:floorplan}.
+
+A set of captured RSSI tables (as seen in Table~\ref{tab:rssi}) were fed into the simulation, which proceeded to run the routing algorithm.
+The simulation produced energy use data, along with power settings, and routes taken for all devices.
+
+\section{Hardware Implementation}\label{sec:hardware}
+In order to verify the new cost function in a real environment, a hardware platform was developed and used.
+The system function in a similar fashion to the simulation, except that the actual routing and power control data was fed back to the ED's.
+
 \subsection{Hardware Platform}
@@
 
-Two experimental setups were used.
-For the first, the subject remained relatively static in a sitting position.
-The second consisted of a more dynamic environment, where the subject walked around a room as seen in Figure~\ref{fig:floorplan}.
-The target RSSI ($RSSI^T$) was arbitrarily chosen to be $-60 dBm$.
+For both simulation and hardware experiments, the target RSSI ($RSSI^T$) was arbitrarily chosen to be $-60 dBm$.
 
@@
 
-\subsection{Cycle of Operations}
+\subsection{Cycle of Operations}\label{subsec:cycleofoperations}
 A normal cycle of operations with $n$ devices works as follows:
-\begin{enumerate}
-\item AP sends synchronization beacon which includes routing and power control tables.
-\item Each ED transmits its own RSSI table back to the AP, while simultaneously listening to other ED messages and storing the received power from each.
-\item Once all EDs have transmitted their data, the AP sends a table with the RSSI data from all devices. (See Table~\ref{tab:rssi} in Appendix~\ref{appendix:tables} for an example)
-\item The host uses the RSSI table to compute the routes along with the required powers to meet the selected links.
-\item The host sends both routing and power tables back to the AP so that a new cycle may begin.
-\end{enumerate}
+
+\begin{algorithmic}[1]
+\STATE AP sends synchronization beacon which includes routing and power control tables.
+\STATE Each ED transmits its own RSSI table back to the AP, while simultaneously listening to other ED messages and storing the received power from each.
+\STATE Once all EDs have transmitted their data, the AP sends a table with the RSSI data from all devices. (See Table~\ref{tab:rssi} for an example)
+\STATE The host uses the RSSI table to compute the routes along with the required powers to meet the selected links.
+\STATE The host sends both routing and power tables back to the AP so that a new cycle may begin.
+\end{algorithmic}
+
+\begin{table}[htb]
+\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}
+\hline  & AP & ED1 & ED2 & ED3 & ED4 & ED5 & ED6 & ED7 & ED8 \\ 
+\hline AP & - & -60 & -60 & -60 & -60.5 & -60.5 & -60.5 & -60 & -79.5 \\ 
+\hline ED1 & -66 & - & -62 & -75 & -63.5 & -69 & -75.5 & -73 & -67.5 \\ 
+\hline ED2 & -72 & -61.5 & - & -80 & -69 & -70 & -74 & -75 & -66 \\ 
+\hline ED3 & -77.5 & -74.5 & -80.5 & - & -67.5 & -70 & -72.5 & -70.5 & -70 \\ 
+\hline ED4 & -39 & -63 & -69.5 & -67 & - & -82.5 & -74.5 & -69 & -65.5 \\ 
+\hline ED5 & -67.5 & -68 & -71 & -69.5 & -84 & - & -55.5 & -51.5 & -62 \\ 
+\hline ED6 & -67 & -75 & -75 & -72.5 & -75 & -55.5 & - & -44.5 & -53 \\ 
+\hline ED7 & -46 & -70 & -77.5 & -70.5 & -69 & -50.5 & -45 & - & -51.5 \\ 
+\hline ED8 & -78 & -67.5 & -66.5 & -71 & -66 & -62 & -54 & -51.5 & - \\ 
+\hline 
+\end{tabular} 
+\caption{Sample RSSI Table}
+\label{tab:rssi}
+\end{table}
 
@@
 
-$\ll$describe change from simulation (RSSI from power controlled transmission, address benefits$\gg$
+\section{Results}
 
-\section{Results}
+Due to the similarities between the simulation and hardware implementation (they both use real RSSI data), the results were fairly similar.
+If fed with the same RSSI data, the algorithm will always produce the same results.
+The only difference with the hardware implementation is that the RSSI table is reduced due to power control.
+Since each ED only uses enough transmit power to reach a specific ED, others might not get the message.
+This was not a problem when the hardware experiment was run.
 
@@
 
-$\ll$describe figure  3 and interpret results – network lifetime (like fig.8 in paper with Jeff)$\gg$
-
-$\ll$describe figure  4 and interpret results – energy per bit (like fig.9 in paper with Jeff)$\gg$
-
 \section{Conclusion}
-$\ll$at the end: different from abstract -  The abstract in past tense + concentrate on meaning of results$\gg$
-
+After analyzing the results, the new link cost function successfully improves network lifetime.
+The new cost function outperformed the traditional one in trials dealing with a dynamic environment, where the subject was walking.
+When the subject was not moving, both link costs performed about the same.
+Note that, in this case, performance is with respect to network lifetime, and not overall power consumption.
+While the new link cost was successful, there is room for improvement.
 One improvement that can be done deals with the example from Figure~\ref{fig:walk4-c100}.
@@
 A better implementation would not include off-line devices in the calculation.
+A second modification would relate actual battery, or other power source, charge to the algorithm.
+This would allow for devices to join and leave the network while maintaining proper operation.
+It would also allow devices with different power sources to be included and used.
 
@@
 
-\appendix
-\section{Algorithms}\label{appendix:algorithms}
-\begin{algorithm}[!ht]
-\caption{Dijkstra's Algorithm with new cost function}
-\label{alg:dijkstra}
-\begin{algorithmic}
-\FOR{$i = 1$ to \# of Nodes} 
-	\STATE $node_i.visited \leftarrow 0$
-	\STATE $node_i.distance \leftarrow \infty$
-	\STATE $node_i.previous \leftarrow node_i$
-\ENDFOR
-\STATE $node_{AP}.distance \leftarrow 0$
-\STATE $energy_{min} = min(node.energy)$
-\WHILE{unvisited nodes available}
-	\STATE $node_{source} \leftarrow $ node with smallest distance
-	\FOR{$i = 1$ to \# of Links}
-		\IF{$link_i.source = node_{source}$}
-			\STATE $node_{dest} \leftarrow link_i.destination$
-			\IF{$node_{dest}.visited = 0$}
-				\STATE $cost \leftarrow link_i.power * (1 + node_{dest}.energy/energy_{min} )^M/2$
-				\STATE $newdistance \leftarrow node_{source}.distance + cost$
-				\IF{$distance < node_{dest}.distance$} 
-				\STATE$node_{dest}.distance \leftarrow newdistance$
-				\STATE$node_{dest}.previous \leftarrow node_{source}$
-				\ENDIF
-			\ENDIF
-		\ENDIF
-	\ENDFOR
-	\STATE $node_{source}.visited \leftarrow 1$
-\ENDWHILE
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}[!ht]
-\caption{Energy Used Calculation}
-\label{alg:energyused}
-\begin{algorithmic}
-\FOR{$i = 1$ to \# of Nodes}
-	\STATE $node_{ptr} = node_i$
-	\IF{$node_{ptr}.previous = node_{ptr}$}
-		\STATE No path to $node_i$
-	\ELSE
-		\WHILE{$node_{ptr}.previous \neq node_{ptr}$}
-			\STATE $link_{ptr} \leftarrow $ link between $node_{ptr}$ and $node_{ptr}.previous$
-			\STATE $node_{ptr}.energy += link_{ptr}.power$
-			\STATE $node_{ptr} \leftarrow = node_{ptr}.previous$
-		\ENDWHILE
-	\ENDIF
-\ENDFOR
-\end{algorithmic}
-\end{algorithm}
-
-\section{Tables}\label{appendix:tables}
-\begin{table}[htb]
-\label{tab:rssi}
-\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}
-\hline  & AP & ED1 & ED2 & ED3 & ED4 & ED5 & ED6 & ED7 & ED8 \\ 
-\hline AP & - & -60 & -60 & -60 & -60.5 & -60.5 & -60.5 & -60 & -79.5 \\ 
-\hline ED1 & -66 & - & -62 & -75 & -63.5 & -69 & -75.5 & -73 & -67.5 \\ 
-\hline ED2 & -72 & -61.5 & - & -80 & -69 & -70 & -74 & -75 & -66 \\ 
-\hline ED3 & -77.5 & -74.5 & -80.5 & - & -67.5 & -70 & -72.5 & -70.5 & -70 \\ 
-\hline ED4 & -39 & -63 & -69.5 & -67 & - & -82.5 & -74.5 & -69 & -65.5 \\ 
-\hline ED5 & -67.5 & -68 & -71 & -69.5 & -84 & - & -55.5 & -51.5 & -62 \\ 
-\hline ED6 & -67 & -75 & -75 & -72.5 & -75 & -55.5 & - & -44.5 & -53 \\ 
-\hline ED7 & -46 & -70 & -77.5 & -70.5 & -69 & -50.5 & -45 & - & -51.5 \\ 
-\hline ED8 & -78 & -67.5 & -66.5 & -71 & -66 & -62 & -54 & -51.5 & - \\ 
-\hline 
-\end{tabular} 
-\caption{Sample RSSI Table}
-\end{table}
-
 \end{document}

### 2011-09-15 docs/thesis/thesis.tex
Updated thesis.tex and added figures.

@@
 \begin{document}
-\title{Masters Thesis}
+\title{ (thesis title?)\\
+Masters Thesis}
 \author{Alvaro Prieto\\
@@
 \section{Summary of Contributions}
+The following is a list of contributions presented in this work.
 \begin{itemize}
-\item BSN Software Libraries
-\item wireless ECG platform
-\item New routing algorithm
+\item \textbf{Software platform for WBAN research.} - The software platform was developed using only free and open source software (FOSS) tools.
+Radio libraries for the TI CC430 device were ported over from TI provided code and libraries for the CC2500 radio were written from scratch.
+Other features include timer control for scheduling and time-stamping, serial communication, some digital I/O, and analog input.
+Included in the platform were limited software tools for data capture and visualization on a host computer.
+
+\item \textbf{Wireless ECG platform} - In order to test and demonstrate the WBAN platform, a wireless ECG system was implemented.
+The system included three wireless devices, an access point, and a host computer.
+Software for synchronized sampling was developed and tested successfully using the WBAN platform.
+Three ECG leads were sampled at 300Hz and displayed on a host computer in real-time.
+
+\item \textbf{Software platform for routing algorithm development} - In order to develop a novel routing algorithm, a platform to test and compare with others was developed.
+The application uses Dijkstra's algorithm for best path selection, and allows for the testing of various link-cost functions.
+Both simulation and real-time operation are supported.
+One unique feature is the generation of routing graphic generation for network visualization.
+
+\item \textbf{Specialized cost function for maximizing Body Area Network lifetime using global routing algorithms} - (Use abstract of paper here?)
 \end{itemize}
@@
 
-\section{Software Libraries}
+\section{Software Libraries}\label{sec:wbansoftwarelib}
 One of the primary contributions of this project consists of software libraries for working with WBANs.
@@
 The analog-to-digital(A/D) ports were used to attach any needed sensors.
+The actual CC430 model used was the CC430x6137.
 
@@
 This device is manufactured by TI and sold as a development platform.
-The MSP430-RF2500 device consists of an MSP430 microcontroller and a CC2500 radio.
+The MSP430-RF2500 device consists of an MSP430F2274 microcontroller and a CC2500 radio.
 There were several reasons why this device was used.
@@
 Features of the new radio libraries include power control, addressing and channel switching.
-
-\section{Specialized Cost Function for Maximizing Body Area Networks Lifetime Using Global Routing Algorithms}
-$\ll$ Insert most of the paper here. $\gg$
 
@@
 
+\begin{figure}[htb]
+\begin{center}
+\includegraphics[width=0.8\textwidth]{figures/heartbeats.pdf}
+\end{center}
+\caption{Sample Measurement from Single ECG Lead}
+\label{fig:heartbeats}
+\end{figure}
+
 \subsection{Host Interface}
@@
 
+\section{Algorithm Development Platform}
+A platform for testing and analyzing new cost functions for global routing algorithms was developed in C.
+The platform allows for a network of nodes and links between the nodes to be created.
+It computes the best path from any node to a central source node.
+Finally, it produces various statistics, such as energy use, routing maps, and graphical visualizations.
+
+Initially, the goal of the platform was to test and compare routing with various link costs in an efficient manner.
+After a good link cost function was found, it would then be implemented in a microcontroller and used.
+This did not turn out to be the case.
+The first problem encountered dealt with the network structure used for testing.
+The program required a list of nodes and a list of links between the nodes, along with the link costs.
+Developing a simulated network and links that accurately represented a real network was unfeasible.
+The choice was then made to closely couple the hardware design and the simulator.
+
+The best way to come up with a network for simulation was to get all of the node and link information from a real network.
+The WBAN platform was then used to develop a system for measuring RSSI data between all nodes in a network.
+This system allowed for the capture of real network data, both nodes and links, that could later be used to run routing simulations.
+The new program auto-generated a list of nodes, links, and link costs from the previously-measured RSSI data.
+This new approach effectively removed the need to come up with special networks for testing.
+
+In order to better understand how the new routing system performed, a series of output files were generated during simulation.
+The program was set up to account for the energy use of each node as it routed packets through the network.
+Visualizing the energy use was vital for the analysis of network lifetime.
+Other available data included the routing information itself, which allowed for the analysis of the routes as the algorithm runs.
+Finally, the program produced a graphical output.
+
+The graphical output began as a simple experiment to see a visual representation of the network.
+It was later expanded to generate graphs of the network, and all available connections, as the RSSI data was sampled.
+Due to the large number of links between all of the nodes, this graphic was not useful.
+The final implantation, which can be seen in Figure~\ref{fig:networkgraph1}, only displayed the links used in a specific round of routing.
+The link labels display the transmit power required by each node to meet the specific link.
+
 \begin{figure}[htb]
 \begin{center}
-\includegraphics[width=1\textwidth]{figures/heartbeats.pdf}
+\includegraphics[width=0.6\textwidth]{figures/networkgraph1.pdf}
 \end{center}
-\caption{Sample Measurement from Single ECG Lead}
-\label{fig:heartbeats}
+\caption{Sample Network Visualization}
+\label{fig:networkgraph1}
 \end{figure}
+
+\section{Specialized Cost Function for Maximizing Body Area Networks Lifetime Using Global Routing Algorithms}\label{sec:costfunction}
+$\ll$ Insert most of the paper here. $\gg$
 
@@
 ...
+
+In Chapter\ref{sec:wbansoftwarelib},
 

### 2011-09-15 docs/thesis/paper.tex
Updated thesis and paper tex files.

@@
 \label{eqn:link_cost}
-C^j_{j,k} = \frac{RSSI^T}{\alpha_{j,k}} \times \left( \frac{1+\left(\frac{E^k_i}{E^{min}_{i}} \right)^M}{2} \right)
+C^i_{j,k} = \frac{RSSI^T}{\alpha_{j,k}} \times \left( \frac{1+\left(\frac{E^k_i}{E^{min}_{i}} \right)^M}{2} \right)
 \end{equation}
@@
 If fed with the same RSSI data, the algorithm will always produce the same results.
-The only difference with the hardware implementation is that the RSSI table is reduced due to power control.
-Since each ED only uses enough transmit power to reach a specific ED, others might not get the message.
+The only difference with the hardware implementation is that the RSSI table is reduced due to the use of power control.
+Since each ED only uses enough transmit power to reach a specific ED, others might not receive the message.
 This was not a problem when the hardware experiment was run.
@@
 All EDs consume energy at different rates for both $M=0$ and $M=100$.
+A possible explanation for this is that the links between EDs remain close to the same as time passes.
+This can become a problem because certain EDs may only have a single path to choose from.
+If this is the case, they have no choice but to use the same link continuously.
+Not having a choice of paths impairs the ability of the routing algorithm to distribute energy use among all devices.
 
 Figures~\ref{fig:walk2-c0}~and~\ref{fig:walk2-c100}, on the other hand, show an increase in network lifetime.
-At the end of the 5 minute sampling time with $M=0$, the device which consumed the most power was ED7 with approximately 1.7 mW.
+For example, at the end of the 5 minute sampling run with $M=0$, the device which consumed the most power was ED7 with approximately 1.7 mW.
 On the other hand, with $M=100$, all EDs consume power in a uniform fashion and end up with less than 1.3 mW each.
@@
 When the subject was not moving, both link costs performed about the same.
-Note that, in this case, performance is with respect to network lifetime, and not overall power consumption.
+Note that, in this case, performance is defined with respect to network lifetime, and not overall power consumption.
+
 While the new link cost was successful, there is room for improvement.
@@
 
+The use of power control, coupled with this new link cost function, should allow WBANs to run efficiently for long periods of time, while maximizing the network lifetime.
+Networks requiring all devices to be on-line to function will greatly benefit from this routing method.
+Other networks can also benefit.
+Due to the normalization of energy use in the network, devices will deplete their energy sources at similar times.
+This could be beneficial as all devices can be recharged or replaced simultaneously, instead of constantly monitoring and replacing individual devices.
+
 \bibliography{sources/thesis.bib}{}

### 2011-09-15 docs/thesis/thesis.tex
Updated thesis and paper tex files.

@@
 \documentclass{article}
+\usepackage{algorithm}
+\usepackage{algorithmic}
 \usepackage{graphicx}
 \begin{document}
-\title{ (thesis title?)\\
+\title{Specialized Cost Function for Maximizing Body Area Network Lifetime Using Global Routing Algorithms\\
 Masters Thesis}
@@
 
-\begin{abstract}
-...
-\end{abstract}
+\addcontentsline{toc}{section}{\protect\numberline{}Abstract}
+\section*{Abstract}
+Wireless Body Area Networks (WBANs) consist of several wireless sensors located around a human body.
+These sensors may measure several biological signals, movement, and temperature.
+Due to major improvements in power consumption and constantly shrinking devices, WBANs are becoming ubiquitous.
+As a side effect present because of the small form factor of these devices, the battery size is limited.
+While the sensors themselves may be extremely power efficient, all of the measured data must be transmitted over a much less efficient wireless link.
+One benefit of WBANs is that they rarely include more than a dozen wireless devices over a small area.
+This constraint allows for the use of routing techniques not suitable for larger wireless sensor networks(WSNs).
+
+This work primarily focuses on maximizing the lifetime of WBANs.
+It also includes a basic software framework for developing WBANs, a sample Wireless Electrocardiogram (ECG) application, and a simple link cost algorithm development platform.
 
@@
 \item \textbf{Software platform for WBAN research.} - The software platform was developed using only free and open source software (FOSS) tools.
-Radio libraries for the TI CC430 device were ported over from TI provided code and libraries for the CC2500 radio were written from scratch.
+Radio libraries for the Texas Instruments (TI) CC430 device were ported over from TI reference code, while libraries for the CC2500 radio were written from scratch.
 Other features include timer control for scheduling and time-stamping, serial communication, some digital I/O, and analog input.
@@
 
-\item \textbf{Specialized cost function for maximizing Body Area Network lifetime using global routing algorithms} - (Use abstract of paper here?)
+\item \textbf{Specialized cost function for maximizing WBAN lifetime using global routing algorithms} - A novel routing algorithm is developed that focuses not on single devices, but the network as a whole.
+Current methods focus on maximizing individual device lifetime or minimizing the energy-per-bit.
+While those are generally good goals, certain systems require all devices functioning to be considered operational.
+Keeping that in mind, the proposed system maximizes the network lifetime by normalizing power use across the network.
+For the purpose of this paper, \textbf{network lifetime} is defined as the time it takes any single node to deplete its energy source.
+
 \end{itemize}
@@
 Features of the new radio libraries include power control, addressing and channel switching.
+
+\begin{figure}[htb]
+\begin{center}
+\includegraphics[width=0.9\textwidth]{figures/ez430-rf2500.pdf}
+\end{center}
+\caption{EZ430-RF2500 with Battery Pack}
+\label{fig:ez430}
+\end{figure}
 
@@
 
+\subsection{Initial Development}
 Initially, the goal of the platform was to test and compare routing with various link costs in an efficient manner.
@@
 
+\subsection{Hardware Coupling}
 The best way to come up with a network for simulation was to get all of the node and link information from a real network.
@@
 
+\subsection{Simulation Output}
 In order to better understand how the new routing system performed, a series of output files were generated during simulation.
@@
 Other available data included the routing information itself, which allowed for the analysis of the routes as the algorithm runs.
-Finally, the program produced a graphical output.
-
-The graphical output began as a simple experiment to see a visual representation of the network.
-It was later expanded to generate graphs of the network, and all available connections, as the RSSI data was sampled.
-Due to the large number of links between all of the nodes, this graphic was not useful.
-The final implantation, which can be seen in Figure~\ref{fig:networkgraph1}, only displayed the links used in a specific round of routing.
-The link labels display the transmit power required by each node to meet the specific link.
+During run-time, the program prints out text-based routing information in a terminal window(Figure~\ref{fig:textoutput}).
+This provided real-time feedback that was useful during hardware debugging.
 
@@
 
+Finally, the program produced a graphical output.
+The graphical output began as a simple experiment to see a visual representation of the network.
+It was later expanded to generate graphs of the network, and all available connections, as the RSSI data was sampled.
+Due to the large number of links between all of the nodes, this graphic was not useful.
+The final implementation, which can be seen in Figure~\ref{fig:networkgraph1}, only displayed the links used in a specific round of routing.
+The link labels display the transmit power required by each node to meet the specific link.
+The generated graphs can later be put together to form an animation that shows the network changing in action.
+
+\begin{figure}[htb]
+\begin{center}
+\includegraphics[width=0.5\textwidth]{figures/textoutput.pdf}
+\end{center}
+\caption{Sample Run-time Output}
+\label{fig:textoutput}
+\end{figure}
+
 \section{Specialized Cost Function for Maximizing Body Area Networks Lifetime Using Global Routing Algorithms}\label{sec:costfunction}
-$\ll$ Insert most of the paper here. $\gg$
+\subsection{Introduction}
+
+There are several routing protocols designed specifically for WBANs.
+Braem et al.
+developed the CICADA protocol, which consists of a spanning tree architecture with a time-division scheme for transmission scheduling~\cite{protocol:CICADA}.
+The primary downside to this method is that nodes closer to the root will deplete their energy source faster due to the need to relay messages from children nodes.
+Other protocols focus on the reliability of the network~\cite{routing:storeandforward}.
+Dynamic networks are a problem with WBANs.
+If sensors are placed in the extremities, the network will change as the body changes positions.
+Quwaider et al.
+developed a protocol tolerant to network changes~\cite{routing:storeandforward}.
+They propose a store-and-forward method that maximizes the likelihood of a packet reaching its destination.
+Each packet is stored by multiple devices and retransmitted, which consumes more power.
+One solution for this problem was proposed by Ehyaie et al.~\cite{relay:networklife}, it consists of using dedicated, non-sensing, relay nodes with larger power sources.
+While this method increases the network lifetime, it requires more hardware.
+Finally, Nabi et al.
+propose a similar store-and-forward method, but integrates Transmit Power Adaptation (TPA)~\cite{relay:transmitpoweradaptation}.
+Nodes keep track of neighbors and use power control to use the smallest transmit power while maintaining a specific link quality.
+
+In order to provide accurate channel conditions, RSSI measurements of an eight ED network were used as input to our simulations(\ref{sec:simulation}).
+Real-time hardware implementations were also run to demonstrate the algorithm(\ref{sec:hardware}).
+The new cost function succeeded in improving network lifetimes when run in a dynamic environment.
+Static environments did not produce successful results.
+
+\begin{figure}[!htb]
+\begin{center}
+\includegraphics[width=0.5\textwidth]{figures/body.pdf}
+\end{center}
+\caption{Experimental Setup (Not all links shown)}
+\label{fig:body}
+\end{figure}
+
+\subsection{Proposed Cost Function}
+In the proposed system, we define a new link cost function that is then used in a routing algorithm.
+Traditionally, the power required to make a link possible is used as the link cost.
+In this case, the total energy used by a device is factored into the equation.
+If a device has used more energy than the others, its use as a relay will be discouraged.
+
+\subsubsection{Device Energy}
+Each device's normalized energy use is calculated as shown in Equation~\ref{eqn:accumulated_energy}, where $j$ denotes the device ID, $i$ is the current polling round, $\alpha_{j,k}$ is the channel attenuation for the selected link, and $RSSI^T$ is the target RSSI.
+The current energy used is incremented by the energy used to transmit a single packet while maintaining $RSSI^T$.
+The channel attenuation for the selected link,$\alpha_{j,k}$, can be seen in Equation~\ref{eqn:alpha}, where RSSI is the received power measured by device $k$ and $P_{tx}$ is the transmitted power used by device $j$.
+
+\subsubsection{Link Cost}
+The actual link cost is computed by calculating the energy used by the device, \emph{if} that link is selected, and multiplying by a cost factor(Equation~\ref{eqn:link_cost}).
+The cost factor calculation takes the energy used by the destination device, $E^k_i$, and divides it by the current energy minimum, $E^{min}_{i}$, which is the energy used by the device with the minimum energy used by round $i$.
+This ratio is then raised to the power $M$, which determines how strong the effect will be.
+If a device's energy is much greater than the current minimum, it will be avoided as a relay.
+The $M$ factor in Equation~\ref{eqn:link_cost} can be any number in $0 < M < \infty$.
+The cost factor is normalized so when $M=0$, it returns to the traditional cost of required power to meet the link.
+
+Since the link costs are directly related to the destination device, they are calculated during the routing algorithm computation.
+For this paper, Dijkstra's algorithm~\cite{dijkstra:algorithm} is used to select the best route.
+
+\begin{equation}
+\label{eqn:alpha}
+\alpha_{j,k} = \frac{RSSI}{P_{tx}}
+\end{equation}
+\begin{equation}
+\label{eqn:accumulated_energy}
+E^j_{(i)} = E^j_{(i-1)} + \frac{RSSI^T}{\alpha}
+\end{equation}
+\begin{equation}
+\label{eqn:link_cost}
+C^i_{j,k} = \frac{RSSI^T}{\alpha_{j,k}} \times \left( \frac{1+\left(\frac{E^k_i}{E^{min}_{i}} \right)^M}{2} \right)
+\end{equation}
+
+\subsubsection{Assumptions}
+This algorithm works under several assumptions.
+The first being that this is a centralized network with one Access Point (AP) with virtually unlimited energy and computing power.
+The second is that each wireless sensor, or End Device(ED), can act as a relay for messages from other nodes.
+The Third is that the link quality, or channel attenuation, between all EDs and AP is known.
+Finally, all EDs begin with an equal energy used, that is, they all have the same energy stored.
+All routing calculations take place in the AP.
+Figure~\ref{fig:body} shows a sample network with EDs labeled 1 through 8, and one AP.
+
+\subsection{Performance Analysis}
+In what follows, we analyze the performance of the proposed cost function.
+The new cost function was tested in both simulated and experimental environments.
+In order to gauge the efficiency of the cost function, it will be compared to an identical system which uses a traditional link cost function.
+The traditional link cost function consists of the required power to meet a link with the desired $RSSI^T$.
+This can be achieved by setting a value of $M=0$ in the new link cost function.
+
+Performance is determined by the time it takes any node's accumulated normalized energy to cross an arbitrary threshold.
+The threshold could be equivalent to the amount of energy stored in a device battery, but for this experiment, is purely arbitrary.
+
+\subsubsection{Algorithm Implementation}
+Detailed pseudo-code for the routing algorithm can be found in Algorithm~\ref{alg:dijkstra}. Similarly, code for energy used calculation can be found in Algorithm~\ref{alg:energyused}.
+
+\begin{algorithm}[!ht]
+\caption{Dijkstra's Algorithm with new cost function}
+\label{alg:dijkstra}
+\begin{algorithmic}[1]
+\FOR{$i = 1$ to \# of Nodes} 
+	\STATE $node_i.visited \leftarrow 0$
+	\STATE $node_i.distance \leftarrow \infty$
+	\STATE $node_i.previous \leftarrow node_i$
+\ENDFOR
+\STATE $node_{AP}.distance \leftarrow 0$
+\STATE $energy_{min} = min(node.energy)$
+\WHILE{unvisited nodes available}
+	\STATE $node_{source} \leftarrow $ node with smallest distance
+	\FOR{$i = 1$ to \# of Links}
+		\IF{$link_i.source = node_{source}$}
+			\STATE $node_{dest} \leftarrow link_i.destination$
+			\IF{$node_{dest}.visited = 0$}
+				\STATE $cost \leftarrow link_i.power * (1 + node_{dest}.energy/energy_{min} )^M/2$
+				\STATE $newdistance \leftarrow node_{source}.distance + cost$
+				\IF{$distance < node_{dest}.distance$} 
+				\STATE$node_{dest}.distance \leftarrow newdistance$
+				\STATE$node_{dest}.previous \leftarrow node_{source}$
+				\ENDIF
+			\ENDIF
+		\ENDIF
+	\ENDFOR
+	\STATE $node_{source}.visited \leftarrow 1$
+\ENDWHILE
+\end{algorithmic}
+\end{algorithm}
+
+\begin{algorithm}[!ht]
+\caption{Energy Used Calculation}
+\label{alg:energyused}
+\begin{algorithmic}[1]
+\FOR{$i = 1$ to \# of Nodes}
+	\STATE $node_{ptr} = node_i$
+	\IF{$node_{ptr}.previous = node_{ptr}$}
+		\STATE No path to $node_i$
+	\ELSE
+		\WHILE{$node_{ptr}.previous \neq node_{ptr}$}
+			\STATE $link_{ptr} \leftarrow $ link between $node_{ptr}$ and $node_{ptr}.previous$
+			\STATE $node_{ptr}.energy += link_{ptr}.power$
+			\STATE $node_{ptr} \leftarrow = node_{ptr}.previous$
+		\ENDWHILE
+	\ENDIF
+\ENDFOR
+\end{algorithmic}
+\end{algorithm}
+
+\subsection{Simulation}\label{sec:simulation}
+In order to test the new cost function in an efficient manner, a simulation was used.
+Measurements of real channel data were acquired by running the procedure described in~\ref{subsec:cycleofoperations} without the routing or power control enabled.
+Each ED continuously sampled RSSI data from all other devices and transmitted it to the host.
+
+Two experimental setups were used to capture RSSI data.
+For the first, the subject(See Figure~\ref{fig:halfbody}) remained relatively static in a sitting position.
+The second consisted of a more dynamic environment, where the subject walked around a room as seen in Figure~\ref{fig:floorplan}.
+
+A set of captured RSSI tables (as seen in Table~\ref{tab:rssi}) were fed into the simulation, which proceeded to run the routing algorithm.
+The simulation produced energy use data, along with power settings, and routes taken for all devices.
+
+\subsection{Hardware Implementation}\label{sec:hardware}
+In order to verify the new cost function in a real environment, a hardware platform was developed and used.
+The system function in a similar fashion to the simulation, except that the actual routing and power control data was fed back to the ED's.
+
+\subsubsection{Hardware Platform}
+The hardware platform used for this test consists of the Texas Instruments (TI) EZ430-RF2500.
+This device includes both an MSP430F2274 microcontroller along with a CC2500 2.4GHz transceiver.
+The CC2500 is configured to run at 250kbps.
+A single device, labeled the Access Point (AP), is connected via a USB to Serial link, running at 115200 BAUD, to the host computer.
+All other devices, labeled End Devices (EDs), are battery powered.
+In this implementation, the AP acts only as a bridge between the host and the end devices.
+All routing and power control calculations are done on the host.
+
+The host is a laptop computer with an Intel(R) Core(TM) i5-2410M CPU and 4GB of RAM.
+Eight EDs were placed on a 170cm, 70kg male subject as shown in Figure~\ref{fig:body}.
+The host was carried on a backpack and connected to the AP via a USB cable.
+The data was sampled, and the routing algorithm was run, at a rate of 5Hz.
+This sampling rate is appropriate as long as the subject, or its environment, does not move faster than this.$\ll$What was the name for this?$\gg$ If data needs to be sampled at a faster rate, some hardware and firmware changes might be necessary, but the host software will still work.
+
+For both simulation and hardware experiments, the target RSSI ($RSSI^T$) was arbitrarily chosen to be $-60 dBm$.
+
+\begin{figure}[!htb]
+\begin{center}
+\includegraphics[width=0.6\textwidth]{figures/floorplan.pdf}
+\end{center}
+\caption{Walking environment.}
+\label{fig:floorplan}
+\end{figure}
+
+\subsubsection{Cycle of Operations}\label{subsec:cycleofoperations}
+A normal cycle of operations with $n$ devices works as follows:
+
+\begin{algorithmic}[1]
+\STATE AP sends synchronization beacon which includes routing and power control tables.
+\STATE Each ED transmits its own RSSI table back to the AP, while simultaneously listening to other ED messages and storing the received power from each.
+\STATE Once all EDs have transmitted their data, the AP sends a table with the RSSI data from all devices. (See Table~\ref{tab:rssi} for an example)
+\STATE The host uses the RSSI table to compute the routes along with the required powers to meet the selected links.
+\STATE The host sends both routing and power tables back to the AP so that a new cycle may begin.
+\end{algorithmic}
+
+\begin{table}[htb]
+\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}
+\hline  & AP & ED1 & ED2 & ED3 & ED4 & ED5 & ED6 & ED7 & ED8 \\ 
+\hline AP & - & -60 & -60 & -60 & -60.5 & -60.5 & -60.5 & -60 & -79.5 \\ 
+\hline ED1 & -66 & - & -62 & -75 & -63.5 & -69 & -75.5 & -73 & -67.5 \\ 
+\hline ED2 & -72 & -61.5 & - & -80 & -69 & -70 & -74 & -75 & -66 \\ 
+\hline ED3 & -77.5 & -74.5 & -80.5 & - & -67.5 & -70 & -72.5 & -70.5 & -70 \\ 
+\hline ED4 & -39 & -63 & -69.5 & -67 & - & -82.5 & -74.5 & -69 & -65.5 \\ 
+\hline ED5 & -67.5 & -68 & -71 & -69.5 & -84 & - & -55.5 & -51.5 & -62 \\ 
+\hline ED6 & -67 & -75 & -75 & -72.5 & -75 & -55.5 & - & -44.5 & -53 \\ 
+\hline ED7 & -46 & -70 & -77.5 & -70.5 & -69 & -50.5 & -45 & - & -51.5 \\ 
+\hline ED8 & -78 & -67.5 & -66.5 & -71 & -66 & -62 & -54 & -51.5 & - \\ 
+\hline 
+\end{tabular} 
+\caption{Sample RSSI Table}
+\label{tab:rssi}
+\end{table}
+
+\subsubsection{Details}
+To minimize the number of control packets from being transmitted, the EDs are not individually polled.
+The only control packet that is sent is the synchronization beacon, which also carries the routing and power tables.
+Once the EDs are synchronized, they transmit their data on a pre-defined schedule to avoid collisions.
+Each ED has a network ID ranging from 1 to $n$.
+The time between synchronization packets is divided into time-slots than are used by each ED.
+The time slot used depends on the network ID of each device.
+This removes the need to come up with time schedules during runtime.
+
+The routing table is a simple array which lists the destination for each ED packet.
+The ED does not need to know the entire route its packets will take, but only next device in the path.
+Similarly, the power table lists the transmit power setting each device needs to use.
+The size of these tables is directly proportional to the number of devices in the network.
+
+\begin{figure}[!htb]
+\begin{center}
+\includegraphics[width=0.6\textwidth]{figures/halfbody.pdf}
+\end{center}
+\caption{Subject with devices for RSSI capture}
+\label{fig:halfbody}
+\end{figure}
+
+\subsection{Results}
+
+Due to the similarities between the simulation and hardware implementation (they both use real RSSI data), the results were fairly similar.
+If fed with the same RSSI data, the algorithm will always produce the same results.
+The only difference with the hardware implementation is that the RSSI table is reduced due to the use of power control.
+Since each ED only uses enough transmit power to reach a specific ED, others might not receive the message.
+This was not a problem when the hardware experiment was run.
+
+\begin{figure}[!ht]
+\includegraphics[width=0.8\textwidth]{figures/sit1-c0.pdf}
+\caption{Sitting M=0}
+\label{fig:sit1-c0}
+\end{figure}
+
+\begin{figure}[!ht]
+\includegraphics[width=0.8\textwidth]{figures/sit1-c100.pdf}
+\caption{Sitting M=100}
+\label{fig:sit1-c100}
+\end{figure}
+
+Figures~\ref{fig:sit1-c0}~and~\ref{fig:sit1-c100} show the resulting accumulated energies while the subject was in a sitting position.
+The experiment was run with both an $M=0$ as a reference, and $M=100$.
+Unfortunately, the results with a static subject were rather neutral.
+All EDs consume energy at different rates for both $M=0$ and $M=100$.
+A possible explanation for this is that the links between EDs remain close to the same as time passes.
+This can become a problem because certain EDs may only have a single path to choose from.
+If this is the case, they have no choice but to use the same link continuously.
+Not having a choice of paths impairs the ability of the routing algorithm to distribute energy use among all devices.
+
+Figures~\ref{fig:walk2-c0}~and~\ref{fig:walk2-c100}, on the other hand, show an increase in network lifetime.
+For example, at the end of the 5 minute sampling run with $M=0$, the device which consumed the most power was ED7 with approximately 1.7 mW.
+On the other hand, with $M=100$, all EDs consume power in a uniform fashion and end up with less than 1.3 mW each.
+One important observation is that the network with $M=0$ consumed less overall power (~9.6mW) than the one with $M=100$ (~10.1mW).
+This is expected, due to the fact that the new cost function specifically avoids the most efficient path for a specific node in favor of the most efficient path for the whole network.
+While more energy is consumed overall, no single ED consumes significantly more power.
+
+To demonstrate the network lifetime improvement, an ED will be considered have depleted it's power source after consuming 1mW of normalized energy.
+In the previous example, the trial with $M=0$ reaches 1mW after approximately 175 seconds, while the case with $M=100$ reaches it around 240 seconds.
+
+\begin{figure}[!ht]
+\includegraphics[width=0.8\textwidth]{figures/walk2-c0.pdf}
+\caption{Walking M=0}
+\label{fig:walk2-c0}
+\end{figure}
+
+\begin{figure}[!ht]
+\includegraphics[width=0.8\textwidth]{figures/walk2-c100.pdf}
+\caption{Walking M=100}
+\label{fig:walk2-c100}
+\end{figure}
+
+A good example of the algorithm working was a result of hardware problems.
+During of the experiments, a single ED went off-line for several seconds and was manually restarted.
+Figure~\ref{fig:walk4-c100} shows the effect the ED failure had on the network.
+Because a device that is powered off does not consume power, it became the one with the least energy used.
+When it returned to the network, it was immediately used as a relay by other nodes.
+
+\begin{figure}[!ht]
+\includegraphics[width=0.8\textwidth]{figures/walk4-c100.pdf}
+\caption{Walking M=100}
+\label{fig:walk4-c100}
+\end{figure}
 
 \section{Conclusion}
-...
+The developed WBAN software libraries were successfully used to implement both the wireless ECG system and the full-body routing system.
+The wireless ECG system was able to sample data at a rate of 300Hz and display it in the host computer.
+The routing algorithm testing platform facilitated the development of the new link cost function to maximize network lifetime.
 
-In Chapter\ref{sec:wbansoftwarelib},
+After analyzing the results, the new link cost function successfully improves network lifetime.
+The new cost function outperformed the traditional one in trials dealing with a dynamic environment, where the subject was walking.
+When the subject was not moving, both link costs performed about the same.
+Note that, in this case, performance is defined with respect to network lifetime, and not overall power consumption.
+
+While the new link cost was successful, there is room for improvement.
+One improvement that can be done deals with the example from Figure~\ref{fig:walk4-c100}.
+While the failed ED was not part of the network, it was still taken into account when it came to link cost calculation.
+A better implementation would not include off-line devices in the calculation.
+A second modification would relate actual battery, or other power source, charge to the algorithm.
+This would allow for devices to join and leave the network while maintaining proper operation.
+It would also allow devices with different power sources to be included and used.
+
+The use of power control, coupled with this new link cost function, should allow WBANs to run efficiently for long periods of time, while maximizing the network lifetime.
+Networks requiring all devices to be on-line to function will greatly benefit from this routing method.
+Other networks can also benefit.
+Due to the normalization of energy use in the network, devices will deplete their energy sources at similar times.
+This could be beneficial as all devices can be recharged or replaced simultaneously, instead of constantly monitoring and replacing individual devices.
 

### 2011-09-15 sim/lib/dijkstra.c
Removed many printout messages in readcsv.c and dikstra.c Set dot output to SVG for graph generation.

@@ -513,8 +513,6 @@ void print_shortest_path( uint8_t node_id )
     print_node_name( p_node->id );
     while( p_node->p_previous != p_node )
     {
-      printf("-(%g)",
-              find_link( p_node->p_previous->id, p_node->id )->links_power);
       p_node = p_node->p_previous;
       printf("->");
       print_node_name( p_node->id );

### 2011-09-15 sim/readcsv/main.c

@@ -190,7 +190,7 @@ uint8_t read_power_line ( FILE* fp_powers, energy_t* power_line )
   char csv_line[INBUFSIZE];   // Buffer for reading a line in the file
   char *p_item;
   uint16_t item_index = 0;
-  printf("Reading power line\n");
+
   if ( NULL != fgets( csv_line, sizeof(csv_line), fp_powers ) )
   {
     p_item = strtok( csv_line, "," );
@@ -200,11 +200,11 @@ uint8_t read_power_line ( FILE* fp_powers, energy_t* power_line )
     
       // Convert string to RSSI value and store in rssi_line
       power_line[item_index] = (energy_t)strtod( p_item, NULL );
-      printf( "%g ", power_line[item_index] );
+      
       item_index++;
       
     }
-    printf("\n");
+   
     // Read table successfully
     return 1;
   }
@@ -223,13 +223,11 @@ uint8_t read_table ( FILE* fp_csv_file, energy_t p_rssi_table[][MAX_DEVICES+1] )
 {
   char csv_line[INBUFSIZE];   // Buffer for reading a line in the file
   uint16_t line_index = 0;
-  printf("Reading table\n");
   while( NULL != fgets( csv_line, sizeof(csv_line), fp_csv_file ) )
   {
     // Detect empty line
     if( csv_line[0] == '\n' )
     {
-      printf("Read table\n");
       // Read table successfully
       return 1;
     }
@@ -325,7 +323,8 @@ void *graph_thread()
       //system("neato -Tsvg -ograph.svg graph.gv");
       //system("dot -Tsvg -ograph.svg graph.gv");
       //sprintf( command, "neato -Tjpg -oimages/graph%04d.jpg images/graph.gv", file_number);
-      sprintf( command, "dot -Tjpg -Gsize=4,4! -Gratio=fill -oimages/graph%05d.jpg %s", frame, filename );
+      //sprintf( command, "dot -Tjpg -Gsize=4,4! -Gratio=fill -oimages/graph%05d.jpg %s", frame, filename );
+      sprintf( command, "dot -Tsvg -Gsize=5,5! -Gratio=fill -oimages/graph%05d.svg %s", frame, filename );
       system( command );
 
       // Animate with ffmpeg -r 10 -b 500000 -i graph%05d.jpg ./animation.mp4

### 2011-09-16 docs/thesis/thesis.tex
Re-formatted thesis.

@@
 \documentclass{article}
+
 \usepackage{algorithm}
@@
 \usepackage{graphicx}
+\usepackage{fullpage}
+\usepackage[top=1.25in, bottom=1.25in, left=1.5in, right=1.0in]{geometry}
+
+\usepackage{setspace}
+\doublespacing
+
+%\addtolength{\oddsidemargin}{0.5in}
+%\addtolength{\evensidemargin}{0.5in}
+%\addtolength{\topmargin}{0.25in}
+%\addtolength{\textwidth}{-0.5in}
+
 \begin{document}
@@
 BS/MS Electrical Engineering\\
+Electrical and Microelectronic Engineering Department\\
+Rochester Institute of Technology\\
+Rochester, NY\\
 \texttt{axp8664@rit.edu}}
+
+\thispagestyle{empty}
 \maketitle
@@
 
-This work primarily focuses on maximizing the lifetime of WBANs.
-It also includes a basic software framework for developing WBANs, a sample Wireless Electrocardiogram (ECG) application, and a simple link cost algorithm development platform.
+Presented in this work is a novel global routing algorithm link-cost function to maximize network lifetime in WBANs.
+Also included are a basic software framework for developing WBANs, a sample Wireless Electrocardiogram (ECG) application, and a simple link cost algorithm development platform.
 
-\section{Summary of Contributions}
+\pagebreak
+
+\listoffigures
+
+\pagebreak
+
+\section*{Summary of Contributions}
 The following is a list of contributions presented in this work.
@@
 For the purpose of this paper, \textbf{network lifetime} is defined as the time it takes any single node to deplete its energy source.
+
+\item \textbf{Publication} A. Prieto, G. R. Tsouri, "Efficient Global Routing Algorithm for Prolonging Network Lifetime in Body Area Networks", in preparation IEEE Wireless Communication Letters.
 
@@
 This is expected, due to the fact that the new cost function specifically avoids the most efficient path for a specific node in favor of the most efficient path for the whole network.
-While more energy is consumed overall, no single ED consumes significantly more power.
+While more energy is consumed overall, no single ED consumes much more power than others.
+
+Another way of interpreting these results is that energy use in the network is more efficient.
+When a single ED depletes its energy source using the traditional link cost, there is a significant amount of energy left unused in the network.
+At the same time, when an ED dies in a network using the new link cost, there should be much less energy left in the rest of the network.
 
@@
 The developed WBAN software libraries were successfully used to implement both the wireless ECG system and the full-body routing system.
-The wireless ECG system was able to sample data at a rate of 300Hz and display it in the host computer.
+The wireless ECG system was able to sample data at a rate of 300Hz and display in real-time through the host computer.
 The routing algorithm testing platform facilitated the development of the new link cost function to maximize network lifetime.
 
-After analyzing the results, the new link cost function successfully improves network lifetime.
+After analyzing the results, the new link cost function successfully improved network lifetime.
 The new cost function outperformed the traditional one in trials dealing with a dynamic environment, where the subject was walking.
@@
 
-\listoffigures
-
-%\listoftables
-
 \bibliography{sources/thesis.bib}{}

### 2011-09-16 docs/thesis/thesis.tex
Changed title, removed page number from title page, added links.

@@
 
-%\addtolength{\oddsidemargin}{0.5in}
-%\addtolength{\evensidemargin}{0.5in}
-%\addtolength{\topmargin}{0.25in}
-%\addtolength{\textwidth}{-0.5in}
+\usepackage{hyperref}
 
 \begin{document}
-\title{Specialized Cost Function for Maximizing Body Area Network Lifetime Using Global Routing Algorithms\\
-Masters Thesis}
+\title{Body Area Network Platform Utilizing Energy-Efficient Routing of Physiological Data}
 \author{Alvaro Prieto\\
+Masters Thesis\\
 BS/MS Electrical Engineering\\
@@
 
+\maketitle
 \thispagestyle{empty}
-\maketitle
 
@@
 
+\addcontentsline{toc}{section}{\protect\numberline{}Table of Contents}
 \tableofcontents
@@
 
+\addcontentsline{toc}{section}{\protect\numberline{}List of Figures}
 \listoffigures
@@
 The generated graphs can later be put together to form an animation that shows the network changing in action.
+A sample video of the generated routing animation can be seen in \url{http://www.youtube.com/watch?v=5Mpn-_FFJaw&hd=1}.
 
@@
 
-\begin{figure}[!ht]
+\begin{figure}[!htb]
 \includegraphics[width=0.8\textwidth]{figures/sit1-c0.pdf}
@@
 
-\begin{figure}[!ht]
+\begin{figure}[!htb]
 \includegraphics[width=0.8\textwidth]{figures/sit1-c100.pdf}
@@
 
-\begin{figure}[!ht]
+\begin{figure}[!htb]
 \includegraphics[width=0.8\textwidth]{figures/walk2-c0.pdf}
@@
 
-\begin{figure}[!ht]
+\begin{figure}[!htb]
 \includegraphics[width=0.8\textwidth]{figures/walk2-c100.pdf}
@@
 
-\pagebreak
-
 \bibliography{sources/thesis.bib}{}

### 2011-09-23 docs/thesis/sources/literature_survey.tex
Final thesis draft.

@@
 
-There are several routing protocols currently used in WSNs~\cite{survey:wirelessrouting}, unfortunately, most are not well suited for BSNs.
+There are several routing protocols currently used in WSNs~\cite{survey:wirelessrouting}, unfortunately, most are not well suited for WBANs.
 Several of the protocols focus on mobility, Quality of Service (QoS), scalability and reliability, among other things.
-In general, BSNs do not have the same requirements as other WSNs.
-For example, in BSNs, scalability is no longer a problem since networks are limited to a human body.
+In general, WBANs do not have the same requirements as other WSNs.
+For example, in WBANs, scalability is no longer a problem since networks are limited to a human body.
 In most cases, power consumption becomes the main limiting factor for overall system performance.
 
-Some routing protocols are specifically designed with BSN constraints in mind.
+Some routing protocols are specifically designed with WBAN constraints in mind.
 One example is the Wireless Autonomous Spanning tree Protocol (WASP) \cite{protocol:WASP}.
@@
 
-Other BSN routing methods focus on postural mobility~\cite{routing:storeandforward}.
+Other WBAN routing methods focus on postural mobility~\cite{routing:storeandforward}.
 Quwaider et al.
@@
 
-Most current routing methods for BSNs use network trees.
+Most current routing methods for WBANs use network trees.
 One of the main problems present with network trees is that the nodes closer to the main node deplete their energy supply faster than children nodes.
@@
 
-So far, most BSNs have used tree-like networks to join nodes together.
+So far, most WBANs have used tree-like networks to join nodes together.
 Tree networks, while simple, are negatively affected by node failures and movement.
@@
 %In order to determine the improvements of novel methods, reference systems need to be available to compare with.
-Both~\cite{mac:tdma}~and~\cite{mac:lowdutycycle} provide a suitable reference system for a synchronized, non-relaying BSN.
+Both~\cite{mac:tdma}~and~\cite{mac:lowdutycycle} provide a suitable reference system for a synchronized, non-relaying WBAN.
 Both proposed systems present, in detail, a TDMA MAC protocol specifically designed with energy efficiency in mind.

### 2011-09-23 docs/thesis/thesis.tex
Final thesis draft.

@@
 \begin{document}
-\title{Body Area Network Platform Utilizing Energy-Efficient Routing of Physiological Data}
-\author{Alvaro Prieto\\
-Masters Thesis\\
-BS/MS Electrical Engineering\\
+\title{Wireless Body Area Network Platform Utilizing Energy-Efficient Routing of Physiological Data}
+\author{by\\
+Alvaro Prieto\\
+Master of Science Thesis\\
+in\\
+Electrical Engineering\\
+Approved by:\\
+\makebox[3.5in]{\hrulefill}\\
+(Gill Tsouri, Ph. D.)\\
+\makebox[3.5in]{\hrulefill}\\
+(Andres Kwasinski, Ph. D.)\\
+\makebox[3.5in]{\hrulefill}\\
+(Nirmala Shenoy, Ph. D.)\\
+\makebox[3.5in]{\hrulefill}\\
+(Sohail A. Dianat, Ph. D.)\\
 Electrical and Microelectronic Engineering Department\\
+Kate Gleason College of Engineering\\
 Rochester Institute of Technology\\
-Rochester, NY\\
+Rochester, New York\\
 \texttt{axp8664@rit.edu}}
@@
 Due to the simple nature of the planned WBAN, a much less complex approach could be used.
-The main application only required streaming data from various wireless sensors.$\ll$ Should I elaborate on the design constraints here?
-The number of devices is known, the time of transmission is known etc...$\gg$ This could be accomplished with a very minimal radio library.
+The main application only required streaming data from various wireless sensors.
+This could be accomplished with a very minimal radio library.
 
@@
 The data was sampled, and the routing algorithm was run, at a rate of 5Hz.
-This sampling rate is appropriate as long as the subject, or its environment, does not move faster than this.$\ll$What was the name for this?$\gg$ If data needs to be sampled at a faster rate, some hardware and firmware changes might be necessary, but the host software will still work.
+This sampling rate is appropriate as long as the subject, or its environment, does not move faster than this.
+If data needs to be sampled at a faster rate, some hardware and firmware changes might be necessary, but the host software will still work.
 
@@
 \begin{table}[htb]
+\begin{center}
+
 \begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}
@@
 \label{tab:rssi}
+\end{center}
 \end{table}
@@
 \begin{figure}[!htb]
-\includegraphics[width=0.8\textwidth]{figures/sit1-c0.pdf}
+\begin{center}
+\includegraphics[width=0.75\textwidth]{figures/sit1-c0.pdf}
 \caption{Sitting M=0}
 \label{fig:sit1-c0}
+\end{center}
 \end{figure}
@@
 \begin{figure}[!htb]
-\includegraphics[width=0.8\textwidth]{figures/sit1-c100.pdf}
+\begin{center}
+\includegraphics[width=0.75\textwidth]{figures/sit1-c100.pdf}
 \caption{Sitting M=100}
 \label{fig:sit1-c100}
+\end{center}
 \end{figure}
@@
 Figures~\ref{fig:walk2-c0}~and~\ref{fig:walk2-c100}, on the other hand, show an increase in network lifetime.
-For example, at the end of the 5 minute sampling run with $M=0$, the device which consumed the most power was ED7 with approximately 1.7 mW.
-On the other hand, with $M=100$, all EDs consume power in a uniform fashion and end up with less than 1.3 mW each.
-One important observation is that the network with $M=0$ consumed less overall power (~9.6mW) than the one with $M=100$ (~10.1mW).
+For example, at the end of the 5 minute sampling run with $M=0$, the device which consumed the most power was ED7 with approximately 1.06 $\mu$J.
+On the other hand, with $M=100$, all EDs consume power in a uniform fashion and end up with less than 0.81 $\mu$J each.
+One important observation is that the network with $M=0$ consumed less overall power (~6 $\mu$J) than the one with $M=100$ (~6.31$\mu$J).
 This is expected, due to the fact that the new cost function specifically avoids the most efficient path for a specific node in favor of the most efficient path for the whole network.
@@
 
-To demonstrate the network lifetime improvement, an ED will be considered have depleted it's power source after consuming 1mW of normalized energy.
-In the previous example, the trial with $M=0$ reaches 1mW after approximately 175 seconds, while the case with $M=100$ reaches it around 240 seconds.
+To demonstrate the network lifetime improvement, an ED will be considered have depleted it's power source after consuming 0.625 $\mu$J of energy.
+In the previous example, the trial with $M=0$ reaches 0.625 $\mu$J after approximately 175 seconds, while the case with $M=100$ reaches it around 240 seconds.
 
 \begin{figure}[!htb]
-\includegraphics[width=0.8\textwidth]{figures/walk2-c0.pdf}
+\begin{center}
+\includegraphics[width=0.75\textwidth]{figures/walk2-c0.pdf}
 \caption{Walking M=0}
 \label{fig:walk2-c0}
+\end{center}
 \end{figure}
@@
 \begin{figure}[!htb]
-\includegraphics[width=0.8\textwidth]{figures/walk2-c100.pdf}
+\begin{center}
+\includegraphics[width=0.75\textwidth]{figures/walk2-c100.pdf}
 \caption{Walking M=100}
 \label{fig:walk2-c100}
+\end{center}
 \end{figure}
@@
 \begin{figure}[!ht]
-\includegraphics[width=0.8\textwidth]{figures/walk4-c100.pdf}
+\begin{center}
+\includegraphics[width=0.75\textwidth]{figures/walk4-c100.pdf}
 \caption{Walking M=100}
 \label{fig:walk4-c100}
+\end{center}
 \end{figure}

### 2011-09-23 host/threadtest/main.c

@@ -50,7 +50,7 @@ int main( int argc, char *argv[] )
   // Make sure input is correct
   if( argc < 6 )
   {
-    printf("Usage: %s port baudrate C(0.0-1.0) [graph (0,1)] timeout\n", 
+    printf("Usage: %s port baudrate C(0.0-1000.0) [graph (0,1)] timeout\n", 
                                                                       argv[0]);
     return 0;
   }
@@ -68,8 +68,9 @@ int main( int argc, char *argv[] )
   printf("Data collection will start in ~%d seconds...\n", timeout);
 
   sleep(timeout);
+#ifdef ENABLE_MUSIC  
   system("banshee --play");
-
+#endif
   // Initialize routing an power tables
   memset( (uint8_t*)power_table, 0xff, MAX_DEVICES );
   memset( (uint8_t*)routing_table, ( MAX_DEVICES + 1 ), MAX_DEVICES );
@@ -281,9 +282,11 @@ void sigint_handler( int32_t sig )
 
     // Close the serial port
     serial_close();
-    
+
+#ifdef ENABLE_MUSIC    
     // Stop playing music
     system("banshee --pause");
+#endif
     
     printf("\nExiting...\n");
     exit(sig);
